### 2018-02-20 generativeModel.tex
initial commit

\documentclass{article}

\usepackage[margin=0.1in]{geometry}
\usepackage{booktabs}
\usepackage{tipa}
\usepackage{amssymb}
\usepackage{tikz}
\usetikzlibrary{fit}
\usepackage{color}
\usepackage{multirow}
\usepackage{hyperref}

\newcommand{\nextForm}[1]{\rotatebox[origin=c]{270}{$_{\curvearrowright}$}$_{#1}$}

\begin{document}

\begin{figure}
  \begin{tikzpicture}

    \node(UG)[purple,align = center] at (-1,-1) {\textbf{Programming language}\\\textbf{(Universal Grammar)}};
    \newcommand*{\universalX}{13}%

    \node(sounds) at (\universalX - 3.5,-0.3) {Phoneme Inventory};
    \node(soundExamples)[draw,align = center] at (\universalX - 3.5,-1.3) {\textipa{@, p, E, 2, k}\\\textipa{a, 3, A, O, d}\\$\cdots$};

    \node at (\universalX + 0.5,-0.3) {Program space};
    \node(programSpace)[red,draw,align = left] at (\universalX + 0.5,-1.3) {$\mathcal{M}\to \mathcal{A} | $\verb|stem|$|\mathcal{M}+\mathcal{M}$\\$\mathcal{P}\to \mathcal{R}\circ \mathcal{R}\circ \cdots \circ \mathcal{R}$};

    \node(featureSpace) at (\universalX - 7,-0.3) {Feature space};
    \node(featureExamples)[draw,align = center] at (\universalX - 7,-1.3) {\verb|voiced|, \verb|vowel|\\\verb|obstruent|, \verb|nasal|\\$\cdots$};

    \draw[thick,->] (featureExamples.east) -- (soundExamples.west);
    \draw[thick,->] (soundExamples.east) -- (programSpace.west);

    \node(universalBox)[purple,draw,ultra thick, fit = (sounds) (featureSpace) (featureExamples) (soundExamples) (programSpace)] {};
    \draw(lowerCausalArrow)[thick,->] (featureExamples.south) .. controls (\universalX - 3,-2.5) ..  (programSpace.south);

    \node(morphology)[blue,align = center] at (-1,-5) {\textbf{Language-specific}\\\textbf{morphophonology}};
    \draw[ultra thick,->] (UG.south) -- (morphology.north);

    \newcommand*{\syntheticX}{15.5}%
    \newcommand*{\ChineseX}{10.5}%

    \node(AABmorphologyLabel) at (\syntheticX-0.5,-3.5) {AAB Morphology};
    \node(AABmorphology) [draw,align = left] at (\syntheticX-0.5,-4){ \verb|A|$+$\verb|B|};

    \node(AABphonologyLabel) at (\syntheticX - 0.5,-5.5) {AAB Phonology};
    \node(AABphonology) [draw,align = left] at (\syntheticX - 0.5,-6) {%
      $r_1$:      $\varnothing\to \sigma_i$\verb| / | $\sigma_i$\verb|_|$\sigma$};
    \node(AABmorphologyBox)[blue,draw,ultra thick, fit = (AABmorphologyLabel) (AABphonology) (AABmorphology)] {};

    \node(ChinesemorphologyLabel) at (\ChineseX - 0.5,-3.5) {Chinese Morphology};
    \node(Chinesemorphology) [draw,align = left] at (\ChineseX - 0.5,-4){ \verb|stem|+/\textipa{d@}/};

    \node(ChinesephonologyLabel) at (\ChineseX - 0.5,-5.5) {Chinese Phonology};
    \node(Chinesephonology) [draw,align = left] at (\ChineseX - 0.5,-6) {%
      $r_1$:      $\varnothing\to \sigma_i$\verb| / |$\sigma_i$ \verb|_| \textipa{d@}};
    \node(ChinesemorphologyBox)[blue,draw,ultra thick, fit = (ChinesemorphologyLabel) (Chinesephonology) (Chinesemorphology)] {};

    \node(polishMorphologyLabel) at (4.5 - 0.5,-3.5) {Polish Morphology};
    \node(polishMorphology) [draw,align = left] at (4.5 - 0.5,-4.3) {\verb|singular|$\to$ \verb|stem|\\\verb|plural|$\to$ \verb|stem|$ + $ /\textipa{i}/};

    \node at (4.5 - 0.5,-5.5+0.4) {Polish Phonology};
    \node(polishPhonology) [draw,align = left] at (4.5 - 0.5,-6.3+0.4) {%
      $r_1$: \textipa{o}$\to$\textipa{u}$/$\verb|_[-nasal +voice]#|\\%
      $r_2$: \verb|[-sonorant]|$\to$\verb|[-voice]|$/$\verb|_#|};
    \node(polishMorphologyBox)[blue,draw,ultra thick, fit = (polishMorphologyLabel) (polishPhonology)] {};

    \draw[ultra thick,->] (universalBox.south) -- (AABmorphologyBox.north);
    \draw[ultra thick,->] (universalBox.south) -- (ChinesemorphologyBox.north);
    \draw[ultra thick,->] (universalBox.south) -- (polishMorphologyBox.north);

    \node(lexicon)[magenta,align = above] at (-1,-10) {\textbf{Observed data}};
    \draw[ultra thick,->] (morphology.south)-- (lexicon.north);

    \node(p1) [align = left] at (\syntheticX + 1,-7.5) {\verb|A|$+$\verb|B:| \textipa{pofe}};
    \node(w1) [align = left] at (\syntheticX + 1,-8) {\verb|A|$+$\verb|B:| \textipa{wug@}};
    \node(p2) [draw,align = left] at (\syntheticX - 1 - 1,-9.5) {\textipa{pofe}\nextForm{r_1}\\\textipa{popofe}};
    \draw [thick,->] (p1.west) -- (p2.north);

    \node(w2) [draw,align = left] at (\syntheticX + 1 - 1,-9.5) {\textipa{wug@}\nextForm{r_1}\\\textipa{wuwug@}};
    \draw [thick,->] (w1.south) -- (w2.north);
    \node(aablexicon) [label = {[name = l] \small \emph{lexicon:}},black,draw,fit = (w1) (p1)] {};    
    \node[magenta,draw,align= left](AABvocabulary) at (\syntheticX - 1,-11.5){\textipa{popofe}, \textipa{wuwug@}, $\cdots$ };
    \draw [thick,->] (p2.south) -- (AABvocabulary.north);
    \draw [thick,->] (w2.south) -- (AABvocabulary.north);
    \node(AABlexiconBox)[black,draw,ultra thick, fit = (p2) (w2) (AABvocabulary)] {}; %(l) (aablexicon) 
    \draw [ultra thick,->] (AABmorphologyBox.south) -- (AABlexiconBox.north);
%    \draw [->] (soundExamples.south) -- (aablexicon.west);

    \node(p1) [align = left] at (\ChineseX + 1.5,-7.5) {\textipa{xau}};
    \node(w1) [align = left] at (\ChineseX + 1.5,-8) {\textipa{\c{c}in}};
    \node(p2) [draw,align = left] at (\ChineseX - 1 - 1,-9.5) {\textipa{xaud@}\nextForm{r_1}\\\textipa{xauxaud@}};
    \draw [thick,->] (p1.west) -- (p2.north);

    \node(w2) [draw,align = left] at (\ChineseX + 1 - 1,-9.5) {\textipa{\c{c}ind@}\nextForm{r_1}\\\textipa{\c{c}in\c{c}ind@}};
    \draw [thick,->] (w1.south) -- (w2.north);

    \node(chineseLexicon) [label = {[name = l] \small \emph{lexicon:}},black,draw,fit = (w1) (p1)] {};

    \node[magenta,draw,align= left](Chinesevocabulary) at (\ChineseX - 1,-11.5){\textipa{xauxaud@}, \textipa{\c{c}in\c{c}ind@}, $\cdots$ };
    \draw [thick,->] (p2.south) -- (Chinesevocabulary.north);
    \draw [thick,->] (w2.south) -- (Chinesevocabulary.north);
    \node(ChineselexiconBox)[black,draw,ultra thick, fit = (Chinesevocabulary) (p2) (w2)] {};
    \draw [ultra thick,->] (ChinesemorphologyBox.south) -- (ChineselexiconBox.north);

    \node(crib) at (6.5,-7.6) {\verb|horn:| \textipa{rog}};
    \node(polishLexicon)[label = {[name = l] \small \emph{lexicon:}},black,draw,fit = (crib)] {};
    \node(Cingular)[draw,align = left] at (3.5-0.5,-9.5) {\verb|singular:|\\ \textipa{rog}\nextForm{r_1}\\\textipa{rug}\nextForm{r_2}\\\textipa{ruk}};
    \node(plural)[draw,align = left] at (5.5-0.5,-9.5) {\verb|plural:|\\ \textipa{rogi}\nextForm{r_1}\\\textipa{rogi}\nextForm{r_2}\\\textipa{rogi}};
    \draw [thick,->] (crib.south) -- (Cingular.north);
    \draw [thick,->] (crib.south) -- (plural.north);

    \node[magenta,draw,align = left] (polishObservation) at (4.5-0.5,-11.5) {\verb|horn, singular:| \textipa{ruk}\\\verb|horn, plural:| \textipa{rogi}};
    \draw [thick,->] (Cingular.south) -- (polishObservation.north);
    \draw [thick,->] (plural.south) -- (polishObservation.north);

    \node(polishData)[black,draw,ultra thick, fit = (Cingular) (plural) (polishObservation)] {};
    \draw [ultra thick,->] (polishMorphologyBox.south) -- (polishData.north);

  \end{tikzpicture}
  %% \caption{{\color{red}Red}: Both the morphology $\mathcal{M}$ and
  %%   phonology $\mathcal{P}$ are programs either build through
  %%   concatenation or function composition
  %%   (Tables~\ref{morphologyGrammar} \&~\ref{phonologyGrammar}). These programs explain how words are built in both natural languages ({\color{blue}blue}) and synthetic languages used in infant studies of rule learning ({\color{green}green}). Learners observe pronunciations of words ({\color{orange}orange}) in order to recover these programs, but with the aid of a strong inductive bias (Universal Grammar, {\color{purple}purple})}.
  %% \label{beautiful}
\end{figure}

\end{document}

### 2018-02-20 outline.tex
initial commit

\documentclass{article}

\usepackage{amsmath}
\usepackage[margin=0.1in]{geometry}
\usepackage{booktabs}
\usepackage{tipa}
\usepackage{amssymb}
\usepackage{tikz}
\usetikzlibrary{fit}
%\usepackage{color}
\usepackage[dvipsnames]{xcolor}
\usepackage{multirow}
\usepackage{hyperref}

 % For algorithms
\usepackage{algorithm}
\usepackage{algorithmic}

\title{Discovering the Rules of language using Bayesian Program Learning}
%\author{Kevin Ellis, Timothy O'Donnell}

\newcommand{\nextForm}[1]{\rotatebox[origin=c]{270}{$_{\curvearrowright}$}$_{#1}$}
\DeclareMathOperator*{\argmin}{arg\,min}

\begin{document}
\maketitle

%% How do linguists come up with phonological rules, how do kids learn
%% artificial grammars, and how does one acquire pig latin?  The
%% solutions to these problems share a common representation, which we
%% show can be modeled as a program, and the corresponding learning
%% problems modeled as program induction.  This framing lets us apply
%% ideas from Bayesian Program Learning to induce grammars, which
%% combines program synthesis techniques with a compression-based inductive
%% bias. This lets the models capture phonological phenomena like vowel
%% harmony or stress patterns and learn synthetic grammars used in prior
%% studies of artificial grammar learning.  Going beyond individual
%% grammar learning problems, we consider the problem of jointly
%% inferring many related rule systems. By solving many textbook
%% phonology problems, we can ask the model what kind of inductive bias
%% best explains the attested phenomena.

\begin{abstract}
    %% How do linguists come up with phonological rules? How do kids
    %% learn pig latin?  We develop computational models of these and
    %% other phenomena in language.  The unifying element of the models
    %% is to treat grammars as programs, which lets us apply ideas from
    %% the field of program synthesis to learn grammars. This lets the
    %% models capture phonological phenomena like vowel harmony or stress
    %% patterns and learn synthetic grammars used in prior studies of
    %% rule learning.  Going beyond individual grammar learning problems,
    %% we consider the problem of jointly inferring many related rule
    %% systems. By solving many textbook phonology problems, we can
    %% ask the model what kind of inductive bias best explains the
  %% attested phenomena.
Both children and linguists confront a similar problem of inference:
given utterances produced by speakers, together with aspects of the
meaning of those utterances, discover the grammatical principles that
relate form to meaning. We study this abstract computational
problem within the domain of morphophonology, contributing a computational model
that learns phenomena from many natural languages and generalizes in
humanlike ways from data used in behavioral studies of artificial grammar
learning.

Our work draws on two analogies. The \emph{child-as-linguist analogy}
holds that both children and linguists must solve the same abstract
inductive reasoning problem, even though the nature of the input data and underlying mental algorithms
are surely different in precise detail. Accordingly we isolate the
problem of learning morphophonological systems, and show that a single
solution to this problem can capture both linguistic analyses from
natural languages and infant rule learning of artificial languages.
We adopt the framework of ``Bayesian Program learning'' (BPL) -- in
which learning is formulated a synthesizing a program which compactly
describes the input data. This \emph{learning-as-programming analogy}
lets us exploit recent techniques from the field of program synthesis to
induce morphophonological rules from data.
While \emph{child-as-linguist} poses the
computational problem, \emph{learning-as-programming} offers a
solution.
  \end{abstract}

\section{Introduction}

Both children and linguists confront a similar problem of inference:
given utterances produced by speakers, together with aspects of the
meaning of those utterances, discover the grammatical principles that
relate form to meaning. Here, we study this abstract computational
problem within the domain of morphophonology, the part of language
that builds up words from simpler parts such as stems, prefixes, and
suffixes (morphology) and then specifies how the words are realized as
sound patterns (phonology).  Our contribution is a computational model
that learns phenomena from 26 natural languages and generalizes in
humanlike ways from data used in infant studies of artificial grammar
learning.

Our work draws on two analogies. The \emph{child-as-linguist analogy}
holds that both children and linguists must solve the same abstract
inductive reasoning problem, even though the nature of the input data,
the process of theory construction, and underlying mental algorithms
are surely different in precise detail. Accordingly we isolate the
problem of learning morphophonological systems, and show that a single
solution to this problem can capture both linguistic analyses from
natural languages and infant rule learning of artificial languages.

We adopt the framework of ``Bayesian Program learning'' (BPL) -- in
which learning is formulated a synthesizing a program which compactly
describes the input data. This \emph{learning-as-programming analogy}
explains the flexibility and richness of a grammar
as a consequence of 
While \emph{child-as-linguist} poses the
computational problem, \emph{learning-as-programming} offers a
solution.

We develop a programming language tailored to expressing
morphophonological generalizations.
For children this programming language is innate and is
commonly called ``Universal Grammar'' (UG). Linguists instead must
uncover this programming language by looking at many languages and
searching for commonalities. Learning what makes up UG unfolded over
evolutionary time for children and is ongoing within linguistics.  BPL
reinterprets the problem of discovering UG.  In BPL, UG is the
programming language combined with an inductive bias over programs.
Here we estimate this inductive bias empirically from naturally
occurring morphophonological systems as a way of learning to learn
grammars.

\section{Morphophonological phenomena}

What is phonology supposed to explain? Familiar phenomena from
English help motivate phonological grammars.  The plural of \emph{cat}
is pronounced \textipa{k\ae ts} (``cats''), while the plural of
\emph{dog} is pronounced \textipa{dagz} (``dogs''). A phonological
rule accounts for the alternation between \textipa{s} and \textipa{z}:
\textipa{z} becomes \textipa{s} at the end of words whenever the
preceding sound is unvoiced (does not involve vibrating the vocal
cords). This generalization holds throughout the entirety of English.
The count system of Tibetan presents another puzzle.
``1'' in Tibetan is pronounced
\textipa{\|x{j}ig}, ``10'' is \textipa{\|x{j}u},
yet ``11'' ($ = 10 + 1$) is \textipa{\|x{j}ug\|x{j}ig},
rather than \textipa{\|x{j}u\|x{j}ig}.
The explanation lies in Tibetan phonology,
which deletes word initial consonants followed by another consonant (Table~\ref{Tibetan}).
In this analysis of Tibetan,
``1'' has a latent ``underlying'' representation as \textipa{g\|x{j}ig}
but is pronounced as \textipa{\|x{j}ig}.

The aim of phonology is to explain pronunciations of words in terms of
latent underlying representations (eg, \textipa{g\|x{j}ig} for Tibetan
``1'') and the rules that give observed surface pronunciations from
underlying forms (eg, deleting word initial consonants).
Context-sensitive rewrite systems are a Turing complete representation
that also naturally express phonological
rules~\cite{chomsky1968sound}.  We use context-sensitive rewrites as a
representation for programs that implement a language's phonology.
For example, the rewrite in Table~\ref{Tibetan},
\verb|C|$\to\varnothing /$\verb|#_C|, deletes ($\to\varnothing $)
consonants (\verb|C|) at word boundaries (\verb|#|) when followed by
another consonant (\verb|C|).
Thus phonological rule learners, whether they be children, linguists, or machines, must solve a program induction problem.

\subsection{Inducing Programs}
We model the program learning problem as Bayesian inference, treating the pronunciations of words as observed data and the phonology as a latent variable whose values are programs. So the model's beliefs about possible programs are:
\begin{equation}
  P(\text{program}|\text{words})\propto P(\text{program}|\text{UG})\prod_{w\in\text{words}}\sum_{u\in\text{underlying forms}} P(u)P(w|u,\text{program})
  \end{equation}
Here, Universal Grammar (UG) is an inductive bias over programs. This inductive bias encodes knowledge of the program space (what rules are possible in human language and what rules are impossible), imparting a bias towards more naturally structured rules, like rules written in terms of syllables and consonants.

\subsection{The role and importance of universal grammar}
                We claim any plausible theory of linguistic rule learning needs a universal grammar. Studies of infant grammar learning suggests that babies come equipped with prior knowledge of what language should look like. This prior knowledge constrains the space of allowed generalizations, accounting for the astonishing pace at which children acquire language. This line of reasoning is a ``poverty-of-the-stimulus'' style argument: if children can induce so much grammar from such meager data, something else must bridge the gap. Because universal grammar also constrains the space of allowed grammars, we would expect that humans do not learn certain generalizations, even when the generalizations seem to hold in the data. Indeed this is the case: this ``surfeit-of-the-stimulus'' style argument  explains phenomena in natural languages (for example, statistical regularities that seem to hold in Turkish are not actually used by speakers) and in infant artificial language learning (for example, infants acquire certain phonological generalizations from just a few examples, while never acquiring other generalizations that are seemingly just a simple). While previous accounts of universal grammar were abstract, our model is concrete and computational. We define universal grammar as a probability mass function that puts zero mass on impossible or unattested generalizations, places more mass on programs licensing the generalizations that people learn quickly and naturally, and then smoothly falls off in probability mass as the programs become less natural.

\subsection{Learning natural language phonology}                
Our model learns a diverse set of phonological generalizations found in natural languages. We compiled a multilingual data set of words from standard linguistic textbooks. In these data sets, our model recovers allophonic relationships (e.g., l vs. r in Japanese) and systems that form different inflections of words (e.g., English verb inflections, or the Tibetan count system). Systems like these have received intense study within computational linguistics. But no prior model could explain the diversity of phonological phenomena found even in a textbook. To explain many diverse phenomena we need a rich, flexible representation -- programs -- coupled to a linguisticly informed inductive bias -- universal grammar. Our model learns simple generalizations like English verbal inflections, as well as phenomena like vowel harmony (in the language Kikuria, the word tiga becomes tegera in a different inflection, or ruga becomes rogera), or tonal rules in languages with tone (in the language Kerewe, different inflections cause tones to appear, disappear, or move to different parts of a word).

For these results, the learning-as-programming analogy does the work for us. BPL moves beyond the impoverished representations of prior learners (neural networks as in, or rigid classes of symbolic rules as in) instead represented generalizations as programs. We claim that programs are a sufficiently general representation to support the kinds of rules found in phonology.

\begin{table}\centering
\begin{tabular*}{10cm}{lcl}
  \toprule
  Example data&Phonology\\\midrule
\begin{tabular}{ll}
1&\textipa{\|x{j}ig}\\
4&\textipa{\|x{s}i}\\
5&\textipa{Na}\\
9&\textipa{gu}\\
10&\textipa{\|x{j}u}\\
11 ($= 10 + 1$) & \textipa{\|x{j}ug\|x{j}ig}\\
14 ($= 10 + 4$) & \textipa{\|x{j}ub\|x{s}i}\\
15 ($= 10 + 5$) & \textipa{\|x{j}uNa}\\
19 ($= 10 + 9$) & \textipa{\|x{j}urgu}\\
40 ($= 4 + 10$) & \textipa{\|x{s}ib\|x{j}u}\\
50 ($= 5 + 10$) & \textipa{Nab\|x{j}u}\\
90 ($= 9 + 10$) & \textipa{gub\|x{j}u}
\end{tabular}
&
\begin{tabular}{l}
  \emph{Consonant cluster reduction:}\\
  \verb|C|$\to\varnothing $ $/$ \verb|# _ C|\\
  \end{tabular}

  \\
  \bottomrule  \end{tabular*}
\caption{Tibetan count system.}
\label{Tibetan}
  \end{table}

\pagebreak

%% 2 insights: analogy between children and linguists. analogy between
%% grammars and programs.

%% The core problem we're pushing on is \textbf{structure learning in
%%   language}. It makes sense that you could learn parameters and
%% lexical items, but could you learn something with more the structure
%% of a grammar? Here we answer that question in the affirmative within a
%% restricted domain (phonology) on restricted problems (smaller data
%% sets and simpler sets of rules).

%% Why focus on some problems that feel more like computational linguistics and other problems that feel more like developmental psychology? Connection to child-as-linguist~\cite{chomsky1988current}, child-as-scientist, child-as-hacker\cite{Rule2015RepresentingAL}.

%% What is the connection between what children do and what linguists do? From the point of view of this research project,
%% the difference lies in the nature and role of universal grammar.
%% For a linguist, universal grammar is an unknown to be discovered.
%% When analyzing a new data set, a linguist might need to entertain a broader or more expressive universal grammar.
%% For the child, universal grammar is an innate part of our genetic endowment.

\section{The Model}

\begin{figure}[h!]\centering
  \includegraphics[width = 0.9\textwidth]{generativeModel.pdf}
  \caption{{\color{red}Red}: Both the morphology $\mathcal{M}$ and
    phonology $\mathcal{P}$ are programs either build through
    concatenation or function composition
    (Tables~\ref{morphologyGrammar} \&~\ref{phonologyGrammar}). These programs explain how words are built in both natural and artificial grammars ({\color{blue}blue}). Learners observe pronunciations of words ({\color{magenta}}) in order to recover these programs, but with the aid of a strong inductive bias (Universal Grammar, {\color{purple}purple})}.
  \label{beautiful}
\end{figure}

\section{Artificial Grammar Learning}

The fundamental hypothesis underlying AGL research is that
artificial grammar learning must engage some shared resource with first language acquisition --
and so we can gain insights into first language acquisition by
instead studying AGL in controlled experiments.

Our contribution to AGL research is the demonstration that
a single BPL model can explain (1) that infants quickly acquire
certain simple grammars very easily, and at the same time (2)
those same infants go on to learn
the very complicated grammar of their first language.
We collected grammars from AGL datasets~\cite{gerken2010infants,marcus1999rule,frank2011three}
and presented our BPL learner with data drawn from these grammars (Table~\ref{artificialGrammarTable}),
finding that our model recovers the generative processes that
 underlies these data sets.
\begin{table}[h!]
\centering\begin{tabular}{lcll}
  \toprule
  Grammar&Example input to learner&MAP grammar&Natural language analogues\\\midrule
  \begin{tabular}{l}
    ABA (same/different/same)\\
  Marcus et al. 1999.     
    \end{tabular}&
  \begin{tabular}{l}
    wofewo\\
    lovilo\\
    fimufi
  \end{tabular}&
  \begin{tabular}{l}
$\varnothing\to \sigma_i$\verb| / #_|$\sigma \sigma_i$
  \end{tabular}& Reduplication (eg, Tagalog)
  \\\midrule
  \begin{tabular}{l}
      ABB (different/same/same)\\
  Marcus et al. 1999. 
    \end{tabular}&
  \begin{tabular}{l}
        wofefe\\
    lovivi\\
    fimumu
  \end{tabular}&
  \begin{tabular}{l}
$\varnothing\to \sigma_i$\verb| / |$\sigma_i$\verb|_#|
  \end{tabular}&
  Reduplication (eg, Tagalog)\\\midrule
  \begin{tabular}{l}
      ABx (different/different/constant)
    \end{tabular}&
  \begin{tabular}{l}
        wofeka\\
    lovika\\
    fimuka
  \end{tabular}&
  \begin{tabular}{l}
stem$ + x$
  \end{tabular}&concatenative morphology
  \\\midrule
  \begin{tabular}{l}
    AAx (same/same/constant)\\Gerken 2006. %~\cite{gerken2010infants}
    \end{tabular}&
  \begin{tabular}{l}
        wowoka\\
    loloka\\
    fifika
  \end{tabular}&
  \begin{tabular}{l}
    stem + $x$\\
$\varnothing\to \sigma_i$\verb| / #_|$\sigma_i$
  \end{tabular}&
  \begin{tabular}{l}
    reduplication\\concatenative morphology
    \end{tabular}\\\midrule
  \begin{tabular}{l}
    AxA (same/constant/same)\\Gerken 2006. %~\cite{gerken2010infants}
    \end{tabular}&
  \begin{tabular}{l}
        wokawo\\
    lokalo\\
    fikafi
  \end{tabular}&
  \begin{tabular}{l}
$x + $ stem\\
    $\varnothing$ $\to\sigma_i$  / \verb|# _| $\sigma$ $\sigma_i$
  \end{tabular}&
  \begin{tabular}{l}
    Infixing\\Reduplication
    \end{tabular}
  \\\midrule
  \begin{tabular}{l}
    Pig Latin
    \end{tabular}&
  \begin{tabular}{l}
    \textipa{pIg}$\to$\textipa{Igpe}\\
    \textipa{latIn}$\to$\textipa{atIle}\\
    \textipa{\ae sk}$\to$\textipa{\ae ske}\\
  \end{tabular}&
  \begin{tabular}{l}
      $\varnothing\to$\verb|C|$_i$ $/$ \verb|#C|$_i$\verb| [ ]* _ #|\\%
      $\varnothing\to$ \textipa{e} $/$ \verb| _ #|\\%
 \verb|C|$\to\varnothing$ $/$ \verb|# _|
  \end{tabular}  & \begin{tabular}{l}
    \textbf{Tim, is there an  analog?}\\
    \textbf{word initial segments moves to  }\\
    \textbf{word final?}
    \end{tabular}
  \bottomrule  \end{tabular}
\caption{Using BPL to model artificial grammar learning. Learner given 5 examples.}\label{artificialGrammarTable}
\end{table}

Just like in first language acquisition,
these AGL learning setups introduce a trade-off between
grammars that are likely and therefore small (``parsimony'')  and grammars that
best fit the data.
For example, a learner could infer a grammar that just memorized to the data (perfect fit but poor parsimony)
or it could infer a grammar that can generate every
possible word (parsimonious but a poor fit).
Effective grammar learners must navigate this trade-off.

To analyze this trade-off, we calculated its shape in the form of a
\textbf{pareto front}~\cite{mattson2005pareto}.
In our setting, a pareto front is the set of all grammars that
are not worse than another grammar along
the two competing axes of parsimony and fit to data.
Intuitively, grammars on the pareto front
are ones which an ideal Bayesian or MDL learner prefers,
\emph{independent} of how the learner decides to
relatively weight the prior and likelihood.
Figure~\ref{quadrupleFrontier}
diagrams the Pareto fronts for two AGL experiments
as the number of
example words provided to the learner is varied.
What these Pareto fronts show is (1)
the set of grammars entertained by the learner,
and (2) how the learner weighs these grammars against each other
as measured by the prior (parsimony) and the likelihood (fit to the data).
We believe that the Pareto front is a useful way of
viewing the space of possible grammars,
and understanding the different trade-offs
between the grammars.

\begin{figure}[H]

\includegraphics[width = 0.5\textwidth]{abb1.png} \includegraphics[width = 0.5\textwidth]{abb3.png}
\includegraphics[width = 0.5\textwidth]{aax1.png} \includegraphics[width = 0.5\textwidth]{aax3.png}

  \caption{Pareto fronts for ABB  (Marcus~1999; top two) \& AAX  (Gerken~2006; bottom two) learning problems fo either one example word (left column) or three example words  (right column). {\color{red}Red shade}: ground truth grammar. {\color{red!60}Pink shade}: shares structure with ground truth grammar. White shade: incorrect generalizations. As the number of examples increases, the Pareto fronts develop a sharp kink around the ground truth grammar, which indicates a stronger preference for the correct grammar. With one example the kinks can still exists but are less pronounced.}\label{quadrupleFrontier}
\end{figure}

\subsection{Beyond artificial grammars: Applying BPL to natural language}

A successful model of
artificial grammar learning should also explain at least some aspects of first language acquisition.
As a first step in this direction,
we take the same model that we applied to
these AGL learning problems and use it to solve phonology textbook exercises.
Each of these textbook exercises is an inductive reasoning
problem where the learner must
explain surface pronunciations
in terms of a latent generative model (a program).
Figure~\ref{naturalLanguageExample}
shows our model solving a
representative textbook problem.

\textbf{TIM: Should I talk about UG? We needed it for Tibetan. We probably don't have space.}

\begin{figure}\centering
  \begin{minipage}{7cm}\centering
    \textbf{Textbook Tibetan counting problem:}
    \begin{tabular}{ll}\toprule
Number & Surface form
\\ \midrule
10 & \textipa{d\super Zu}\\
1 & \textipa{d\super Zig}\\
11 & \textipa{d\super Zugd\super Zig}\\
4 & \textipa{Si}\\
%14 & \textipa{d\super ZubSi}\\
40 & \textipa{Sibd\super Zu}\\
9 & \textipa{gu}\\
19 & \textipa{d\super Zurgu}\\
%90 & \textipa{gubd\super Zu}\\
5 & \textipa{Na}\\
%15 & \textipa{d\super ZuNa}\\
%50 & \textipa{Nabd\super Zu}\\
\bottomrule
\end{tabular}
  \end{minipage}%
  \hspace{0.5cm}%
  \begin{minipage}{6cm}\centering
    \textbf{Our model's solution:}

    Induced phonological rule: \texttt{C}$\to\varnothing /$\texttt{\#\_C} \\
    \hspace{0.5cm}\emph{(Reduce word initial consonant clusters)}

    Induced underlying representations:
\\\begin{tabular}{ll}\toprule
1 & \textipa{gd\super Zig}\\
4 & \textipa{bSi}\\
5 & \textipa{Na}\\
9 & \textipa{rgu}\\
10 & \textipa{bd\super Zu}
\bottomrule
\end{tabular}    
  \end{minipage}
  \caption{Example textbook phonology exercise (left) and the solution our model discovers for it (right). To solve this problem the model must jointly infer unobserved underlying representations for each of the Tibetan numbers, along with a phonological rule that explains their surface pronunciations. The same code that calculated the Pareto fronts (Figure~\ref{quadrupleFrontier}) produces this analysis of the Tibetan count system.}\label{naturalLanguageExample}
  \end{figure}

%\pagebreak
%% \begin{figure}
%%   \begin{tikzpicture}
%%     %  \draw[help lines] (0,0) grid (10,-10);
%%     \node(UG)[purple,align = center,rotate = 90] at (1,-1) {\textbf{Universal}\\\textbf{Grammar}};
%%     %\draw[ultra thick] (2,0) rectangle (13,-2.3);

%%     \node(sounds) at (7,-0.3) {Phoneme Inventory};
%%     \node(soundExamples)[draw,align = center] at (7,-1.3) {\textipa{@, p, E, 2, k}\\\textipa{a, 3, A, O, d}\\$\cdots$};

%%     \node at (11,-0.3) {Program space};
%%     \node(programSpace)[draw,align = left] at (11,-1.3) {$\mathcal{M}\to \mathcal{A} | $\verb|stem|$|\mathcal{M}+\mathcal{M}$\\$\mathcal{P}\to \mathcal{R}\circ \mathcal{R}\circ \cdots \circ \mathcal{R}$};

%%     \node(featureSpace) at (3.5,-0.3) {Feature space};
%%     \node(featureExamples)[draw,align = center] at (3.5,-1.3) {\verb|voiced|, \verb|vowel|\\\verb|obstruent|, \verb|nasal|\\$\cdots$};

%%     \draw[thick,->] (featureExamples.east) -- (soundExamples.west);
%%     \draw[thick,->] (soundExamples.east) -- (programSpace.west);

%%     \node(universalBox)[purple,draw,ultra thick, fit = (sounds) (featureSpace) (featureExamples) (soundExamples) (programSpace)] {};
%%     \draw(lowerCausalArrow)[thick,->] (featureExamples.south) .. controls (7,-2.5) ..  (programSpace.south);

%%     \node(morphology)[blue,align = above,rotate = 90] at (1,-5) {\textbf{Programs}};
%%     \draw[ultra thick,->] (UG.west) -- (morphology.east);

%%     \node(ABAmorphologyLabel) at (10.5,-3.5) {Pig Latin Morphology};
%%     \node(ABAmorphology) [draw,align = left] at (10.5,-4){ \verb|observation|$\to$\verb|stem|};

%%     \node(ABAphonologyLabel) at (10.5,-5.3) {Pig Latin Phonology};
%%     \node(ABAphonology) [draw,align = left] at (10.5,-6.3) {%
%% $r_1$:      $\varnothing\to$\verb|C|$_i$ $/$ \verb|#C|$_i$\verb| [ ]* _ #|\\%
%%       $r_2$:      $\varnothing\to$ \textipa{e} $/$ \verb| _ #|\\%
%%     $r_3$: \verb|C|$\to\varnothing$ $/$ \verb|# _|};
%%     \node(ABAmorphologyBox)[blue,draw,ultra thick, fit = (ABAmorphologyLabel) (ABAphonology)] {};

%%     %  \draw[ultra thick] (2,-3) rectangle (7,-7);

%%     \node(polishMorphologyLabel) at (4.5,-3.5) {Polish Morphology};
%%     \node(polishMorphology) [draw,align = left] at (4.5,-4.3) {\verb|singular|$\to$ \verb|stem|\\\verb|plural|$\to$ \verb|stem|$ + $ /\textipa{i}/};

%%     \node at (4.5,-5.5) {Polish Phonology};
%%     \node(polishPhonology) [draw,align = left] at (4.5,-6.3) {%
%%       $r_1$: \textipa{o}$\to$\textipa{u}$/$\verb|_[-nasal +voice]#|\\%
%%       $r_2$: \verb|[-sonorant]|$\to$\verb|[-voice]|$/$\verb|_#|};
%%     \node(polishMorphologyBox)[blue,draw,ultra thick, fit = (polishMorphologyLabel) (polishPhonology)] {};

%%     \draw[ultra thick,->] (universalBox.south) -- (ABAmorphologyBox.north);
%%     \draw[ultra thick,->] (universalBox.south) -- (polishMorphologyBox.north);

%%     \node(lexicon)[magenta,align = center,rotate = 90] at (1,-10) {\textbf{Words}};
%%     \draw[ultra thick,->] (morphology.west)-- (lexicon.east);

%%     \node(p1) [align = left] at (9.5,-8) {\verb|LATIN:| \textipa{l\ae tIn}};
%%     \node(p2) [draw,align = left] at (9.5,-9.5) {\textipa{l\ae tIn}\nextForm{r_1}\\\textipa{l\ae tInl}\nextForm{r_2}\\\textipa{l\ae tInle}\nextForm{r_3}\\\textipa{\ae tInle}}; %\\\textipa{l\ae tInle}}};
%%     \draw [thick,->] (p1.south) -- (p2.north);

%%     \node(w1) [align = left] at (12,-8) {\verb|ASK:| \textipa{\ae sk}};
%%     \node(w2) [draw,align = left] at (12,-9.5) {\textipa{\ae sk}\nextForm{r_1}\\\textipa{\ae sk}\nextForm{r_2}\\\textipa{\ae ske}\nextForm{r_3}\\\textipa{\ae ske}};
%%     \draw [thick,->] (w1.south) -- (w2.north);

%%     \node[magenta,draw,align= left](ABAvocabulary) at (10.5,-11.5){\textipa{\ae tInle}, \textipa{\ae ske}, $\cdots$ };
%%     \draw [thick,->] (p2.south) -- (ABAvocabulary.north);
%%     \draw [thick,->] (w2.south) -- (ABAvocabulary.north);
%%     \node(ABAlexiconBox)[draw,ultra thick, fit = (p1) (w1) (ABAvocabulary)] {};
%%     \draw [ultra thick,->] (ABAmorphologyBox.south) -- (ABAlexiconBox.north);

%%     \node(crib) at (4.5,-8) {\verb|horn:| \textipa{rog}};
%%     \node(Cingular)[draw,align = left] at (3.5,-9.5) {\verb|singular:|\\ \textipa{rog}\nextForm{r_1}\\\textipa{rug}\nextForm{r_2}\\\textipa{ruk}};
%%     \node(plural)[draw,align = left] at (5.5,-9.5) {\verb|plural:|\\ \textipa{rogi}\nextForm{r_1}\\\textipa{rogi}\nextForm{r_2}\\\textipa{rogi}};
%%     \draw [thick,->] (crib.south) -- (Cingular.north);
%%     \draw [thick,->] (crib.south) -- (plural.north);

%%     \node[magenta,draw,align = left] (polishObservation) at (4.5,-11.5) {\verb|horn, singular:| \textipa{\v{z}wop}\\\verb|horn, plural:| \textipa{\v{z}wubi}};
%%     \draw [thick,->] (Cingular.south) -- (polishObservation.north);
%%     \draw [thick,->] (plural.south) -- (polishObservation.north);

%%     \node(polishLexicon)[draw,ultra thick, fit = (crib) (Cingular) (plural) (polishObservation)] {};
%%     \draw [ultra thick,->] (polishMorphologyBox.south) -- (polishLexicon.north);

%%   \end{tikzpicture}
%% \end{figure}

%\bibliographystyle{named}
\bibliographystyle{unsrt} \bibliography{main}

\pagebreak
\section{Appendix}

\begin{table*}[h]
\centering
\begin{tabular*}{\textwidth}{ll}
\toprule
Grammar rule  & English description \\\midrule
$\mathcal{P}\to \mathcal{R}\circ \mathcal{R}\circ \cdots \circ \mathcal{R}$&Phonology $\mathcal{P}$ is a sequence of compositions of rewrites $\mathcal{R}$\\
$\mathcal{R}\to \mathcal{F}\longrightarrow \mathcal{C} / \mathcal{T}\_\mathcal{T}$& Rewrite focus $\mathcal{F}$ to structural change $\mathcal{C}$ between triggers $\mathcal{T}$\\
$\mathcal{T}\to \# \mathcal{T}' | \mathcal{T}'$ & Triggers optionally match end of string\\
$\mathcal{T}'\to \epsilon | \mathcal{X} \mathcal{T}' | \mathcal{X}^* \mathcal{T}'$ & Triggers are sequences of matrices $\mathcal{X}$ possibly with Kleene star\\
$\mathcal{X}\to a|t|s|\cdots$&Matrices can be constant phonemes\\
$\mathcal{X}\to [\pm \mathcal{G} \; \pm \mathcal{G} \; \cdots \; \pm \mathcal{G}]$& Matrices check features $\mathcal{G}$\\
$\mathcal{X}\to \sigma$&Matrices can be whole syllables\\
$\mathcal{G}\to \text{voice}|\text{nasal}|\cdots$&Standard phonological features\\
$\mathcal{F}\to \mathcal{X}$&Focus can be a feature matrix\\
$\mathcal{F}\to \varnothing$&Focus can be empty (insertion rule)\\
$\mathcal{F}\to\mathbb{Z}$&Focus can be one of the triggers (copies it)\\
$\mathcal{C}\to\mathcal{X}$&Structural change can be a feature matrix\\
$\mathcal{C}\to\varnothing$&Deletion rule\\
\bottomrule
\end{tabular*}
\caption{Grammar over phonologies. The non-terminal $\mathcal{P}$ is the start symbol for the grammar.}
\label{phonologyGrammar}
\end{table*}

\begin{table*}[h]
\centering
\begin{tabular*}{\textwidth}{ll}
\toprule
Grammar rule  & English description \\\midrule
$\mathcal{W}\to (\mathcal{M}) \mathcal{M} (\mathcal{M})$&Morphology builds words $\mathcal{W}$ by concatenating morphemes $\mathcal{M}$. Prefix/Suffix optional (in parentheses)\\
$\mathcal{M}\to$ sequence of phonemes\bottomrule
\end{tabular*}
\caption{Grammar over morphologies. The non-terminal $\mathcal{W}$ is the start symbol for the grammar.}
\label{morphologyGrammar}
\end{table*}

\pagebreak
\subsection{Inference}

The solver takes as input:
\begin{itemize}
\item A set of inflected surface forms, $\{x_n\}$. Here the variable $x_n$ refers to all of the inflections of a word. So an example set of 3 inflected surface forms might be $\{(\text{fish},\text{fished}), (\text{walk},\text{walked}), (\text{eat},\text{ate})\}$.
\item A bound on the number of rules to consider synthesizing, $B_R$.
\end{itemize}
It produces as output:
\begin{itemize}
\item A set of underlying forms, $\{u_n\}$.
\item A morphology consisting of a prefix ($\{p_i\}$) and a suffix ($\{s_i\}$)  for each inflection
\item  A sequence of rules, $\{r_j\}_{j = 1}^{B_R}$
\end{itemize}
It is \emph{guaranteed} to discover the rules, morphology, and underlying forms minimizing:
\begin{equation}
  \text{solver}(\{x_n\},B_R) = \argmin_{\{r_j\}_{j = 1}^{B_R}, \{u_n\}, \{p_i\}, \{s_i\}} \sum_j \text{cost}(r_j) + \sum_i (\text{cost}(p_i) + \text{cost}(s_i) ) + \sum_n \text{cost}(u_n)
\end{equation}
In exchange for this strong guarantee, however, it has only a very
weak guarantee on the amount of time it will take to discover the
solution: in the worst case, it will take time exponential size of the
problem. In practice it takes close to a day for it to do exact search over 3 rules.
For easy problems, like the AGL experiments, I can get away with 4 rules if I wait for a few hours.

One can also ask the solver for the top $K$ solutions minimizing the cost function, which I will write:
\begin{equation}
  \text{solver}_K(\{x_n\},B_R) = \text{the }K\text{ minimum cost solutions.}
\end{equation}
One can also provide additional constraints to the solver, for example telling it what the underlying forms are,
or telling it what some of the rules are. I will write these constraints by adding an extra argument to the solver
that looks like conditioning in Bayesian inference. For example,
\begin{equation}
  \text{solver}_K(\{x_n\},B_R|u_1 = \text{fish}) = \text{the }K\text{ minimum cost solutions where }u_1\text{ has to be fish.}
  \end{equation}

So now the game is to leverage this solver as a ``black box'' inside of an incomplete search algorithm.
A key subroutine in my attempts at building such a search algorithm is
\texttt{solveIncrementalChange}, which takes as input an existing set of rules ($\{r_j^e\}_{j = 1}^{B_R}$)
along with a \textbf{search radius} ($R$),
and solves for a solution that changes at most \textbf{search radius} rules.
\begin{equation}
  \text{solveIncrementalChange}_K (\{x_n\}, \{r_j^e\}_{j = 1}^{B_R}, R)
   = \text{solver}_K (\{x_n\}, B_R+1 | r_j\not= r_j^e \text{ at most }R\text{ times})
\end{equation}
So \texttt{solveIncrementalChange} can add a rule (see $B_R+1$). I didn't build in any explicit mechanism for reordering rules -- if you set the search radius $R$ to 2, then it could effectively swap 2 rules.

In addition to synthesizing large programs, we also have to deal with
large data sets. Armando already came up with a nice trick for this called CEGIS\footnote{counterexample
  guided inductive synthesis}.
The idea behind CEGIS is that you synthesize a program from one example,
and then you check to see if that program is inconsistent with any other examples.
If not then you are done.
Otherwise, you have found a \textbf{counterexample}. You add the counterexample to your
training data set and repeat.

Combining \texttt{solveIncrementalChange} with CEGIS gives the
skeleton of the inference algorithm that I have been experimenting
with, called \texttt{incrementalCEGIS}. See the pseudocode at the top of the next page:
\pagebreak
\begin{algorithmic}
  \STATE \textbf{\texttt{incrementalCEGIS}}:
  \STATE {\bfseries Input:} Matrix of inflected forms for $N$ lexemes, $X$
  \STATE {\bfseries Output:} underlying forms, $\{u_n\}$; prefixes $\{p_i\}$; suffixes $\{s_i\}$; rules $\{r_j\}$
  \STATE // Initialize training set $T$ with a single example
  \STATE Set $T = $ inflections for any lexeme in $X$
  \STATE // Find a solution for this single example, using as many rules as you need ($\infty$)
  \STATE Set $\{r_j\}_{j = 1}^{B_R}, \{p_i\}, \{s_i\} = \text{solver}(T,\infty)$
  \REPEAT
  \STATE // Find the set of counterexamples $C$ inconsistent with our current guess as to what the solution is
  \STATE Set $C = \{x | x\in X \text{ and } x \text{ inconsistent with } \{r_j\}_{j = 1}^{B_R},\{p_i\}, \{s_i\}\}$
  \STATE // If there are no counterexamples as we are done
  \STATE \textbf{if} $C = \varnothing$ \textbf{ then return }$\{r_j\}_{j = 1}^{B_R},\{p_i\}, \{s_i\}$

  \STATE // Start out with a search radius of $1$ and increase it until we find a way of accommodating another example
  \FOR{$R = 1,2,3,4,...$}
  \FOR {$c\in C$}
  \STATE Set synthesisResult$ = \text{solveIncrementalChange}(T\cup\{c\},\{r_j\},R)$
  \IF{synthesisResult$\not= \bot$}
  \STATE Set $\{r_j\}_{j = 1}^{B_R}, \{p_i\}, \{s_i\} = $synthesisResult
  \STATE Add $c$ to the set of training examples, $T$
  \STATE \textbf{break out of loop over }$R$ // Successfully updated both our solution and added another example to the training set
\end{algorithmic}
\vspace{1cm}

One important feature of this approach is that it does not reuse
previous inferences about the morphology or the underlying forms.
I think we want to add this.
For example, we might allow it to change up to $R$ underlying forms,
or disallow changes in the morphology once the training set is big enough.

Variations of \textbf{incrementalCEGIS} that I have tried:
\begin{itemize}
\item[] Maintaining a population of solutions. Instead of taking the top 1 solution every time you synthesize, try taking the top $K$. There is some trade-off here between the cost of solving for more solutions and
  the benefit of not needing as many counterexamples. I have not done any systematic experiments,
  but my general sense is that this trade-off favored $K = 1$ in practice (so not maintaining a population of solutions).
\item[] Letting the solver insert the new rule anywhere vs constraining it to either append or prepended the new rule. Sketch uses up a lot of memory whenever I let it insert the rule anywhere -- not sure why this is.
\item[] Giving sketch more problems to solve but making each problem smaller. Here what I do is have an outer loop that iterates over all of the different ways you could edit an existing set of rules, and then invoke the solver for each of these. So for example I invoke the solver telling it to change only the first rule,
  then I invoke it telling it to change only the second rule, etc, finally telling it to keep all the rules the same but add a new rule.
\end{itemize}

\section{Related work}

Minimum generalization learner: \cite{Albright03rulesvs}. The interesting similarity here is that we both learned the structure of the rules. But it's restriction to the supervised case hampers its ability to really capture what a grammar is about (a generative model learned without supervision) and it misses out on lots of cool phenomena.

Maximum entropy learner:~\cite{goldwater2003learning}. Whether you take an OT perspective or a generative perspective, the structure of the rules (or constraints) has to come from somewhere. How do you learn that structure? This is a program induction problem.

Unsupervised phonological rule learners:
~\cite{cotterell-peng-eisner-2015} learns weighted fst's while
\cite{rulebased} learns context-sensitive rewrites. Here we show that
these results and more can be reformulated as a program induction
problem, and that doing so buys you (1) connections to probabilistic
programming and Bayesian program learning; (2) theoretical guarantees;
(3) a bunch more phenomena modeled, like supervised cases (pig latin),
UG, and infant studies.

References from a discussion with Adam Albright:

LISA (an acronym standing for something) babbling

Alex Cristia: Can infants learn phonology in the lab?

David Stampe: Spontaneous Devoicing. Evidence of UG

Chapter 8 of SPE~\cite{chomsky1968sound}: What is the right evaluation metric for rules?

Collin Wilson: paletization generalizations as more evidence of UG in phonology

Degree of articulatory effort as an inductive bias

Rules shouldn't be too disruptive of the underlying form. Should also minimize confuseability. Final devoicing is a good example of this. PMAP. ``weak contrast''

\section{Experiments}

\subsection{Phonological Rule Induction}

Several metrics of success here: (1) how well does it compress the
data? (2) how well does it predict held out forms? (3) how natural or
interpretable are the resulting solutions?

Here are cases where I think it learns something cool:

\noindent\begin{tabular*}{\textwidth}{lclll}
  \toprule
  Language&Example data&Underlying forms&Phonology&Phenomena\\\midrule
Gen&
\begin{tabular}{l}
\textipa{sra}\\
\textipa{agble}\\
\textipa{drE}\\
\textipa{hlE}
\end{tabular}
&\textipa{l} underlying
  & \textipa{l}$\to$\textipa{r}$/$\verb|[+coronal] _|
  &
\textipa{l}$\sim$\textipa{r} alternation
  \\
  \bottomrule  \end{tabular*}

\noindent\begin{tabular*}{\textwidth}{lclll}
  \toprule
  Language&Example data&Phonology&Morphology&Phenomena\\\midrule
  Tibetan&
\begin{tabular}{ll}
1&\textipa{\|x{j}ig}\\
4&\textipa{\|x{s}i}\\
5&\textipa{Na}\\
9&\textipa{gu}\\
10&\textipa{\|x{j}u}\\
11 ($= 10 + 1$) & \textipa{\|x{j}ug\|x{j}ig}\\
14 ($= 10 + 4$) & \textipa{\|x{j}ub\|x{s}i}\\
15 ($= 10 + 5$) & \textipa{\|x{j}uNa}\\
19 ($= 10 + 9$) & \textipa{\|x{j}urgu}\\
40 ($= 4 + 10$) & \textipa{\|x{s}ib\|x{j}u}\\
50 ($= 5 + 10$) & \textipa{Nab\|x{j}u}\\
90 ($= 9 + 10$) & \textipa{gub\|x{j}u}
\end{tabular}
&
\begin{tabular}{l}
  \verb|[-nasal]|$\to\varnothing $ $/$ \verb|# _|\\
  \emph{or,}\\
  \verb|C|$\to\varnothing $ $/$ \verb|# _ C|\\
  \end{tabular}
  &
\emph{given in problem}
  &
Consonant cluster reduction
  \\
  \bottomrule  \end{tabular*}

\begin{tabular*}{\textwidth}{lclll}
  \toprule
  Language&Example data&Phonology&Morphology&Phenomena\\\midrule
  Kerewe&
\begin{tabular}{llll}
\textipa{kubala} & \textipa{kubalana} & \textipa{kubalila} & \textipa{kubalilana}\\
\textipa{kugaya} & \textipa{kugayana} & \textipa{kugayila} & \textipa{kugayilana}\\
\textipa{kub\'ala} & \textipa{kub\'al\'ana} & \textipa{kub\'al\'ila} & \textipa{kub\'al\'ilana}\\\hline
\textipa{kut\'ub\'ala} & \textipa{kuk\'ib\'ala} & \textipa{kut\'ub\'alila} & \textipa{kuk\'it\'ubalila}\\
\textipa{kut\'ug\'aya} & \textipa{kuk\'ig\'aya} & \textipa{kut\'ug\'ayila} & \textipa{kuk\'it\'ugayila}\\
\textipa{kut\'ub\'ala} & \textipa{kuk\'ib\'ala} & \textipa{kut\'ub\'alila} & \textipa{kuk\'it\'ubalila}
\end{tabular}
&
\begin{tabular}{l}
  \verb|[]|$\to$\verb|[-H]|$/$\verb|[+H] [] _| \\
  \verb|[]|$\to$\verb|[+H]|$/$\verb|[+H] [] _ []|
  \end{tabular}
  &
\begin{tabular}{ll}
\textipa{ku}$+$stem$+$\textipa{a}\\
\textipa{ku}$+$stem$+$\textipa{ana}\\
\textipa{ku}$+$stem$+$\textipa{ila}\\
\textipa{ku}$+$stem$+$\textipa{ilana}\\
\textipa{kut\'u}$+$stem$+$\textipa{a}\\
\textipa{kuk\'i}$+$stem$+$\textipa{a}\\
\textipa{kut\'u}$+$stem$+$\textipa{ila}\\
\textipa{kuk\'it\'u}$+$stem$+$\textipa{ila}
  \end{tabular}
  &
\begin{tabular}{ll}
  Tone spreading\\
  Rule ordering
  \end{tabular}
  \\
  \bottomrule  \end{tabular*}

\noindent\begin{tabular*}{\textwidth}{lclll}
  \toprule
  Language&Example data&Phonology&Morphology&Phenomena\\\midrule
  Polish&
\begin{tabular}{ll}
\textipa{dom} & \textipa{domi}\\
\textipa{kot} & \textipa{koti}\\
\textipa{lut} & \textipa{lodi}\\
\textipa{vus} & \textipa{vozi}\\
\textipa{wuk} & \textipa{wugi}\\
\textipa{ruk} & \textipa{rogi}\\
\textipa{bur} & \textipa{bori}\\
\textipa{\|x{s}um} & \textipa{\|x{s}umi}
\end{tabular}
&\begin{tabular}{l}
   \textipa{o}$\to$\textipa{u}$/$\verb|_[-nasal +voice]#|\\%
\verb|[-sonorant]|$\to$\verb|[-voice]|$/$\verb|_#|
   \end{tabular}
  &
\begin{tabular}{ll}
  stem\\
  stem$ + $\textipa{i}
  \end{tabular}
  &
\begin{tabular}{ll}
  Final devoicing\\
  Rule ordering
  \end{tabular}
  \\
  \bottomrule  \end{tabular*}

\noindent\begin{tabular*}{\textwidth}{lclll}
  \toprule
  Language&Example data&Phonology&Morphology&Phenomena\\\midrule
Makonde  &
\begin{tabular}{lll}
\textipa{am\'aNga} & \textipa{am\'ile} & \textipa{\'ama}\\
\textipa{ak\'aNga} & \textipa{ak\'ile} & \textipa{\'aka}\\
\textipa{av\'aNga} & \textipa{av\'ile} & \textipa{\'ova}\\
\textipa{am\'aNga} & \textipa{am\'ile} & \textipa{\'oma}\\
\textipa{ut\'aNga} & \textipa{ut\'ile} & \textipa{\'uta}\\
\textipa{av\'aNga} & \textipa{av\'ile} & \textipa{\'eva}\\
\textipa{tav\'aNga} & \textipa{tav\'ile} & \textipa{t\'ava}\\
\textipa{uNg\'aNga} & \textipa{uNg\'ile} & \textipa{\'uNga}\\
\textipa{pat\'aNga} & \textipa{pat\'ile} & \textipa{p\'ota}
\end{tabular}
&
\begin{tabular}{l}
\verb|[]|$\to$\verb|[-stress]| $/$ \verb| _ [ ]* [ +stress]|\\
\verb|[+middle -stress]|$\to$\textipa{a}$/$\verb| _ [ ]|
  \end{tabular}
  &
\begin{tabular}{ll}
  stem$ + $\textipa{\'aNga}\\
  stem$ + $\textipa{\'ile}\\
  stem$ + $\textipa{a}
  \end{tabular}
  &
\begin{tabular}{ll}
Stress patterns\\
Neutralizing rule
  \end{tabular}
  \\
  \bottomrule  \end{tabular*}

\noindent\begin{tabular*}{\textwidth}{lclll}
  \toprule
  Language&Example data&Phonology&Morphology&Phenomena\\\midrule
  Kikuria&
\begin{tabular}{ll}
\textipa{rema} & \textipa{remera}\\
\textipa{roma} & \textipa{romera}\\
\textipa{tiga} & \textipa{tegera}\\
\textipa{ruga} & \textipa{rogera}\\
\textipa{hoora} & \textipa{hoorera}\\
\textipa{siika} & \textipa{seekera}\\
\textipa{huuta} & \textipa{hootera}\\
\textipa{suraaNga} & \textipa{suraaNgera}
\end{tabular}
&
\verb|[+high]|$\to$\verb|[+middle]|$ /$\verb| _ [-low]* [+middle]|
  &
\begin{tabular}{ll}
  stem$ + $\textipa{a}\\
  stem$ + $\textipa{era}
  \end{tabular}
  &
Vowel harmony
  \\
  \bottomrule  \end{tabular*}

\noindent\begin{tabular*}{\textwidth}{lclll}
  \toprule
  Language&Example data&Phonology&Morphology&Phenomena\\\midrule
  Hungarian&
  \begin{tabular}{llll}
\textipa{a:g\super y} & \textipa{a:g\super yban} & \textipa{a:k\super yto:l} & \textipa{a:g\super ynak}\\
\textipa{\"o:r} & \textipa{\"o:rben} & \textipa{\"o:rt\"o:l} & \textipa{\"o:rnek}\\
\textipa{ku:t} & \textipa{ku:dban} & \textipa{ku:tto:l} & \textipa{ku:tnak}\\
\textipa{re:s} & \textipa{re:zben} & \textipa{re:st\"o:l} & \textipa{re:snek}\\
\textipa{rab} & \textipa{rabban} & \textipa{rapto:l} & \textipa{rabnak}\\
\textipa{vi:z} & \textipa{vi:zben} & \textipa{vi:st\"o:l} & \textipa{vi:znek}\\
\textipa{fal} & \textipa{falban} & \textipa{falto:l} & \textipa{falnak}\\
\textipa{test} & \textipa{tezdben} & \textipa{testt\"o:l} & \textipa{testnek}
  \end{tabular}&
  \begin{tabular}{l}
    \verb|V|$\to$\verb|[+mid +tense +front]|$ /$\\\verb|  [ +front ] [ ]* _ |\\
         \verb|[ ]|$\to$\verb|[+voice]|$ / $\\\verb| _ [+bilabial]|\\
         \verb|[-sonorant]|$\to$\verb|[-voice]|$ / $\\\verb| _ [-voice]|
    \end{tabular}
  &
  \begin{tabular}{l}
    stem\\
    stem$ + $\textipa{ban}\\
    stem$ + $\textipa{to:l}\\
    stem$ + $\textipa{nak}
  \end{tabular}
  &
  \begin{tabular}{l}
    Vowel harmony\\
    Voicing assimilation\\
    Rule ordering
    \end{tabular}
  \\
\bottomrule  \end{tabular*}

\pagebreak
These problems are solved:
\begin{itemize}
\item Kikurai: fricatives alternate with stops after nasals
\item Greek: Velar stops alternate with palletized versions before front vowels
\item Farsi: Trills alternate with flaps
\item Osage: coronal stops become dentals before central vowels
\item Amharic: alternation between \textipa{@} \& \textipa{E}
\item Gen: l/r alternation
\item Kishambaa: voiced/unvoiced nasals alternate
\item  Thai: stops are unreleased word finally
\item Palauan: a word initial neutralizing rule
\item Quechua: Velar becomes uvular when followed by uvular (spreading)
\item Lhasa Tibetan: no contrast between velar/uvular, or voiced/voiceless stops or fricatives
\item Axininca Campa: stops become glides
\item Kikuyu: infinitive prefix can surface as either k or \textipa{G}
\item Korean: vowel harmony, aspiration only surfaces in certain contexts
\item Hungarian: vowel harmony and voicing assimilation
\item Kikuria: vowel harmony
\item Farsi: a deletion rule explains singular/plural
\item Tibetan: initial consonant cluster reduction explains counting system
\item Makonde: stress patterns; unstressed vowels are neutralized
\item North Saami: 3 neutralization rules explain nominative sg essive
\item Samoan: 2 deletion rules that explain words that sound the same in one inflection being different in another
\item     Russian: devoicing of word final obstruent
\item English: verbal inflections (voicing assimilation, epenthesis)
\item Finnish: nominative/partive explained by vowel raising and vowel harmony
\item Kerewe: Interacting tone rules
\item Polish: vowel alternations interacting with devoicing of word final obstruent in singular/plural
\item Ancient Greek: voicing assimilation, deaspiration, deletion rule. Order matters.
\item Serbo-Croatian: predictable stress, devoicing, neutralizing, epenthesis
\end{itemize}
17 textbook problems are not solved. 2 of these need a tier-based representation. A couple more of these are due to limitations in the program learner's representation of rules. The rest are because they involve scaling to more than 3 rules, which is about the practical limit for exact search with these methods.
\pagebreak

\subsubsection{An approximation to UG: Program fragments}

Within-rule reuse of structure:
\begin{tabular}{ll}
  Tibetan: & {\color{red}\texttt{C}}$\to\varnothing /$\verb|#_|{\color{red}\texttt{C}} \\
  English: &          \verb|[-sonorant]|$\to${\color{red}\texttt{[-voice]}}$ / ${\color{red}\texttt{[-voice]}}\verb| _ |\\
Kikuria: &\verb|[ +high ]|$\to$ {\color{red}\texttt{[ +middle ]}} $/$ \verb| _ [ -low ]*| {\color{red}\texttt{[ +middle ]}}
\end{tabular}

\vspace{0.5cm}\noindent Within-language reuse of structure:
\begin{tabular}{ll}
  Axininca Campa: &
  \begin{tabular}{l}
    \textipa{p}{\color{red}$\to$}\textipa{w} {\color{red}$/$ \texttt{[ ] \_}}\\
    \textipa{k}{\color{red}$\to$}\textipa{y}\, {\color{red}$/$ \texttt{[ ] \_}}
  \end{tabular}\\
  English:&
  \begin{tabular}{l}
    {\color{red}$\varnothing\to$\textipa{@} $/$} \verb|[+sib] _ [+sib]|\\
    {\color{red}$\varnothing\to$\textipa{@} $/$} \verb|[+cor +stop] _ [+cor +stop]|
  \end{tabular}
\end{tabular}

\vspace{0.5cm}\noindent Cross-language reuse of structure:
\begin{tabular}{ll}
  Russian/Polish:& \verb|[-sonorant]|$\to$\verb|[-voice]|$/$\verb|_#|\\
  Samoan/Tibetan:&
  \begin{tabular}{l}
    {\color{red}\texttt{C}$\to\varnothing /$}\verb|_#|\\
    {\color{red}\texttt{C}$\to\varnothing /$}\verb|#_C|
  \end{tabular}\\
  Ancient Greek/Hungarian:&
  \begin{tabular}{l}
    {\color{red}\texttt{[-sonorant]}$\to$\texttt{[-voice}} \texttt{-aspirated]} {\color{red} $/$ \texttt{ \_ [-voice]}}\\
    {\color{red}\texttt{[-sonorant]}$\to$\texttt{[-voice}} \texttt{]} {\color{red} $/$ \texttt{ \_ [-voice]}}
    \end{tabular}
  \end{tabular}

\begin{figure}[h]
  \includegraphics[width = \textwidth]{featureFrequencies.png}
  \caption{Which features are frequently used in correct rules? Voicing, sonorant, nasality are very frequent, and are also near the top of the feature geometry hierarchy. Feature geometry was not given to the system, but you can see at least this coarse level of structure emerge from the data. Vowel features also frequent due to vowel harmony occurring in lots of languages.}
\end{figure}

\begin{figure}[h]
  \includegraphics[width = \textwidth]{skeletonFrequencies.png}
  \caption{What is the general structure of rules that occur in phonology, ignoring the particular constants that occur in the rule? Here $K$ stands in for an arbitrary phoneme; $\pm$features stands in for an arbitrary feature matrix. From this plot you can glean that (1) neutralizing rules are very common (top two most frequent rules); (2) harmony/assimilation rules are also frequent; (4) deletion rules tend to occur at word boundaries. There are about $5\times 10^4$ possible syntactic structures for rules, but only 15 are actually attested (relative frequency$ > 1$). Here I think that the sparsity of the problem sets is hurting us: these results come from 26 problems, so the fact that you see only 15 general schemas isn't too crazy if you don't believe in strong innate constraints on the structure of rules. But looking at this data you also see a generalization that holds across all of the rules learned: the structures tend to be pretty small, so you never see a rule that checks feature values of more than two adjacent segments.}
\end{figure}

%% \subsection{Historical productivity}

%% There is an interesting tension between discovering regularities
%% that are psychologically real (like things that are used productively)
%% and discovering regularities that are in the data but only due to
%% historical accidents. Our system cannot distinguish them.

%% Are these deep analyses real? We don't answer this question, but we do
%% raise it. We are agnostic. These are the kinds of regularities that a
%% good linguist would notice, and our system notices them too.

\subsection{Artificial Phonologies}

It would be cool to model all of the grammars in~\cite{frank2011three}. Right now we can model AAB/ABA patterns like in Gary Marcus's famous infant experiments. Other infant studies with artificial grammars:~\cite{gerken2006decisions}. MNPQ problem:~\cite{smith1966grammatical}.

Analogs in real languages: agreement morphology (eg gender in Spanish
- there is a constraint that two things have to be the same);
reduplication (this could literally be modeled using the same ABA
system); obligatory contour principle (opposite of harmony - two
things have to be different, just like the ssd/sds distinction).

For any of these problems, there is a trade-off between how
complicated of a program you are willing to use and how badly you want
to compress the data. Here, compressing the data means representing
the surface forms with short underlying representations. This
trade-off is of course present in the phonology problem sets as well.
But the problem sets are complicated and have lots of data, precluding
any in-depth analysis of the trade-off except quantitatively. For
these synthetic grammar learning problems, we can dig in and see
exactly what the trade-off looks like for different amounts of data.
See figure~\ref{fronts} for an analysis of this trade-off along with
examples of what the programs look like.

\section{Applications}
The point of this is to motivate why you would want to solve exactly
this problem even though it doesn't do anything more immediately
useful than solve your phonology homework. Practical applications that could come from this line of work:

\begin{itemize}
\item Low-resource language acquisition: \url{http://sapience.dec.ens.fr/bootphon/} for some languages there is very little data and very few speakers
\item Mathematica for Phonology: a lot of the time physicists don't do
  integrals; they just punch them into Mathematica. This is a tool
  that provably finds the ``best'' grammar for some data, sort of like
  how Mathematica finds the correct integral of some expression. And
  just like some integrals are beyond the reach of computer algebra
  systems, some data sets might be beyond the reach of this
  hypothetical tool. Analogy to the work
  in~\cite{schmidt2009distilling}. According to Tim it would be more
  practical for the field for knowledge is to have a system that can
  work out the consequences of a given rule set, maybe pointing out
  places where it's inconsistent with the data or suggesting
  underlying forms. These alternative uses could also be accommodated.
\item A probe for UG: the prior over phonology is a parameter of the
  algorithm which you could in principle estimate from the data or
  compare different settings of the prior to see how well they
  compress the data. Or, come up with the right constraints so that
  the system doesn't generalize in the same places where humans don't
  generalize (surfeit of the stimulus:~\cite{becker2011surfeit}). You
  can approach universal grammar from two directions: what do I need
  to build then so that it will learn the same things that humans
  learn? And what do I need to take out in order to stop it from
  learning things that humans never learn? 
\end{itemize}

\section{Discussion}

Deep learnability issue: what is the actual computational complexity
of learning these grammars? The actual hardness is an empirical
question which you can upper bound by coming up with better heuristics
for inducing grammars. Attempts at \emph{lower bounding} the
complexity of learning has generally been way too conservative. We
want to know the actual hardness of the learning problem, and
constructing a practical program learner shows that it might not be
too hard with the right UG.

We must emphasize that there is \textbf{no general-purpose
  learner}. So we need UG. This tool lets us dig in and study it.
There's always going to be a trade-off between how much you build in and how much data/computation is needed to recover the correct generalizations. Our compromise in the space of this trade-off is to:
\begin{itemize}
\item  Not play in the space of arbitrary turing machines but instead search through a finite program space (bounded program size runtime and memory consumption; program structure given by a grammar)
  \item Make contentfull proposals about what the structure of
    universal grammar might be, and test them empirically on many
    languages
\end{itemize}

\textbf{Comparison with deep learning:} In a sense this project is like
the opposite of deep learning. It's about the kind of learning that
yields explanations of the causal processes behind data sets; where
the learner comes equipped with explicit background knowledge (UG)
that we can tweak finely and systematically.

One similarity though is that some of the phenomena we model, like
infant learning of synthetic grammars, are the sort of things that
humans pick up unconsciously. A lot of deep learning might be
analogous to humans rote learning some motor or perceptual task.

\end{document}

### 2018-02-20 paper.tex
initial commit

% Use only LaTeX2e, calling the article.cls class and 12-point type.

\documentclass[12pt]{article}

% Users of the {thebibliography} environment or BibTeX should use the
% scicite.sty package, downloadable from *Science* at
% www.sciencemag.org/about/authors/prep/TeX_help/ .
% This package should properly format in-text
% reference calls and reference-list numbers.

\usepackage{scicite}

% Use times if you have the font installed; otherwise, comment out the
% following line.

\usepackage{times}

% The preamble here sets up a lot of new/revised commands and
% environments.  It's annoying, but please do *not* try to strip these
% out into a separate .sty file (which could lead to the loss of some
% information when we convert the file to other formats).  Instead, keep
% them in the preamble of your main LaTeX source file.

% The following parameters seem to provide a reasonable page setup.

\topmargin 0.0cm
\oddsidemargin 0.2cm
\textwidth 16cm 
\textheight 21cm
\footskip 1.0cm

%The next command sets up an environment for the abstract to your paper.

\newenvironment{sciabstract}{%
\begin{quote} \bf}
{\end{quote}}

% If your reference list includes text notes as well as references,
% include the following line; otherwise, comment it out.

\renewcommand\refname{References and Notes}

% The following lines set up an environment for the last note in the
% reference list, which commonly includes acknowledgments of funding,
% help, etc.  It's intended for users of BibTeX or the {thebibliography}
% environment.  Users who are hand-coding their references at the end
% using a list environment such as {enumerate} can simply add another
% item at the end, and it will be numbered automatically.

\newcounter{lastnote}
\newenvironment{scilastnote}{%
\setcounter{lastnote}{\value{enumiv}}%
\addtocounter{lastnote}{+1}%
\begin{list}%
{\arabic{lastnote}.}
{\setlength{\leftmargin}{.22in}}
{\setlength{\labelsep}{.5em}}}
{\end{list}}

% Include your paper's title here

\title{Discovering the Rules of Language via Bayesian Program Learning}

% Place the author information here.  Please hand-code the contact
% information and notecalls; do *not* use \footnote commands.  Let the
% author contact information appear immediately below the author names
% as shown.  We would also prefer that you don't change the type-size
% settings shown here.

\author
{John Smith,$^{1\ast}$ Jane Doe,$^{1}$ Joe Scientist$^{2}$\\
\\
\normalsize{$^{1}$Department of Chemistry, University of Wherever,}\\
\normalsize{An Unknown Address, Wherever, ST 00000, USA}\\
\normalsize{$^{2}$Another Unknown Address, Palookaville, ST 99999, USA}\\
\\
\normalsize{$^\ast$To whom correspondence should be addressed; E-mail:  jsmith@wherever.edu.}
}

% Include the date command, but leave its argument blank.

\date{}

%%%%%%%%%%%%%%%%% END OF PREAMBLE %%%%%%%%%%%%%%%%

\begin{document}

% Double-space the manuscript.

\baselineskip24pt

% Make the title.

\maketitle

% Place your abstract within the special {sciabstract} environment.

\begin{sciabstract}
  This document presents a number of hints about how to set up your
  {\it Science\/} paper in \LaTeX\ .  We provide a template file,
  \texttt{scifile.tex}, that you can use to set up the \LaTeX\ source
  for your article.  An example of the style is the special
  \texttt{\{sciabstract\}} environment used to set up the abstract you
  see here.
\end{sciabstract}

% In setting up this template for *Science* papers, we've used both
% the \section* command and the \paragraph* command for topical
% divisions.  Which you use will of course depend on the type of paper
% you're writing.  Review Articles tend to have displayed headings, for
% which \section* is more appropriate; Research Articles, when they have
% formal topical divisions at all, tend to signal them with bold text
% that runs into the paragraph, for which \paragraph* is the right
% choice.  Either way, use the asterisk (*) modifier, as shown, to
% suppress numbering.

\section*{Introduction}

In this file, we present some tips and sample mark-up to assure your
\LaTeX\ file of the smoothest possible journey from review manuscript
to published {\it Science\/} paper.  We focus here particularly on
issues related to style files, citation, and math, tables, and
figures, as those tend to be the biggest sticking points.  Please use
the source file for this document, \texttt{scifile.tex}, as a template
for your manuscript, cutting and pasting your content into the file at
the appropriate places.

{\it Science\/}'s publication workflow relies on Microsoft Word.  To
translate \LaTeX\ files into Word, we use an intermediate MS-DOS
routine \cite{tth} that converts the \TeX\ source into HTML\@.  The
routine is generally robust, but it works best if the source document
is clean \LaTeX\ without a significant freight of local macros or
\texttt{.sty} files.  Use of the source file \texttt{scifile.tex} as a
template, and calling {\it only\/} the \texttt{.sty} and \texttt{.bst}
files specifically mentioned here, will generate a manuscript that
should be eminently reviewable, and yet will allow your paper to
proceed quickly into our production flow upon acceptance \cite{use2e}.

\section*{Formatting Citations}

Citations can be handled in one of three ways.  The most
straightforward (albeit labor-intensive) would be to hardwire your
citations into your \LaTeX\ source, as you would if you were using an
ordinary word processor.  Thus, your code might look something like
this:

\begin{quote}
\begin{verbatim}
However, this record of the solar nebula may have been
partly erased by the complex history of the meteorite
parent bodies, which includes collision-induced shock,
thermal metamorphism, and aqueous alteration
({\it 1, 2, 5--7\/}).
\end{verbatim}
\end{quote}

\noindent Compiled, the last two lines of the code above, of course, would give notecalls in {\it Science\/} style:

\begin{quote}
\ldots thermal metamorphism, and aqueous alteration ({\it 1, 2, 5--7\/}).
\end{quote}

Under the same logic, the author could set up his or her reference list as a simple enumeration,

\begin{quote}
\begin{verbatim}
{\bf References and Notes}

\begin{enumerate}
\item G. Gamow, {\it The Constitution of Atomic Nuclei
and Radioactivity\/} (Oxford Univ. Press, New York, 1931).
\item W. Heisenberg and W. Pauli, {\it Zeitschr.\ f.\ 
Physik\/} {\bf 56}, 1 (1929).
\end{enumerate}
\end{verbatim}
\end{quote}

\noindent yielding

\begin{quote}
{\bf References and Notes}

\begin{enumerate}
\item G. Gamow, {\it The Constitution of Atomic Nuclei and
Radioactivity\/} (Oxford Univ. Press, New York, 1931).
\item W. Heisenberg and W. Pauli, {\it Zeitschr.\ f.\ Physik} {\bf 56},
1 (1929).
\end{enumerate}
\end{quote}

That's not a solution that's likely to appeal to everyone, however ---
especially not to users of B{\small{IB}}\TeX\ \cite{inclme}.  If you
are a B{\small{IB}}\TeX\ user, we suggest that you use the
\texttt{Science.bst} bibliography style file and the
\texttt{scicite.sty} package, both of which we are downloadable from our author help site
(http://www.sciencemag.org/about/authors/prep/TeX_help/).  You can also
generate your reference lists by using the list environment
\texttt{\{thebibliography\}} at the end of your source document; here
again, you may find the \texttt{scicite.sty} file useful.

Whether you use B{\small{IB}}\TeX\ or \texttt{\{thebibliography\}}, be
very careful about how you set up your in-text reference calls and
notecalls.  In particular, observe the following requirements:

\begin{enumerate}
\item Please follow the style for references outlined at our author
  help site and embodied in recent issues of {\it Science}.  Each
  citation number should refer to a single reference; please do not
  concatenate several references under a single number.
\item Please cite your references and notes in text {\it only\/} using
  the standard \LaTeX\ \verb+\cite+ command, not another command
  driven by outside macros.
\item Please separate multiple citations within a single \verb+\cite+
  command using commas only; there should be {\it no space\/}
  between reference keynames.  That is, if you are citing two
  papers whose bibliography keys are \texttt{keyname1} and
  \texttt{keyname2}, the in-text cite should read
  \verb+\cite{keyname1,keyname2}+, {\it not\/}
  \verb+\cite{keyname1, keyname2}+.
\end{enumerate}

\noindent Failure to follow these guidelines could lead
to the omission of the references in an accepted paper when the source
file is translated to Word via HTML.

\section*{Handling Math, Tables, and Figures}

Following are a few things to keep in mind in coding equations,
tables, and figures for submission to {\it Science}.

\paragraph*{In-line math.}  The utility that we use for converting
from \LaTeX\ to HTML handles in-line math relatively well.  It is best
to avoid using built-up fractions in in-line equations, and going for
the more boring ``slash'' presentation whenever possible --- that is,
for \verb+$a/b$+ (which comes out as $a/b$) rather than
\verb+$\frac{a}{b}$+ (which compiles as $\frac{a}{b}$).  Likewise,
HTML isn't tooled to handle certain overaccented special characters
in-line; for $\hat{\alpha}$ (coded \verb+$\hat{\alpha}$+), for
example, the HTML translation code will return [\^{}$(\alpha)$].
Don't drive yourself crazy --- but if it's possible to avoid such
constructs, please do so.  Please do not code arrays or matrices as
in-line math; display them instead.  And please keep your coding as
\TeX-y as possible --- avoid using specialized math macro packages
like \texttt{amstex.sty}.

\paragraph*{Displayed math.} Our HTML converter sets up \TeX\
displayed equations using nested HTML tables.  That works well for an
HTML presentation, but Word chokes when it comes across a nested
table in an HTML file.  We surmount that problem by simply cutting the
displayed equations out of the HTML before it's imported into Word,
and then replacing them in the Word document using either images or
equations generated by a Word equation editor.  Strictly speaking,
this procedure doesn't bear on how you should prepare your manuscript
--- although, for reasons best consigned to a note \cite{nattex}, we'd
prefer that you use native \TeX\ commands within displayed-math
environments, rather than \LaTeX\ sub-environments.

\paragraph*{Tables.}  The HTML converter that we use seems to handle
reasonably well simple tables generated using the \LaTeX\
\texttt{\{tabular\}} environment.  For very complicated tables, you
may want to consider generating them in a word processing program and
including them as a separate file.

\paragraph*{Figures.}  Figure callouts within the text should not be
in the form of \LaTeX\ references, but should simply be typed in ---
that is, \verb+(Fig. 1)+ rather than \verb+\ref{fig1}+.  For the
figures themselves, treatment can differ depending on whether the
manuscript is an initial submission or a final revision for acceptance
and publication.  For an initial submission and review copy, you can
use the \LaTeX\ \verb+{figure}+ environment and the
\verb+\includegraphics+ command to include your PostScript figures at
the end of the compiled PostScript file.  For the final revision,
however, the \verb+{figure}+ environment should {\it not\/} be used;
instead, the figure captions themselves should be typed in as regular
text at the end of the source file (an example is included here), and
the figures should be uploaded separately according to the Art
Department's instructions.

\section*{What to Send In}

What you should send to {\it Science\/} will depend on the stage your manuscript is in:

\begin{itemize}
\item {\bf Important:} If you're sending in the initial submission of
  your manuscript (that is, the copy for evaluation and peer review),
  please send in {\it only\/} a PostScript or PDF version of the
  compiled file (including figures).  Please do not send in the \TeX\ 
  source, \texttt{.sty}, \texttt{.bbl}, or other associated files with
  your initial submission.  (For more information, please see the
  instructions at our Web submission site,
  http://www.submit2science.org/ .)
\item When the time comes for you to send in your revised final
  manuscript (i.e., after peer review), we require that you include
  all source files and generated files in your upload.  Thus, if the
  name of your main source document is \texttt{ltxfile.tex}, you
  need to include:
\begin{itemize}
\item \texttt{ltxfile.tex}.
\item \texttt{ltxfile.aux}, the auxilliary file generated by the
  compilation.
\item A PostScript file (compiled using \texttt{dvips} or some other
  driver) of the \texttt{.dvi} file generated from
  \texttt{ltxfile.tex}, or a PDF file distilled from that
  PostScript.  You do not need to include the actual \texttt{.dvi}
  file in your upload.
\item From B{\small{IB}}\TeX\ users, your bibliography (\texttt{.bib})
  file, {\it and\/} the generated file \texttt{ltxfile.bbl} created
  when you run B{\small{IB}}\TeX.
\item Any additional \texttt{.sty} and \texttt{.bst} files called by
  the source code (though, for reasons noted earlier, we {\it
    strongly\/} discourage the use of such files beyond those
  mentioned in this document).
\end{itemize}
\end{itemize}

% Your references go at the end of the main text, and before the
% figures.  For this document we've used BibTeX, the .bib file
% scibib.bib, and the .bst file Science.bst.  The package scicite.sty
% was included to format the reference numbers according to *Science*
% style.

\bibliography{scibib}

\bibliographystyle{Science}

% Following is a new environment, {scilastnote}, that's defined in the
% preamble and that allows authors to add a reference at the end of the
% list that's not signaled in the text; such references are used in
% *Science* for acknowledgments of funding, help, etc.

\begin{scilastnote}
\item We've included in the template file \texttt{scifile.tex} a new
environment, \texttt{\{scilastnote\}}, that generates a numbered final
citation without a corresponding signal in the text.  This environment
can be used to generate a final numbered reference containing
acknowledgments, sources of funding, and the like, per {\it Science\/}
style.
\end{scilastnote}

% For your review copy (i.e., the file you initially send in for
% evaluation), you can use the {figure} environment and the
% \includegraphics command to stream your figures into the text, placing
% all figures at the end.  For the final, revised manuscript for
% acceptance and production, however, PostScript or other graphics
% should not be streamed into your compliled file.  Instead, set
% captions as simple paragraphs (with a \noindent tag), setting them
% off from the rest of the text with a \clearpage as shown  below, and
% submit figures as separate files according to the Art Department's
% instructions.

\clearpage

\noindent {\bf Fig. 1.} Please do not use figure environments to set
up your figures in the final (post-peer-review) draft, do not include graphics in your
source code, and do not cite figures in the text using \LaTeX\
\verb+\ref+ commands.  Instead, simply refer to the figure numbers in
the text per {\it Science\/} style, and include the list of captions at
the end of the document, coded as ordinary paragraphs as shown in the
\texttt{scifile.tex} template file.  Your actual figure files should
be submitted separately.

\end{document}

### 2018-02-20 talk.tex
initial commit

\documentclass{beamer}

\usepackage{tikz}
\usetikzlibrary{fit}
\usepackage{color}
\usepackage{booktabs}
\usepackage{tipa}
\usepackage{amssymb}
\usepackage{verbatim}
\usepackage[absolute,overlay]{textpos}
\usepackage{pifont}% http://ctan.org/pkg/pifont
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\usetikzlibrary{bayesnet}
\usetikzlibrary{decorations.markings}

\newcommand{\nextForm}[1]{\rotatebox[origin=c]{270}{$_{\curvearrowright}$}$_{#1}$}

\title{Inducing Phonological Rules: Perspectives from Bayesian Program Learning}
\author{Kevin Ellis and Timothy O'Donnell and Joshua B. Tenenbaum and Armando Solar-Lezama}
\institute{MIT}
\date{2017}

\begin{document}

\frame{\titlepage}
%% \begin{frame}{Abstract}
%%     How do linguists come up with phonological rules? How do kids
%%     learn pig latin?  We develop computational models of these and
%%     other phenomena in language.  The unifying element of the models
%%     is to treat grammars as programs, which lets us apply ideas from
%%     the field of program synthesis to learn grammars. This lets the
%%     models capture phonological phenomena like vowel harmony or stress
%%     patterns and learn synthetic grammars used in prior studies of
%%     rule learning.  Going beyond individual grammar learning problems,
%%     we consider the problem of jointly inferring many related rule
%%     systems. By solving many textbook phonology problems, we can
%%     ask the model what kind of inductive bias best explains the
%%     attested phenomena.
%% \end{frame}

\begin{frame}{A language learning experiment}
  %  \fontsize{20}{7.2}\selectfont
  \Huge
  \centering\begin{tabular}{ll}
    \textipa{xau}&\visible<2->{\textipa{xauxaud@}}\\
    \textipa{man}&\visible<2->{\textipa{manmand@}}\\
    \textipa{kwaj}&\visible<2->{\textipa{kwajkwajd@}}\\
    \textipa{\c{c}in}&    \visible<2->{\textipa{\c{c}in\c{c}ind@}}\\
    \visible<3->{\textipa{lej}}&\visible<4->{\textipa{lejlejd@}}

  \end{tabular}
  \\\vspace{1cm}

  \visible<5->{\color{red}\texttt{A}$\to$\texttt{AA}$+$\textipa{d@}}

\end{frame}

\begin{frame}[t]{A language learning experiment}
  \begin{columns}

    \begin{column}{0.4\textwidth}

  \begin{center}
    \huge
    \begin{tabular}{ll}
      \textipa{dom} & \visible<2->{\textipa{domi}}\\
      \textipa{kot} & \visible<2->{\textipa{koti}}\\
      \textipa{lut} & \visible<2->{\textipa{lodi}}\\
      \textipa{vus} & \visible<2->{\textipa{vozi}}\\
      \textipa{wuk} & \visible<2->{\textipa{wugi}}\\
      \textipa{bur} & \visible<2->{\textipa{bori}}\\
      \visible<4->{\textipa{ruk}} & \visible<3->{\textipa{rogi}}\\
    \end{tabular}
  \end{center}
    \end{column}
        \begin{column}{0.6\textwidth}
        \visible<5->{\color{red}\Large  \texttt{A}$\to$\texttt{A}$ + /\text{i}/$

  \textipa{o}$\to$\textipa{u}$/$ \_ [-nasal +voice]\#

  [-sonorant]$\to$[-voice] $/$ \_ \#
  }
    \end{column}
    \end{columns}

\end{frame}

%% \begin{frame}{Learning to build and pronounce words}
%%   dog: [\textipa{dag}]\\
%%   dogs: [\textipa{dagz}]

%%   \vspace{1cm}\pause

%%   cat: [\textipa{k\ae t}]\\
%%   cats: [\textipa{k\ae ts}]\\

%%   \vspace{1cm}\pause

%%   /\textipa{dag}/+/\textipa{z}/ $\to$ /\textipa{dagz}/  $\to$  [\textipa{dagz}]

%%   /\textipa{k\ae t}/+/\textipa{z}/ $\to$ /\textipa{k\ae tz}/  $\to$  [\textipa{k\ae ts}]

%%   \vspace{0.5cm}

%%   z$\to$s / [-voice] \_ \#

%% \end{frame}

\begin{frame}{ABA (same/different/same)}
\centering\Huge  \begin{tabular}{l}
    wofewo\\
    lovilo\\
    fimufi
  \end{tabular}

  \pause

\vspace{1cm}  $\varnothing\to\sigma_i$ / $\sigma_i\sigma$ \_

  \vspace{1cm}

  /wofe/ $\to$ [wofewo]

\end{frame}

%% \begin{frame}{Three kinds of language learning}

%%   Child language acquisition

%%   \pause

%% \vspace{1cm}  Linguists discovering grammars

%%   \pause

%% \vspace{1cm}  Linguistics students doing problem sets

%% \end{frame}

%% \begin{frame}{Pig Latin}
%%   \begin{tabular}{l}
%%     \textipa{pIg}$\to$\textipa{Igpe}\\
%%     \textipa{latIn}$\to$\textipa{atIle}\\
%%     \textipa{\ae sk}$\to$\textipa{\ae ske}\\
%%   \end{tabular}

%%   \pause\vspace{1cm}

%%   \#$\to$[ ]$_i$ / \# [ -vowel ]$_i$ [ ]* \_

%%   \#$\to$e / \_

%% [ -vowel ]$\to\varnothing$ / \# \_ 
%% \end{frame}

\begin{frame}{Motivation}
  \Large \textbf{Problem:} Understand the principles that underlie linguistic generalization.\\\vspace{1cm} \textbf{Approach:} Triangulate these inductive principles with a wide variety of problems/data sets
\end{frame}

\begin{frame}{A (reverse) engineering problem}
  \Huge How is it possible 
  to learn a large number of diverse natural and artificial grammars from 
  relatively small data sets?
\end{frame}

\begin{frame}[t]{Bayesian Program Learning (BPL)}
\begin{textblock*}{5cm}(0cm,1cm)
\visible<2->{      \includegraphics[width=4cm]{BPL1.png}}    \end{textblock*}
\begin{textblock*}{6cm}(0cm,6cm)
\visible<3->{      \includegraphics[width=6cm]{bpl2.png}}    \end{textblock*}
\begin{textblock*}{6cm}(6.5cm,2cm)
\visible<4->{      \includegraphics[width=6cm]{bpl3.png}}    \end{textblock*}
\end{frame}

\begin{frame}{Talk roadmap}

  \begin{itemize}
  \item Bayesian   program learning (BPL) for phonology
  \item Artificial grammar learning: using BPL as a tool for studying simplicity trade-offs
  \item Phonological rules in natural language: BPL as a tool for explaining a breadth of phenomena
  \item Simplicity metrics and universal grammar
  \item A problem with minimum description length as a simplicity metric
    \end{itemize}

  \end{frame}

\begin{frame}{Bayesian Program Learning (BPL) for phonology}
  %\hspace{-2cm}
  %\vspace{-4cm}
  \begin{textblock*}{5cm}(0cm,-3cm)
  \includegraphics[pages={1}, width=12cm]{generativeModel.pdf}    \end{textblock*}
\end{frame}

\begin{frame}{What makes a good model?}
  Probable$\iff$low description length$\iff$simple

  \begin{equation*}
    \text{descriptionLength}(randomEvent) = -\log\mathbb{P} [randomEvent]
  \end{equation*}

    \pause

  \begin{eqnarray*}
    \text{descriptionLength}(\text{model}; \text{data}) &=& \text{descriptionLength}(\text{model})\\& +& \sum_{x\in \text{data}}\text{descriptionLength}(x ; \text{model})
  \end{eqnarray*}

  \pause

  \begin{equation*}
    \text{descriptionLength}(\text{model}) \sim \text{program size}
  \end{equation*}

  \pause

  \begin{gather*}
    \text{descriptionLength}(x ; \text{model})  = \text{size of \emph{x}'s UR in model} \\
    \text{descriptionLength}(\text{[\textipa{pofepo}]}; \text{ABA}) = \text{len(/\textipa{pofe}/)} = 4\\
        \text{ (see Richard Futrell's talk)}
  \end{gather*}

  \pause

  \textbf{Bayesian Program Learning framed as compression}\\
  \textbf{Old idea from Solomonoff}

\end{frame}

\begin{frame}{The program space}
\begin{textblock*}{5cm}(0cm,-3cm)
  \includegraphics[pages={1}, width=12cm]{generativeModel.pdf}    \end{textblock*}
\begin{tikzpicture}[remember picture,overlay]
\draw[green, very thick] (6,2.5) ellipse (3cm and 1.5cm);
\end{tikzpicture}  
  \end{frame}
\begin{frame}{The program space: SPE-style context-sensitive rewrites}
  \begin{tabular*}{\textwidth}{ll}
\toprule
Grammar rule  & English description \\\midrule
$\mathcal{P}\to \mathcal{R}\circ \mathcal{R}\circ \cdots \circ \mathcal{R}$&\textbf{P}honology is compositions of \textbf{r}ewrites\\
$\mathcal{R}\to \mathcal{F}\longrightarrow \mathcal{C} / \mathcal{T}\_\mathcal{T}$& Rewrite \textbf{f}ocus to \textbf{c}hange between \textbf{t}riggers\\
$\mathcal{T}\to \# \mathcal{T}' | \mathcal{T}'$ & \textbf{T}riggers optionally match end of string, \#\\
$\mathcal{T}'\to \epsilon | \mathcal{X} \mathcal{T}' | \mathcal{X}^* \mathcal{T}'$ & \textbf{T}riggers are sequences of matrices $\mathcal{X}$\\
$\mathcal{X}\to a|t|s|\cdots$&Matrices can be constant phonemes\\
$\mathcal{X}\to [\pm \mathcal{E} \; \pm \mathcal{E} \; \cdots \; \pm \mathcal{E}]$& Matrices check features $\mathcal{E}$\\
$\mathcal{E}\to \text{voice}|\text{nasal}|\cdots$&Standard phonological f\textbf{e}atures\\
$\mathcal{F}\to \mathcal{X}$&Focus can be a feature matrix\\
$\mathcal{F}\to\mathbb{Z}$&Focus can be one of the triggers (copies it)\\
$\mathcal{F}\to\varnothing$&Insertion rule\\
$\mathcal{C}\to\mathcal{X}$&Structural change can be a feature matrix\\
$\mathcal{C}\to\varnothing$&Deletion rule\\
$\mathcal{C}\to\mathbb{Z}$&\begin{tabular}{l}
  Structural change constrained to\\ match a triggering feature matrix
  \end{tabular}
\bottomrule
\end{tabular*}

\end{frame}

\begin{frame}{The programs}
\begin{textblock*}{5cm}(0cm,-3cm)
  \includegraphics[pages={1}, width=12cm]{generativeModel.pdf}    \end{textblock*}
\begin{tikzpicture}[remember picture,overlay]
\draw[green, very thick] (3,0.25) ellipse (2cm and 1.5cm);
\end{tikzpicture}  
\end{frame}

\begin{frame}{The programs}
  \begin{itemize}
  \item Morphology: Simple concatenative rules combining
    underlying forms of morphemes based on
    morphological function.
  \end{itemize}
  SING: \# + rog + \#
  \\PLURAL: \# + rog + i + \#

  \begin{itemize}
  \item Phonology: Ordered rules that transform resulting phone sequences.
  \end{itemize}
  \textipa{o}$\to$\textipa{u}$/$ \_ [-nasal +voice]\#\\
  obstruent$\to$ [-voice] $/$\_\#
\end{frame}

\begin{frame}{The lexicon}
\begin{textblock*}{5cm}(0cm,-3cm)
  \includegraphics[pages={1}, width=12cm]{generativeModel.pdf}    \end{textblock*}
\begin{tikzpicture}[remember picture,overlay]
  \draw[green, very thick] (4.5,-1.5) ellipse (0.75cm and 0.75cm);
  \draw[green, very thick] (7.5,-1.5) ellipse (0.75cm and 0.75cm);
  \draw[green, very thick] (10,-1.5) ellipse (0.75cm and 0.75cm);
\end{tikzpicture}  
\end{frame}

\begin{frame}{The lexicon}
  \large

  Inventory of stems (``underlying representations'')

  \vspace{1cm}
  \centering
\begin{columns}[T] % align columns
  \begin{column}{.25\textwidth}
  \textbf{Polish}:    
  \\\begin{tabular}{l}

    rog\\
    klub\\
    dvon\\
    ...
    \end{tabular}

    \end{column}%
%\hfill%
\begin{column}{.25\textwidth}
\textbf{AAB:}
\\\begin{tabular}{l}
pofe\\
wuga\\
...
  \end{tabular}

      \end{column}%
\end{columns}

\vspace{1cm}

Shorter stem$\iff$Better compression$\iff$Higher likelihood of data

\end{frame}

\begin{frame}{}

  \Huge
\centering  But how do you find any good programs in the first place?

  \end{frame}

\begin{frame}{The search problem}
  \textbf{program synthesis techniques from Armando Solar-Lezama}
  \begin{columns}[T] % align columns
    \begin{column}{.6\textwidth}
      \rule{\linewidth}{4pt}
      \emph{SAT/SMT solving:}
\centering  \includegraphics[width = 6cm]{background.png}
    \end{column}%
\hfill%
\begin{column}{.5\textwidth}
  \rule{\linewidth}{4pt}
  \emph{Sketch:}
\includegraphics[width = 5cm]{generatorExample.png}

      \end{column}%
  \end{columns}
  \pause

  \Large
  \textcolor{green}{\cmark} Guarantee: Exact optimization
  \\  \textcolor{red}{\xmark} No guarantee: runtime
  \pause

  2 rules, 1 inflection: $\geq (10^{18}\text{rules})^2\times (10^{8} \text{morphologies}) = 10^{42}\text{models}$

\end{frame}

\begin{frame}{Artificial Grammar Learning}

  Widely studied. \\\\Fundamental hypothesis: AGL engages some shared resources with first language acquisition

  \\\\

  \begin{itemize}
  \item ABA (same/different/same): wofewo, \textipa{pik\ae pi}, gugagu
  \item ABB (different/same/same): wowofe, \textipa{pipik\ae}, guguga
  \item Pig Latin
    \item $\cdots$
    \end{itemize}

\end{frame}

\begin{frame}{Artificial Grammar Learning}

  \begin{centering}
   \includegraphics[width = 11.5cm]{synthetic.png}
  \end{centering}

  Marcus 1999: Babies learn these grammars\\
  Gerken 2006: Babies learn these grammars \emph{from only a few examples}\\

\end{frame}

\begin{frame}{Optimal simplicity front for learning ABB from  1 example}
  \centering  \includegraphics[width = 11cm]{marcusAnimation0.png}   
\end{frame}
\begin{frame}{Optimal simplicity front for learning ABB from  1 example}
  \centering  \includegraphics[width = 11cm]{marcusAnimation1.png}   
\end{frame}
\begin{frame}{Optimal simplicity front for learning ABB from  1 example}
  \centering  \includegraphics[width = 11cm]{marcusAnimation2.png}   
\end{frame}
\begin{frame}{Optimal simplicity front for learning ABB from  1 example}
  \centering  \includegraphics[width = 11cm]{marcusAnimation3.png}    
\end{frame}
\begin{frame}{Optimal simplicity front for learning ABB from  1 example}
  \centering  \includegraphics[width = 11cm]{marcusAnimation4.png}   
\end{frame}

\begin{frame}{Optimal simplicity front for learning ABB: 2 examples}
  \centering  \includegraphics[width = 11cm]{syllableFront2.png}  
\end{frame}

\begin{frame}{Optimal simplicity front for learning ABB: 3 examples}
  \centering  \includegraphics[width = 11cm]{syllableFront3.png}  
\end{frame}

%% \begin{frame}{Learning ABB: 5 examples}
%%   \centering  \includegraphics[width = 11cm]{front5.png}  
%% \end{frame}  safe compile latex

\begin{frame}{Optimal simplicity front for learning ABB without syllables, 3 examples}
    \centering  \includegraphics[width = 11cm]{front3.png}  
\end{frame}

\begin{frame}{Optimal simplicity front for learning ABB without syllables, 6 examples}
    \centering  \includegraphics[width = 11cm]{front6.png}  
  \end{frame}

\begin{frame}{Learning natural language phonology}
  The vision: Induce programs describing diverse phonological systems from many languages. Imagine getting data from field linguists

  \vspace{1cm}
Ours is not the first linguistic rule learner: Ezer Rasin \&  Roni Katzir (this workshop); Albright \& Hayes 2003; Yip \& Sussman 1996; Constantine Lignos \& Charles Yang 2010; Doyle \& Bicknell \& Levy 2014; Colin Wilson 2006

\vspace{1cm}

\large \textbf{What is a good stepping stone toward this vision?}
  \end{frame}

\begin{frame}{Doing your phonology homework}

  \begin{textblock*}{10cm}(0.5cm,1.5cm)\includegraphics[width=5cm]{phonologyCover.png}\end{textblock*}

  \begin{textblock*}{10cm}(2.5cm,1.3cm)\visible<2>{\includegraphics[width=10cm]{phonology1.png}}\end{textblock*}

  %    \begin{textblock*}{10cm}(2.5cm,1.3cm)\visible<3->{\includegraphics[width=10cm]{alternationExample.png}}\end{textblock*}
  \begin{textblock*}{10cm}(0.5cm,1.3cm)\visible<3->{\frame{\includegraphics[width=12cm]{exampleProblem1.png}}}\end{textblock*}
  \begin{textblock*}{10cm}(1cm,2.3cm)\visible<4->{\frame{\includegraphics[width=12cm]{exampleProblem2.png}}}\end{textblock*}
  \begin{textblock*}{10cm}(0.75cm,3.3cm)\visible<5->{\frame{\includegraphics[width=12cm]{exampleProblem4.png}}}\end{textblock*}
\end{frame}

\begin{frame}{What it can do right now}
\tiny  \begin{itemize}
\item \textbf{ Kikurai}: fricatives alternate with stops after nasals
\item \textbf{ Farsi}: Trills alternate with flaps
\item \textbf{ Amharic}: alternation between \textipa{@} \& \textipa{E}
\item \textbf{ Gen}: l/r alternation
\item \textbf{ Greek}: Velar stops alternate with palletized versions before front vowels
  \item \textbf{ Osage}: coronal stops become dentals before central vowels
\item \textbf{ Kishambaa}: voiced/unvoiced nasals alternate
\item \textbf{  Thai}: stops are unreleased word finally
\item \textbf{ Palauan}: a word initial neutralizing rule
\item \textbf{ Quechua}: Velar becomes uvular when followed by uvular (spreading)
\item \textbf{ Lhasa Tibetan}: no contrast between velar/uvular, or voiced/voiceless stops or fricatives
\item \textbf{ Axininca Campa}: stops become glides
\item \textbf{ Kikuyu}: infinitive prefix can surface as either k or \textipa{G}
\item \textbf{ Korean}: vowel harmony, aspiration only surfaces in certain contexts
\item \textbf{ Hungarian}: vowel harmony and voicing assimilation
\item \textbf{ Kikuria}: vowel harmony
\item \textbf{ Farsi}: a deletion rule explains singular/plural
\item \textbf{ Tibetan}: initial consonant cluster reduction explains counting system
\item \textbf{ Makonde}: stress patterns; unstressed vowels are neutralized
\item \textbf{ North Saami}: 3 neutralization rules explain nominative sg essive
\item \textbf{ Samoan}: 2 deletion rules that explain words that sound the same in one inflection being different in another
\item \textbf{     Russian}: devoicing of word final obstruent
\item \textbf{ English}: verbal inflections (voicing assimilation, epenthesis)
\item \textbf{ Finnish}: nominative/partive explained by vowel raising and vowel harmony
\item \textbf{ Kerewe}: Interacting tone rules
\item \textbf{ Polish}: vowel alternations interacting with devoicing of word final obstruent in singular/plural
\item \textbf{ Ancient Greek}: voicing assimilation, deaspiration, deletion rule. Order matters.
\item \textbf{ Serbo-Croatian}: predictable stress, devoicing, neutralizing, epenthesis
\item \textbf{Artificial grammar learning}: Pig Latin, ABB, ABA, AxA, ABx, AAx, Chinese duplication
\end{itemize}

  \begin{textblock*}{5cm}(8.1cm,1cm)
    \Huge
\visible<2->{    Handles 60\% of the textbook}
  \end{textblock*}

  \end{frame}

\begin{frame}[t]{Polish: Final devoicing + Rule ordering}

  \begin{center}
    \begin{tabular}{ll}
      singular&plural\\\hline
      \textipa{dom} & \textipa{domi}\\
      \textipa{kot} & \textipa{koti}\\
      \textipa{lut} & \textipa{lodi}\\
      \textipa{vus} & \textipa{vozi}\\
      \textipa{wuk} & \textipa{wugi}\\
      \textipa{ruk} & \textipa{rogi}\\
      \textipa{bur} & \textipa{bori}\\
      \textipa{\|v{s}um} & \textipa{\|v{s}umi}
    \end{tabular}
  \end{center}

  \pause

\vspace{0.5cm}

  \textipa{o}$\to$\textipa{u}$/$ \_ [-nasal +voice]\#

  [-sonorant]$\to$[-voice] $/$ \_ \#

  \hline

  singular: stem

  plural: stem$ + /\text{i}/$

\end{frame}

%% \begin{frame}[t]{Kikuria: vowel harmony}

%%   \begin{center}
%%     \begin{tabular}{ll}
%%      verb&verb for\\\hline
%% \textipa{rema} & \textipa{remera}\\
%% \textipa{roma} & \textipa{romera}\\
%% \textipa{tiga} & \textipa{tegera}\\
%% \textipa{ruga} & \textipa{rogera}\\
%% \textipa{hoora} & \textipa{hoorera}\\
%% \textipa{siika} & \textipa{seekera}\\
%% \textipa{huuta} & \textipa{hootera}\\
%% \textipa{suraaNga} & \textipa{suraaNgera}
%% \end{tabular}
%%   \end{center}

%%     \pause

%% \vspace{0.5cm}

%% [+high]$\to$[+middle]$ /$ \_ [-low]* [+middle]

%% \hline
%% verb: stem$ + /\text{a}/$

%% verb for: stem$ + /\text{era}/$

%% \end{frame}

%% \begin{frame}[t]{Makonde: stress patterns}

%%   \begin{center}
%%     \begin{tabular}{lll}
%%       repeated imperative&past&imperative\\\hline
%% \textipa{am\'aNga} & \textipa{am\'ile} & \textipa{\'ama}\\
%% %\textipa{ak\'aNga} & \textipa{ak\'ile} & \textipa{\'aka}\\
%% %\textipa{av\'aNga} & \textipa{av\'ile} & \textipa{\'ova}\\
%% \textipa{am\'aNga} & \textipa{am\'ile} & \textipa{\'oma}\\
%% \textipa{ut\'aNga} & \textipa{ut\'ile} & \textipa{\'uta}\\
%% \textipa{av\'aNga} & \textipa{av\'ile} & \textipa{\'eva}\\
%% \textipa{tav\'aNga} & \textipa{tav\'ile} & \textipa{t\'ava}\\
%% \textipa{uNg\'aNga} & \textipa{uNg\'ile} & \textipa{\'uNga}\\
%% \textipa{pat\'aNga} & \textipa{pat\'ile} & \textipa{p\'ota}
%% \end{tabular}
%%   \end{center}

%%     \pause

%% \vspace{0.5cm}

%%   [ ]$\to$[-stress] $/$ \_ [ ]* [ +stress] \\
%%   [+middle -stress]$\to$a $/$ \_ [ ]\\\hline
%% repeated imperative:   stem$ + $\textipa{\'aNga}\\
%% past:  stem$ + $\textipa{\'ile}\\
%% imperative:  stem$ + $\textipa{a}

%% \end{frame}

\begin{frame}{Scaling program synthesis to learn large grammars}

  \begin{textblock*}{10cm}(0.5cm,2cm)\includegraphics[width=6cm]{Croatian1.png}\end{textblock*}
  \begin{textblock*}{10cm}(6.5cm,2cm)\includegraphics[width=6cm]{Croatian2.png}\end{textblock*}

\end{frame}

%% \begin{frame}{Relaxing the requirement of complete search}

%%   \begin{tikzpicture}
%%     \draw[thick]     (0.1,-0.1) rectangle (4.5,6);
%%     \node at (7.5,3) {$X = \text{observed data}$ };
%%     \node at (2.5,6.25) {Space of all grammars};

%%     \node(empty)[align = center] at (0.5,-0.5) {\tiny empty grammar};
%%     \node(emptyp)[fill = cyan,align = center] at (0.5,0.5) {\tiny $\mathcal{G}_0$};    %    \draw[blue,fill = blue] (0.5,0.5) circle (0.1) ;
%%     \draw[thick,->] (empty.north) -- (emptyp.south);

%%     \node(target)[align = center] at (7,5) {\tiny target grammar: best explanation for $X$};
%%     \node(targetGrammar)[green, draw, fill, star, star points=5] at (3.7,5){};%    \draw[yellow,fill = yellow] (2.5,5.5) star (star points = 5);% (0.1) ;
%%     \draw[thick,->] (target.west) -- (targetGrammar.east);

%% \pause
%% \draw[red] (0.5,0.5) circle (2) ;

%% \node at (7.5,2) {\Small $c_1\in X\text{ is a counterexample to }\mathcal{G}_0$};
%% %    };

%%     \pause

%%     \node(g1p)[fill = cyan,align = center] at (1.8,1.8) {\tiny $\mathcal{G}_1$};
%%     \node(g1)[align = center] at (3,-0.5) {\tiny explains $c_1$};  
%%     \draw[thick,->] (g1.north) -- (g1p.south);
%%     \draw[dashed] (emptyp.north) -- (g1p.west);

%%     \pause
%% \draw[red] (g1p.north) circle (2) ;
%% \node at (7.5,1) {\Small $c_2\in X\text{ is a counterexample to }\mathcal{G}_1$};
%% %    };

%%     \pause

%%     \node(g2p)[fill = cyan,align = center] at (3,3.8) {\tiny $\mathcal{G}_2$};
%%     \node(g2)[align = center] at (-0.5,3.9) {\tiny explains $c_2$};  
%%     \draw[thick,->] (g2.east) -- (g2p.west);
%%     \draw[dashed] (g1p.north) -- (g2p.south);

%%     \pause
%%     \draw[dashed] (g2p.north) -- (targetGrammar.south);
%%     \node at (7.5,0) {\Small $c_3\in X\text{ is a counterexample to }\mathcal{G}_2$};
%% \draw[red] (g2p.north) circle (2) ;
%%   \end{tikzpicture}

%% \end{frame}

\begin{frame}{Scaling program synthesis to learn large grammars}

  \begin{textblock*}{10cm}(0.5cm,2cm)\includegraphics[width=6cm]{Croatian1.png}\end{textblock*}
  \begin{textblock*}{10cm}(6.5cm,2cm)\includegraphics[width=6cm]{Croatian2.png}\end{textblock*}

  \begin{textblock*}{10cm}(2cm,3cm)
    \begin{tikzpicture}
      \node[red,fill = white,align = left]{\Large
        [ +vowel ]$\to$[ +stress ] / [ +stress ] [  ]* \_\\\Large
        $\varnothing\to$ a / [ -vowel ] \_ [ -vowel ] \#\\\Large
        [ +fricative ]$\to$[ -voice ] /  \_ k\\\Large
        [ +vowel ]$\to$[ -stress ] /  \_ [  ]* [ +stress ]\\\Large
        [ +lateral ]$\to$o /  \_ \# \\
        stem \\
       stem + / a /\\
       stem + / o /\\
       stem + / i /
};
      \end{tikzpicture}
\end{textblock*}
\end{frame}

\begin{frame}{}

  \Huge

\Huge  Investigating the inductive bias over grammars

  \end{frame}

\begin{frame}[t]{Motivation: Counting in Tibetan}

    \begin{textblock*}{10cm}(0.5cm,2.5cm)\includegraphics[width=12cm]{Tibetan.png}\end{textblock*}

\end{frame}

\begin{frame}[t]{Motivation: Counting in Tibetan}

  \begin{center}
    \begin{tabular}{l|l}
      Number&Pronunciation\\\hline
      1& \textipa{\v{j}ig}\\
      4& \textipa{\v{s}i}\\
      10&\textipa{\v{j}u}\\
      11&\textipa{\v{j}u+g\v{j}ig} (10+1)\\
      40&\textipa{\v{s}i+b\v{j}u} (4+10)
  \end{tabular}
    \end{center}

%%   \vspace{1cm}

%%   Explanation:

%%   \textipa{\v{j}ig} (``1'') is really     \textipa{g\v{j}ig}

%%   \textipa{\v{j}u} (``10'') is really     \textipa{b\v{j}u}

%%   \begin{textblock*}{10cm}(8.3cm,6.5cm)\includegraphics[width=3cm]{Tim.jpeg}\end{textblock*}

%% \textbf{Program:}

%% C$\to \varnothing $/\# C \_

%% (Delete initial cluster of consonants)
\end{frame}

\begin{frame}[t]{Motivation: Counting in Tibetan}

  \begin{center}
    \begin{tabular}{l|l}
      Number&Pronunciation\\\hline
      1& \textipa{\v{j}ig}\\
      4& \textipa{\v{s}i}\\
      10&\textipa{\v{j}u}\\
      11&\textipa{\v{j}u+g\v{j}ig} (10+1)\\
      40&\textipa{\v{s}i+b\v{j}u} (4+10)
  \end{tabular}
    \end{center}

  \vspace{1cm}

  Explanation:

  \textipa{\v{j}ig} (``1'') is really     \textipa{g\v{j}ig}

  \textipa{\v{j}u} (``10'') is really     \textipa{b\v{j}u}

%  \begin{textblock*}{10cm}(8.3cm,6.5cm)\includegraphics[width=3cm]{Tim.jpeg}\end{textblock*}

\textbf{Program:}

C$\to \varnothing $/\# \_ C

(Delete initial cluster of consonants)
\end{frame}

\begin{frame}[t]{Counting in Tibetan}

  \begin{center}
    \begin{tabular}{l|l}
      Number&Pronunciation\\\hline
      1& \textipa{\v{j}ig}\\
      4& \textipa{\v{s}i}\\
      10&\textipa{\v{j}u}\\
      11&\textipa{\v{j}u+g\v{j}ig} (10+1)\\
      40&\textipa{\v{s}i+b\v{j}u} (4+10)
  \end{tabular}
    \end{center}

  \vspace{1cm}

  Explanation:

  \textipa{\v{j}ig} (``1'') is really     \textipa{g\v{j}ig}

  \textipa{\v{j}u} (``10'') is really     \textipa{b\v{j}u}

\textbf{Program:}

[-nasal]$\to \varnothing $/\# \_

\textbf{Delete initial nonnasal???}
\end{frame}

\begin{frame}{}
  \large

%  \centering

  \textbf{Problem:} \\Probable grammar $\not=$ grammars children/linguists prefer

  \pause

  \\\\\vspace{1cm}

  \textbf{Solution:}\\ Get a better inductive bias (a better universal grammar)

  \end{frame}

%% \begin{frame}{Learning which features are important}

%% \centering  \includegraphics[width = 11.5cm]{featureFrequencies.png}

%% \end{frame}

\begin{frame}{Learning what rules look like}

\centering  \includegraphics[width = 11.5cm]{skeletonFrequencies.png}

\end{frame}

\begin{frame}{Learning a (fragment) grammar over grammars}
  Fragment grammars: O'Donnell 2011
  \small
\begin{tabular}{ll}
    FEATUREMATRIX $\to$& [ -voice ]\\
     FEATUREMATRIX $\to$& [ -sonorant ]\\
 FEATUREMATRIX $\to$& [ +voice ]\\
 FEATUREMATRIX $\to$& [ +tense ]\\
 FEATUREMATRIX $\to$& [ +highTone ]\\
 FEATUREMATRIX $\to$& [ +middle ]\\
 TRIGGER $\to$& [ +sibilant ]\\
 TRIGGER $\to$& [ ] [ +highTone ]\\
 TRIGGER $\to$& [ +stop +voice ]\\
 TRIGGER $\to$& [ -low ]\\
 TRIGGER $\to$& [ ]* FEATUREMATRIX\\
 RULE $\to$& [ -vowel ] $\to$ $\varnothing$ / TRIGGER \_ TRIGGER\\
 RULE $\to$& [ +low ] $\to$ CONSTANT / [ ]* FEATUREMATRIX \_\\
 RULE $\to$& [ ] $\to$ [ +voice ] / \_ TRIGGER\\
 RULE $\to$& CONSTANT $\to$ CONSTANT / [ ] \_ TRIGGER\\
 RULE $\to$& [ ] $\to$ [ -highTone ] / TRIGGER \_ TRIGGER\\
 RULE $\to$& [ -sonorant ] $\to$ [ -voice ] / \_ \#\\
 RULE $\to$& [ ] $\to$ FC / TRIGGER \_ [ ] FEATUREMATRIX\\
    \end{tabular}
\end{frame}
\begin{frame}{Learning a (fragment) grammar over grammars}
  Fragment grammars: O'Donnell 2011
  \small
  \begin{tabular}{ll}
    FEATUREMATRIX $\to$& [ -voice ]\\
     FEATUREMATRIX $\to$& [ -sonorant ]\\
 FEATUREMATRIX $\to$& [ +voice ]\\
 FEATUREMATRIX $\to$& [ +tense ]\\
 FEATUREMATRIX $\to$& [ +highTone ]\\
 FEATUREMATRIX $\to$& [ +middle ]\\
 TRIGGER $\to$& [ +sibilant ]\\
 TRIGGER $\to$& [ ] [ +highTone ]\\
 TRIGGER $\to$& [ +stop +voice ]\\
 TRIGGER $\to$& [ -low ]\\
 TRIGGER $\to$& [ ]* FEATUREMATRIX\\
 RULE $\to$& {\color{red}[ -vowel ] $\to$ $\varnothing$ / TRIGGER \_ TRIGGER}\\
 RULE $\to$& [ +low ] $\to$ CONSTANT / [ ]* FEATUREMATRIX \_\\
 RULE $\to$& [ ] $\to$ [ +voice ] / \_ TRIGGER\\
 RULE $\to$& CONSTANT $\to$ CONSTANT / [ ] \_ TRIGGER\\
 RULE $\to$& [ ] $\to$ [ -highTone ] / TRIGGER \_ TRIGGER\\
 RULE $\to$& {\color{red}[ -sonorant ] $\to$ [ -voice ] / \_ \#}\\
 RULE $\to$& [ ] $\to$ FC / TRIGGER \_ [ ] FEATUREMATRIX\\
  \end{tabular}

\end{frame}

\begin{frame}{Learning a (fragment) grammar over grammars}
  Fragment grammars: O'Donnell 2011
  \small
  \begin{tabular}{ll}
    FEATUREMATRIX $\to$& [ -voice ]\\
     FEATUREMATRIX $\to$& [ -sonorant ]\\
 FEATUREMATRIX $\to$& [ +voice ]\\
 FEATUREMATRIX $\to$& [ +tense ]\\
 FEATUREMATRIX $\to$& [ +highTone ]\\
 FEATUREMATRIX $\to$& [ +middle ]\\
 TRIGGER $\to$& [ +sibilant ]\\
 TRIGGER $\to$& [ ] [ +highTone ]\\
 TRIGGER $\to$& [ +stop +voice ]\\
 TRIGGER $\to$& [ -low ]\\
 TRIGGER $\to$& [ ]* FEATUREMATRIX\\
 RULE $\to$& {\color{red}[ -vowel ] $\to$ $\varnothing$ / TRIGGER \_ TRIGGER}\\
 RULE $\to$& [ +low ] $\to$ CONSTANT / [ ]* FEATUREMATRIX \_\\
 RULE $\to$& [ ] $\to$ [ +voice ] / \_ TRIGGER\\
 RULE $\to$& CONSTANT $\to$ CONSTANT / [ ] \_ TRIGGER\\
 RULE $\to$& [ ] $\to$ [ -highTone ] / TRIGGER \_ TRIGGER\\
 RULE $\to$& {\color{red}[ -sonorant ] $\to$ [ -voice ] / \_ \#}\\
 RULE $\to$& [ ] $\to$ FC / TRIGGER \_ [ ] FEATUREMATRIX\\
  \end{tabular}

    \begin{textblock*}{6cm}(3cm,2.5cm)
         \textblockcolor{lightgray}
         \hspace{0.7cm}  \large    In light of this UG, the \\
         \hspace{0.7cm}system revises its rule for \\
         \hspace{0.7cm}Tibetan to match the good \\
         \hspace{0.7cm}linguistics students judgment:\\
{\color{red}\hspace{0.3}[ -vowel ] $\to$ $\varnothing$ / } \# \_ [-vowel]
    \end{textblock*}
\end{frame}

\begin{frame}{}

  \Huge
  Problems with minimum description length

  \end{frame}

\begin{frame}{Problems with compression}
  \begin{tabular}{ll}
\textipa{varit} & \textipa{varihin}\\
\textipa{oahpis} & \textipa{oahpisin}\\
\textipa{bissobeahtset} & \textipa{bissobeahtsehin}\\
\textipa{yaaPmin} & \textipa{yaaPmimin}\\
\textipa{gahpir} & \textipa{gahpirin}\\
\textipa{gaauhtsis} & \textipa{gaauhtsisin}\\
\textipa{be\v{s}tor} & \textipa{be\v{s}torin}\\
\textipa{heevemeahhtun} & \textipa{heevemeahhtunin}\\
\textipa{bissomeahtun} & \textipa{bissomeahtumin}\\
\textipa{laDas} & \textipa{laDasin}\\
\textipa{heaNkkan} & \textipa{heaNkkanin}\\
\textipa{yaman} & \textipa{yamanin}
\end{tabular}

\end{frame}

\begin{frame}{Problems with compression}
  \begin{tabular}{ll}
\textipa{varit} & \textipa{varihin}\\
\textipa{oahpis} & \textipa{oahpisin}\\
\textipa{bissobeahtset} & \textipa{bissobeahtsehin}\\
{\color{red}\textipa{yaaPmin}} & {\color{red}\textipa{yaaPmimin}}\\
\textipa{gahpir} & \textipa{gahpirin}\\
\textipa{gaauhtsis} & \textipa{gaauhtsisin}\\
\textipa{be\v{s}tor} & \textipa{be\v{s}torin}\\
{\color{red}\textipa{heevemeahhtun}} &{\color{red} \textipa{heevemeahhtunin}}\\
{\color{red}\textipa{bissomeahtun}} &{\color{red} \textipa{bissomeahtumin}}\\
\textipa{laDas} & \textipa{laDasin}\\
{\color{red}\textipa{heaNkkan}} &{\color{red} \textipa{heaNkkanin}}\\
{\color{red}\textipa{yaman}} &{\color{red} \textipa{yamanin}}
  \end{tabular}

  \visible<2->{
    \begin{textblock*}{5cm}(7cm,1.5cm)
               \textblockcolor{white}
      \Large
      \begin{tabular}{l}
        \texttt{stem}$ + $/in/\\
        m$\to$n word finally\\\\
        \textipa{yaaPmim}$\to$\textipa{yaaPmin}\\
        \textipa{yaaPmim}$+$\textipa{in}$\to$\textipa{yaaPmimin}
        \end{tabular}
  \end{textblock*}}

\end{frame}

\begin{frame}{Problems with compression}
  \begin{tabular}{ll}
\textipa{varit} & \textipa{varihin}\\
\textipa{oahpis} & \textipa{oahpisin}\\
\textipa{bissobeahtset} & \textipa{bissobeahtsehin}\\
{\color{red}\textipa{yaaPmin}} & {\color{red}\textipa{yaaPmimin}}\\
\textipa{gahpir} & \textipa{gahpirin}\\
\textipa{gaauhtsis} & \textipa{gaauhtsisin}\\
\textipa{be\v{s}tor} & \textipa{be\v{s}torin}\\
{\color{red}\textipa{heevemeahhtun}} &{\color{red} \textipa{heevemeahhtunin}}\\
{\color{red}\textipa{bissomeahtun}} &{\color{red} \textipa{bissomeahtumin}}\\
\textipa{laDas} & \textipa{laDasin}\\
{\color{red}\textipa{heaNkkan}} &{\color{red} \textipa{heaNkkanin}}\\
{\color{red}\textipa{yaman}} &{\color{red} \textipa{yamanin}}
  \end{tabular}

  \begin{textblock*}{5cm}(7cm,1.5cm)\textblockcolor{white}
      \Large
      \begin{tabular}{l}
        \texttt{stem}$ + $/i{\color{red}m}/\\
        m$\to$n word finally\\\\
        \textipa{yaaPmim}$\to$\textipa{yaaPmin}\\
        \textipa{yaaPmim}$+$i{\color{red}m}$\to$\\\textipa{yaaPmimim}$\to$\\\textipa{yaaPmimin}
      \end{tabular}
      \visible<2->{\Huge MDL unchanged!}

  \end{textblock*}
      \visible<3->{\vspace{1cm}\color{red} $    \text{cost}(\text{model}; \text{data}) = \text{cost}(\text{model}) + \sum_{x\in \text{data}}\text{cost}(x ; \text{model}) + \text{runtimeCost}(x ; \text{model})$}
\end{frame}

\begin{frame}{Contributions}

  Engineered new approaches to grammar learning drawn from program synthesis. Allows exact study of simplicity trade-offs between different grammars.

  \vspace{1cm}

  A system that can learn many morphophonologies, both natural and artificial, from relatively small amounts of data

  \vspace{1cm}

  Learning to learn phonological rules as a way of estimating and probing universal grammar

  \pause

  \vspace{1cm}

  \textbf{The end.}

\end{frame}

\end{document}

### 2018-02-20 test.tex
initial commit

\documentclass{article}
\usepackage[margin = 0.1cm]{geometry}
\usepackage{tipa}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{longtable}
\begin{document}

\emph{Kikurai:}\\\begin{longtable}{ll}\toprule
Surface form & UR
\\ \midrule
\textipa{aBa\'{a}nto}&\textipa{aBa\'{a}nto}\\
\textipa{aBam\'{u}ra}&\textipa{aBam\'{u}ra}\\
\textipa{amah\'{i}\'{i}ndi}&\textipa{amah\'{i}\'{i}nri}\\
\textipa{amak\'{E}\'{E}ndO}&\textipa{amak\'{E}\'{E}nrO}\\
\textipa{eB\|x{a}}&\textipa{eB\|x{a}}\\
\textipa{eeNgw\'{e}}&\textipa{eeNGw\'{e}}\\
\textipa{eG\|x{a}}&\textipa{eG\|x{a}}\\
\textipa{ekeBw\'{E}}&\textipa{ekeBw\'{E}}\\
\textipa{hoor\'{a}}&\textipa{hoor\'{a}}\\
\textipa{iBiG\'{u}r\'{u}Be}&\textipa{iBiG\'{u}r\'{u}Be}\\
\textipa{iBir\'{u}\'{u}Ng\'{u}uri}&\textipa{iBir\'{u}\'{u}NG\'{u}uri}\\
\textipa{uGus\'{i}ri}&\textipa{uGus\'{i}ri}\\
\textipa{B\'{a}inu}&\textipa{B\'{a}inu}\\
\textipa{Borj\'{o}}&\textipa{Borj\'{o}}\\
\textipa{it\super SiiNg\'{E}na}&\textipa{it\super SiiNG\'{E}na}\\
\textipa{it\super SiiNg\'{u}r\'{u}Be}&\textipa{it\super SiiNG\'{u}r\'{u}Be}\\
\textipa{GaB\|x{a}}&\textipa{GaB\|x{a}}\\
\textipa{it\super SiiNg\'{u}ta}&\textipa{it\super SiiNG\'{u}ta}\\
\textipa{BErEk\'{a}}&\textipa{BErEk\'{a}}\\
\textipa{iGit\'{u}\'{u}mbe}&\textipa{iGit\'{u}\'{u}mBe}\\
\textipa{G\'{u}\'{u}k\'{a}}&\textipa{G\'{u}\'{u}k\'{a}}\\
\textipa{rem\|x{a}}&\textipa{rem\|x{a}}\\
\textipa{rEEnt\'{a}}&\textipa{rEEnt\'{a}}\\
\textipa{oBoG\'{a}\'{a}k\'{a}}&\textipa{oBoG\'{a}\'{a}k\'{a}}\\
\textipa{oBot\'{e}\'{e}nd\'{e}\'{e}ru}&\textipa{oBot\'{e}\'{e}nr\'{e}\'{e}ru}\\
\textipa{okoG\'{e}\'{e}mb\'{a}}&\textipa{okoG\'{e}\'{e}mB\'{a}}\\
\textipa{oko\'{o}mb\'{a}ra}&\textipa{oko\'{o}mB\'{a}ra}\\
\textipa{okoB\'{a}ra}&\textipa{okoB\'{a}ra}\\
\textipa{oko\'{o}nd\'{O}Ga}&\textipa{oko\'{o}nr\'{O}Ga}\\
\textipa{okor\'{O}Ga}&\textipa{okor\'{O}Ga}\\
\textipa{rom\|x{a}}&\textipa{rom\|x{a}}\\
\textipa{teGet\'{a}}&\textipa{teGet\'{a}}\\
\textipa{uku\'{u}mbuurj\'{a}}&\textipa{uku\'{u}mBuurj\'{a}}\\
\textipa{uruG\'{u}ta}&\textipa{uruG\'{u}ta}\\
\bottomrule\end{longtable}

\begin{longtable}{ll}\toprule
\emph{The surface form...}&\emph{Is underlyingly...}
\\ \midrule
\textipa{b}&\textipa{B}\\
\textipa{d}&\textipa{r}\\
\textipa{g}&\textipa{G}\\
\bottomrule\end{longtable}

\begin{tabular}{l}\emph{Rules: }\\
\verb|[+liquid]| $\to$ \textipa{d} / \verb|[-vowel]| \verb|_| \\\verb|[-sonorant]| $\to$ \verb|[-continuant]| / \verb|[-vowel]| \verb|_| 
\end{tabular}

\pagebreak

\emph{Modern Greek:}\\\begin{longtable}{ll}\toprule
Surface form & UR
\\ \midrule
\textipa{kano}&\textipa{kano}\\
\textipa{kori}&\textipa{kori}\\
\textipa{xano}&\textipa{xano}\\
\textipa{xori}&\textipa{xori}\\
\textipa{x\super jino}&\textipa{xino}\\
\textipa{k\super jino}&\textipa{kino}\\
\textipa{krima}&\textipa{krima}\\
\textipa{xrima}&\textipa{xrima}\\
\textipa{xufta}&\textipa{xufta}\\
\textipa{kufeta}&\textipa{kufeta}\\
\textipa{kali}&\textipa{kali}\\
\textipa{xali}&\textipa{xali}\\
\textipa{x\super jeli}&\textipa{xeli}\\
\textipa{k\super jeri}&\textipa{keri}\\
\textipa{x\super jeri}&\textipa{xeri}\\
\textipa{ox\super ji}&\textipa{oxi}\\
\bottomrule\end{longtable}

\begin{longtable}{ll}\toprule
\emph{The surface form...}&\emph{Is underlyingly...}
\\ \midrule
\textipa{x\super j}&\textipa{x}\\
\textipa{k\super j}&\textipa{k}\\
\bottomrule\end{longtable}

\begin{tabular}{l}\emph{Rules: }\\
\verb|[+back -voice]| $\to$ \verb|[+palletized]| /  \verb|_| \verb|[-liquid -back]|
\end{tabular}

\pagebreak

\emph{Farsi:}\\\begin{longtable}{ll}\toprule
Surface form & UR
\\ \midrule
\textipa{\ae \~rteS}&\textipa{\ae \~rteS}\\
\textipa{fa\~rsi}&\textipa{fa\~rsi}\\
\textipa{q\ae d\~ri}&\textipa{q\ae d\~ri}\\
\textipa{\~rah}&\textipa{\~rah}\\
\textipa{\~rast}&\textipa{\~rast}\\
\textipa{\~riS}&\textipa{\~riS}\\
\textipa{aha\r*{\~r}}&\textipa{aha\r*{\~r}}\\
\textipa{ax\ae \r*{\~r}}&\textipa{ax\ae \r*{\~r}}\\
\textipa{h\ae \~rtow\r*{\~r}}&\textipa{h\ae \~rtow\r*{\~r}}\\
\textipa{Si\r*{\~r}}&\textipa{Si\r*{\~r}}\\
\textipa{aha\v{r}i}&\textipa{aha\~ri}\\
\textipa{b\ae \v{r}ad\ae \r*{\~r}}&\textipa{b\ae \~rad\ae \r*{\~r}}\\
\textipa{t\super Se\v{r}a}&\textipa{t\super Se\~ra}\\
\textipa{da\v{r}id}&\textipa{da\~rid}\\
\textipa{bi\v{r}\ae ng}&\textipa{bi\~r\ae ng}\\
\textipa{Si\v{r}ini}&\textipa{Si\~rini}\\
\bottomrule\end{longtable}

\begin{longtable}{ll}\toprule
\emph{The surface form...}&\emph{Is underlyingly...}
\\ \midrule
\textipa{\v{r}}&\textipa{\~r}\\
\bottomrule\end{longtable}

\begin{tabular}{l}\emph{Rules: }\\
\verb|[+trill]| $\to$ \textipa{\v{r}} / \verb|[+vowel]| \verb|_| \verb|[+voice]|
\end{tabular}

\pagebreak

\emph{Osage:}\\\begin{longtable}{ll}\toprule
Surface form & UR
\\ \midrule
\textipa{d\'{a}br\~i}&\textipa{d\'{a}br\~i}\\
\textipa{\'{a}Dik\super h\~aZ\~a}&\textipa{\'{a}dik\super h\~aZ\~a}\\
\textipa{dat\super Sp\'{e}}&\textipa{dat\super Sp\'{e}}\\
\textipa{t\super SP\'{e}De}&\textipa{t\super SP\'{e}de}\\
\textipa{dakP\'{e}}&\textipa{dakP\'{e}}\\
\textipa{D\'{e}ze}&\textipa{d\'{e}ze}\\
\textipa{d\'{a}l\~i}&\textipa{d\'{a}l\~i}\\
\textipa{D\'{i}e}&\textipa{d\'{i}e}\\
\textipa{daSt\'{u}}&\textipa{daSt\'{u}}\\
\textipa{D\'{i}Ski}&\textipa{d\'{i}Ski}\\
\bottomrule\end{longtable}

\begin{longtable}{ll}\toprule
\emph{The surface form...}&\emph{Is underlyingly...}
\\ \midrule
\textipa{D}&\textipa{d}\\
\bottomrule\end{longtable}

\begin{tabular}{l}\emph{Rules: }\\
\textipa{d} $\to$ \textipa{D} /  \verb|_| \verb|[-low]|
\end{tabular}

\pagebreak

\emph{Amharic:}\\\begin{longtable}{ll}\toprule
Surface form & UR
\\ \midrule
\textipa{f@r@s}&\textipa{f@r@s}\\
\textipa{t@n@sa}&\textipa{t@n@sa}\\
\textipa{jEl1Sl1S}&\textipa{j@l1Sl1S}\\
\textipa{majEt}&\textipa{maj@t}\\
\textipa{g@nz@b}&\textipa{g@nz@b}\\
\textipa{SEgna}&\textipa{S@gna}\\
\textipa{n@\~n}&\textipa{n@\~n}\\
\textipa{m@wd@d}&\textipa{m@wd@d}\\
\textipa{m@nn@sat}&\textipa{m@nn@sat}\\
\textipa{m@mk@r}&\textipa{m@mk@r}\\
\textipa{ZEle}&\textipa{Z@le}\\
\textipa{jEll@m}&\textipa{j@ll@m}\\
\textipa{m@t\super S}&\textipa{m@t\super S}\\
\textipa{m@st@t}&\textipa{m@st@t}\\
\textipa{f@ll@g@}&\textipa{f@ll@g@}\\
\textipa{ag@\~n\~nE}&\textipa{ag@\~n\~n@}\\
\textipa{t@m@t\super St\super SE}&\textipa{t@m@t\super St\super S@}\\
\textipa{mokk@r@}&\textipa{mokk@r@}\\
\textipa{kaZZE}&\textipa{kaZZ@}\\
\textipa{ZEmm@r@}&\textipa{Z@mm@r@}\\
\textipa{lat\super St\super SE}&\textipa{lat\super St\super S@}\\
\textipa{aSSE}&\textipa{aSS@}\\
\textipa{b@kk@l@}&\textipa{b@kk@l@}\\
\textipa{SEm@gg@l@}&\textipa{S@m@gg@l@}\\
\bottomrule\end{longtable}

\begin{longtable}{ll}\toprule
\emph{The surface form...}&\emph{Is underlyingly...}
\\ \midrule
\textipa{E}&\textipa{@}\\
\bottomrule\end{longtable}

\begin{tabular}{l}\emph{Rules: }\\
\textipa{@} $\to$ \textipa{E} / \verb|[-anterior -liquid -high]| \verb|_| 
\end{tabular}

\pagebreak

\emph{Gen:}\\\begin{longtable}{ll}\toprule
Surface form & UR
\\ \midrule
\textipa{agble}&\textipa{agble}\\
\textipa{agoNglo}&\textipa{agoNglo}\\
\textipa{aNOli}&\textipa{aNOli}\\
\textipa{akplO}&\textipa{akplO}\\
\textipa{sabulE}&\textipa{sabulE}\\
\textipa{sra}&\textipa{sla}\\
\textipa{alO}&\textipa{alO}\\
\textipa{atitrwE}&\textipa{atitlwE}\\
\textipa{avlO}&\textipa{avlO}\\
\textipa{blafogbe}&\textipa{blafogbe}\\
\textipa{drE}&\textipa{dlE}\\
\textipa{edrO}&\textipa{edlO}\\
\textipa{exlO}&\textipa{exlO}\\
\textipa{exle}&\textipa{exle}\\
\textipa{hlE}&\textipa{hlE}\\
\textipa{NlO}&\textipa{NlO}\\
\textipa{t\super Sr\~O}&\textipa{t\super Sl\~O}\\
\textipa{\~nr\~a}&\textipa{\~nl\~a}\\
\textipa{klO}&\textipa{klO}\\
\textipa{tre}&\textipa{tle}\\
\textipa{vlu}&\textipa{vlu}\\
\textipa{lO}&\textipa{lO}\\
\textipa{mla}&\textipa{mla}\\
\textipa{pleplelu}&\textipa{pleplelu}\\
\textipa{wla}&\textipa{wla}\\
\textipa{zro}&\textipa{zlo}\\
\textipa{esrO}&\textipa{eslO}\\
\textipa{etro}&\textipa{etlo}\\
\textipa{e\~nr\~O}&\textipa{e\~nl\~O}\\
\textipa{d\super Zro}&\textipa{d\super Zlo}\\
\bottomrule\end{longtable}

\begin{longtable}{ll}\toprule
\emph{The surface form...}&\emph{Is underlyingly...}
\\ \midrule
\textipa{r}&\textipa{l}\\
\bottomrule\end{longtable}

\begin{tabular}{l}\emph{Rules: }\\
\verb|[ ]| $\to$ \verb|[-lateral]| / \verb|[+coronal]| \verb|_| 
\end{tabular}

\pagebreak

\emph{Kishambaa:}\\\begin{longtable}{ll}\toprule
Surface form & UR
\\ \midrule
\textipa{tagi}&\textipa{tagi}\\
\textipa{kitabu}&\textipa{kitabu}\\
\textipa{paalika}&\textipa{paalika}\\
\textipa{ni}&\textipa{ni}\\
\textipa{Nombe}&\textipa{Nombe}\\
\textipa{matagi}&\textipa{matagi}\\
\textipa{dodoa}&\textipa{dodoa}\\
\textipa{goSa}&\textipa{goSa}\\
\textipa{babu}&\textipa{babu}\\
\textipa{ndimi}&\textipa{ndimi}\\
\textipa{Ngoto}&\textipa{Ngoto}\\
\textipa{mbeu}&\textipa{mbeu}\\
\textipa{\r*nt\super humbii}&\textipa{nt\super humbii}\\
\textipa{Nok\super huNguni}&\textipa{Nok\super huNguni}\\
\textipa{\r*mp\super heho}&\textipa{mp\super heho}\\
\bottomrule\end{longtable}

\begin{longtable}{ll}\toprule
\emph{The surface form...}&\emph{Is underlyingly...}
\\ \midrule
\textipa{\r*m}&\textipa{m}\\
\textipa{\r*n}&\textipa{n}\\
\bottomrule\end{longtable}

\begin{tabular}{l}\emph{Rules: }\\
\verb|[+nasal]| $\to$ \verb|[-voice]| /  \verb|_| \verb|[-voice]|
\end{tabular}

\pagebreak

\emph{Thai:}\\\begin{longtable}{ll}\toprule
Surface form & UR
\\ \midrule
\textipa{bil}&\textipa{bil}\\
\textipa{m\"u\"u}&\textipa{m\"u\"u}\\
\textipa{rak\textcorner }&\textipa{rak}\\
\textipa{baa}&\textipa{baa}\\
\textipa{loN}&\textipa{loN}\\
\textipa{br\"u\"u}&\textipa{br\"u\"u}\\
\textipa{haa}&\textipa{haa}\\
\textipa{plaa}&\textipa{plaa}\\
\textipa{dii}&\textipa{dii}\\
\textipa{t\super Saan}&\textipa{t\super Saan}\\
\textipa{t\super hee}&\textipa{t\super hee}\\
\textipa{t\super hruumEEn}&\textipa{t\super hruumEEn}\\
\textipa{k\super hEN}&\textipa{k\super hEN}\\
\textipa{panjaa}&\textipa{panjaa}\\
\textipa{l@@j}&\textipa{l@@j}\\
\textipa{p\super hjaa}&\textipa{p\super hjaa}\\
\textipa{l\"uak\textcorner }&\textipa{l\"uak}\\
\textipa{klaaN}&\textipa{klaaN}\\
\textipa{t\super S\super hat\textcorner }&\textipa{t\super S\super hat}\\
\textipa{traa}&\textipa{traa}\\
\textipa{riip\textcorner }&\textipa{riip}\\
\textipa{OOk\textcorner }&\textipa{OOk}\\
\textipa{p\super hrEE}&\textipa{p\super hrEE}\\
\textipa{ki@}&\textipa{ki@}\\
\textipa{k\super hwaa}&\textipa{k\super hwaa}\\
\textipa{kEE}&\textipa{kEE}\\
\textipa{draj}&\textipa{draj}\\
\textipa{d\"uN}&\textipa{d\"uN}\\
\textipa{kan}&\textipa{kan}\\
\textipa{t\super Su@k\textcorner }&\textipa{t\super Su@k}\\
\textipa{p\super hleeN}&\textipa{p\super hleeN}\\
\textipa{t\super S\super han}&\textipa{t\super S\super han}\\
\textipa{staaN}&\textipa{staaN}\\
\textipa{rap\textcorner }&\textipa{rap}\\
\textipa{jiisip\textcorner }&\textipa{jiisip}\\
\textipa{p\super haa}&\textipa{p\super haa}\\
\textipa{k\super haa}&\textipa{k\super haa}\\
\textipa{dam}&\textipa{dam}\\
\textipa{raaj}&\textipa{raaj}\\
\textipa{tit\textcorner }&\textipa{tit}\\
\textipa{sip\textcorner }&\textipa{sip}\\
\textipa{pen}&\textipa{pen}\\
\bottomrule\end{longtable}

\begin{longtable}{ll}\toprule
\emph{The surface form...}&\emph{Is underlyingly...}
\\ \midrule
\textipa{t\textcorner }&\textipa{t}\\
\textipa{p\textcorner }&\textipa{p}\\
\textipa{k\textcorner }&\textipa{k}\\
\bottomrule\end{longtable}

\begin{tabular}{l}\emph{Rules: }\\
\verb|[-voice]| $\to$ \verb|[+unreleased]| /  \verb|_| \#
\end{tabular}

\pagebreak

\emph{Palauan:}\\\begin{longtable}{ll}\toprule
Surface form & UR
\\ \midrule
\textipa{k@D@}&\textipa{k@D@}\\
\textipa{b@Duk}&\textipa{b@Duk}\\
\textipa{Diak}&\textipa{Diak}\\
\textipa{diak}&\textipa{diak}\\
\textipa{maTtNoT}&\textipa{madtNod}\\
\textipa{De:l}&\textipa{De:l}\\
\textipa{de:l}&\textipa{de:l}\\
\textipa{Dios@P}&\textipa{Dios@P}\\
\textipa{dios@P}&\textipa{dios@P}\\
\textipa{Dik}&\textipa{Dik}\\
\textipa{dik}&\textipa{dik}\\
\textipa{kuT}&\textipa{kud}\\
\textipa{PoDiN@l}&\textipa{PoDiN@l}\\
\textipa{koaT}&\textipa{koad}\\
\textipa{eaN@T}&\textipa{eaN@d}\\
\textipa{N@rar@D@}&\textipa{N@rar@D@}\\
\textipa{baT}&\textipa{bad}\\
\textipa{ieDlP@Dip}&\textipa{ieDlP@Dip}\\
\textipa{k@Deb}&\textipa{k@Deb}\\
\textipa{m@D@Nei}&\textipa{m@D@Nei}\\
\textipa{uDouT}&\textipa{uDoud}\\
\textipa{olDak}&\textipa{olDak}\\
\bottomrule\end{longtable}

\begin{longtable}{ll}\toprule
\emph{The surface form...}&\emph{Is underlyingly...}
\\ \midrule
\textipa{T}&\textipa{d}\\
\bottomrule\end{longtable}

\begin{tabular}{l}\emph{Rules: }\\
\textipa{d} $\to$ \textipa{T} / \verb|[ ]| \verb|_| 
\end{tabular}

\pagebreak

\emph{Quechua (Cuzco dialect):}\\\begin{longtable}{ll}\toprule
Surface form & UR
\\ \midrule
\textipa{qori}&\textipa{quri}\\
\textipa{t\super SoXlu}&\textipa{t\super SuXlu}\\
\textipa{qomir}&\textipa{qumir}\\
\textipa{niNri}&\textipa{niNri}\\
\textipa{moqo}&\textipa{muqu}\\
\textipa{hoqara}&\textipa{huqara}\\
\textipa{p\super hulju}&\textipa{p\super hulju}\\
\textipa{jujaN}&\textipa{jujaN}\\
\textipa{tulju}&\textipa{tulju}\\
\textipa{api}&\textipa{api}\\
\textipa{suti}&\textipa{suti}\\
\textipa{o\;Nqoj}&\textipa{uNquj}\\
\textipa{t\super Silwi}&\textipa{t\super Silwi}\\
\textipa{t\super S\super hit\super SiN}&\textipa{t\super S\super hit\super SiN}\\
\textipa{t\super S\super ha\;Nqaj}&\textipa{t\super S\super haNqaj}\\
\textipa{a\;Nqosaj}&\textipa{aNqusaj}\\
\textipa{qet\super SuN}&\textipa{qit\super SuN}\\
\textipa{pisqo}&\textipa{pisqu}\\
\textipa{musoX}&\textipa{musuX}\\
\textipa{t\super SuNka}&\textipa{t\super SuNka}\\
\textipa{ja\;NqaN}&\textipa{jaNqaN}\\
\textipa{t\super Sulju}&\textipa{t\super Sulju}\\
\textipa{qhelja}&\textipa{qhilja}\\
\textipa{qe\;Nqo}&\textipa{qiNqu}\\
\textipa{t\super SeqaN}&\textipa{t\super SiqaN}\\
\textipa{qaN}&\textipa{qaN}\\
\textipa{noqa}&\textipa{nuqa}\\
\textipa{t\super Saxra}&\textipa{t\super Saxra}\\
\textipa{t\super SeXniN}&\textipa{t\super SiXniN}\\
\textipa{soXta}&\textipa{suXta}\\
\textipa{aXna}&\textipa{aXna}\\
\textipa{ljixlja}&\textipa{ljixlja}\\
\textipa{qosa}&\textipa{qusa}\\
\textipa{qara}&\textipa{qara}\\
\textipa{alqo}&\textipa{alqu}\\
\textipa{se\;Nqa}&\textipa{siNqa}\\
\textipa{karu}&\textipa{karu}\\
\textipa{atoX}&\textipa{atuX}\\
\textipa{qaNkuna}&\textipa{qaNkuna}\\
\textipa{pusaX}&\textipa{pusaX}\\
\textipa{teXwaj}&\textipa{tiXwaj}\\
\textipa{t\super Saki}&\textipa{t\super Saki}\\
\textipa{wateX}&\textipa{watiX}\\
\textipa{aNka}&\textipa{aNka}\\
\textipa{waXtaj}&\textipa{waXtaj}\\
\textipa{haku}&\textipa{haku}\\
\textipa{waqaj}&\textipa{waqaj}\\
\textipa{kaNka}&\textipa{kaNka}\\
\textipa{waxt\super Sa}&\textipa{waxt\super Sa}\\
\textipa{waleX}&\textipa{waliX}\\
\textipa{t\super hakaj}&\textipa{t\super hakaj}\\
\textipa{reXsisqa}&\textipa{riXsisqa}\\
\bottomrule\end{longtable}

\begin{longtable}{ll}\toprule
\emph{The surface form...}&\emph{Is underlyingly...}
\\ \midrule
\textipa{e}&\textipa{i}\\
\textipa{o}&\textipa{u}\\
\textipa{\;N}&\textipa{N}\\
\bottomrule\end{longtable}

\begin{tabular}{l}\emph{Rules: }\\
\verb|[+high]| $\to$ \textipa{\;N} /  \verb|_| \textipa{q}\\\verb|[ ]| $\to$ \verb|[-high]| / \textipa{q} \verb|_| \\\textipa{u} $\to$ \textipa{o} /  \verb|_| \verb|[+back -high]|\\\verb|[+sonorant]| $\to$ \verb|[-high]| / \verb|[-palletized]| \verb|_| \verb|[-anterior]|* \verb|[-nasal +back]|
\end{tabular}

\pagebreak

\emph{Lhasa Tibetan:}\\\begin{longtable}{ll}\toprule
Surface form & UR
\\ \midrule
\textipa{aNgu}&\textipa{aNku}\\
\textipa{aN\.*t\~a\~a}&\textipa{aN\.*t\~a\~a}\\
\textipa{aNba}&\textipa{aNpa}\\
\textipa{apsoo}&\textipa{apsoo}\\
\textipa{amt\super SOO}&\textipa{amt\super SOO}\\
\textipa{tuk\.*t\"u\"u}&\textipa{tuk\.*t\"u\"u}\\
\textipa{amto}&\textipa{amto}\\
\textipa{1Gu}&\textipa{1Gu}\\
\textipa{1mt\super Si}&\textipa{1mt\super Si}\\
\textipa{u\.*t1}&\textipa{u\.*t1}\\
\textipa{uB\textiota \textiota }&\textipa{uB\textiota \textiota }\\
\textipa{ea}&\textipa{ea}\\
\textipa{embo}&\textipa{empo}\\
\textipa{UUtsi}&\textipa{UUtsi}\\
\textipa{qa}&\textipa{ka}\\
\textipa{qaa}&\textipa{kaa}\\
\textipa{qaNba}&\textipa{kaNpa}\\
\textipa{qamba}&\textipa{kampa}\\
\textipa{qam}&\textipa{kam}\\
\textipa{qamtoo}&\textipa{kamtoo}\\
\textipa{qaaBo}&\textipa{kaaBo}\\
\textipa{k1k\.*ti}&\textipa{k1k\.*ti}\\
\textipa{k1Bu}&\textipa{k1Bu}\\
\textipa{k11Nguu}&\textipa{k11Nkuu}\\
\textipa{kik}&\textipa{kik}\\
\textipa{ki\.*tuu}&\textipa{ki\.*tuu}\\
\textipa{k\textiota \textiota cuu}&\textipa{k\textiota \textiota cuu}\\
\textipa{k\textiota \textiota rii}&\textipa{k\textiota \textiota rii}\\
\textipa{kiiGuu}&\textipa{kiiGuu}\\
\textipa{ku}&\textipa{ku}\\
\textipa{kupc1}&\textipa{kupc1}\\
\textipa{kupcaa}&\textipa{kupcaa}\\
\textipa{kEnca}&\textipa{kEnca}\\
\textipa{kEmbo}&\textipa{kEmpo}\\
\textipa{keG\"o\"o}&\textipa{keG\"o\"o}\\
\textipa{kerBa}&\textipa{kerBa}\\
\textipa{qo}&\textipa{ko}\\
\textipa{qomba}&\textipa{kompa}\\
\textipa{qOr}&\textipa{kOr}\\
\textipa{qOOOO}&\textipa{kOOOO}\\
\textipa{t\super S\super hea}&\textipa{t\super S\super hea}\\
\textipa{t\super S\super huGum}&\textipa{t\super S\super huGum}\\
\textipa{topcaa}&\textipa{topcaa}\\
\textipa{t\super ho\~o\~o}&\textipa{t\super ho\~o\~o}\\
\textipa{\.*taa\~a\~a}&\textipa{\.*taa\~a\~a}\\
\textipa{\.*tuG1}&\textipa{\.*tuG1}\\
\textipa{\.*tu\;N\;Go}&\textipa{\.*tuNgo}\\
\textipa{nE\;N\;Gaa}&\textipa{nENgaa}\\
\textipa{pa\;N\;GOO}&\textipa{paNgOO}\\
\textipa{pEEB\~a\~a}&\textipa{pEEB\~a\~a}\\
\textipa{sim\;G\~a\~a}&\textipa{simg\~a\~a}\\
\bottomrule\end{longtable}

\begin{longtable}{ll}\toprule
\emph{The surface form...}&\emph{Is underlyingly...}
\\ \midrule
\textipa{b}&\textipa{p}\\
\textipa{d}&\textipa{t}\\
\textipa{\;G}&\textipa{g}\\
\textipa{\.*d}&\textipa{\.*t}\\
\textipa{\;N}&\textipa{N}\\
\textipa{q}&\textipa{k}\\
\textipa{g}&\textipa{k}\\
\bottomrule\end{longtable}

\begin{tabular}{l}\emph{Rules: }\\
\verb|[-anterior -vowel]| $\to$ \verb|[-sonorant -high]| /  \verb|_| \verb|[+high]|* \verb|[+back -high]|\\\verb|[-coronal]| $\to$ \verb|[+voice]| / \verb|[-vowel]| \verb|_| 
\end{tabular}

\pagebreak

\emph{Axininca Campa:}\\\begin{longtable}{ll|l}\toprule
$\varnothing$ $+$stem$+$ $\varnothing$ & \textipa{no} $+$stem$+$ \textipa{ti} & UR
\\ \midrule
\textipa{toniro} & \textipa{notoniroti} & \textipa{toniro}\\
\textipa{jaarato} & \textipa{nojaaratoti} & \textipa{jaarato}\\
\textipa{kanari} & \textipa{nojanariti} & \textipa{kanari}\\
\textipa{kosiri} & \textipa{nojosiriti} & \textipa{kosiri}\\
\textipa{pisiro} & \textipa{nowisiroti} & \textipa{wisiro}\\
\textipa{porita} & \textipa{noworitati} & \textipa{worita}\\
\bottomrule\end{longtable}
\begin{tabular}{l}\emph{Rules: }\\
\textipa{w} $\to$ \textipa{p} / \# \verb|_| \\\textipa{k} $\to$ \textipa{j} / \verb|[ ]| \verb|_| 
\end{tabular}

\pagebreak

\emph{Kikuyu:}\\\begin{longtable}{l|l}\toprule
\textipa{ko} $+$stem$+$ \textipa{a} & UR
\\ \midrule
\textipa{GotENEra} & \textipa{tENEr}\\
\textipa{Gokuua} & \textipa{kuu}\\
\textipa{Gokoora} & \textipa{koor}\\
\textipa{koruGa} & \textipa{ruG}\\
\textipa{kooria} & \textipa{ori}\\
\textipa{komE\~na} & \textipa{mE\~n}\\
\textipa{kohOta} & \textipa{hOt}\\
\textipa{Got\super Sina} & \textipa{t\super Sin}\\
\textipa{koGeera} & \textipa{Geer}\\
\textipa{koina} & \textipa{in}\\
\textipa{Got\super Suuka} & \textipa{t\super Suuk}\\
\textipa{Gokaja} & \textipa{kaj}\\
\textipa{koGaja} & \textipa{Gaj}\\
\bottomrule\end{longtable}
\begin{tabular}{l}\emph{Rules: }\\
\verb|[ ]| $\to$ \textipa{G} / \# \verb|_| \verb|[ ]| \verb|[-low -voice]|
\end{tabular}

\pagebreak

\emph{Korean:}\\\begin{longtable}{ll|l}\toprule
$\varnothing$ $+$stem$+$ \textipa{a} & $\varnothing$ $+$stem$+$ \textipa{ko} & UR
\\ \midrule
\textipa{ip@} & \textipa{ipko} & \textipa{ip}\\
\textipa{kup@} & \textipa{kupko} & \textipa{kup}\\
\textipa{kap\super ha} & \textipa{kapko} & \textipa{kap\super h}\\
\textipa{cip\super h@} & \textipa{cipko} & \textipa{cip\super h}\\
\textipa{tata} & \textipa{tatko} & \textipa{tat}\\
\textipa{put\super h@} & \textipa{putko} & \textipa{put\super h}\\
\textipa{m@k@} & \textipa{m@kko} & \textipa{m@k}\\
\textipa{t\super Suk@} & \textipa{t\super Sukko} & \textipa{t\super Suk}\\
\textipa{ik@} & \textipa{ikko} & \textipa{ik}\\
\textipa{taka} & \textipa{takko} &  -- \\
\textipa{kaka} & \textipa{kakko} &  -- \\
\textipa{s@k@} & \textipa{s@kko} &  -- \\
\bottomrule\end{longtable}
\begin{tabular}{l}\emph{Rules: }\\
\verb|[ ]| $\to$ \verb|[-aspirated]| /  \verb|_| \verb|[-voice]|\\\textipa{a} $\to$ \textipa{@} / \verb|[-low]| \verb|[ ]| \verb|_| 
\end{tabular}

\pagebreak

\emph{Hungarian:}\\\begin{longtable}{llll|l}\toprule
$\varnothing$ $+$stem$+$ $\varnothing$ & $\varnothing$ $+$stem$+$ \textipa{ban} & $\varnothing$ $+$stem$+$ \textipa{to:l} & $\varnothing$ $+$stem$+$ \textipa{nak} & UR
\\ \midrule
\textipa{kalap} & \textipa{kalabban} & \textipa{kalapto:l} & \textipa{kalapnak} & \textipa{kalap}\\
\textipa{ku:t} & \textipa{ku:dban} & \textipa{ku:tto:l} & \textipa{ku:tnak} & \textipa{ku:t}\\
\textipa{Za:k} & \textipa{Za:gban} & \textipa{Za:kto:l} & \textipa{Za:knak} & \textipa{Za:k}\\
\textipa{re:s} & \textipa{re:zben} & \textipa{re:st\"o:l} & \textipa{re:snek} & \textipa{re:s}\\
\textipa{Sro:f} & \textipa{Sro:vban} & \textipa{Sro:fto:l} & \textipa{Sro:fnak} & \textipa{Sro:f}\\
\textipa{laka:S} & \textipa{laka:Zban} & \textipa{laka:Sto:l} & \textipa{laka:Snak} & \textipa{laka:S}\\
\textipa{ketret\super s} & \textipa{ketred\super zben} & \textipa{ketret\super st\"o:l} & \textipa{ketret\super snek} & \textipa{ketret\super s}\\
\textipa{test} & \textipa{tezdben} & \textipa{testt\"o:l} & \textipa{testnek} & \textipa{tezt}\\
\textipa{rab} & \textipa{rabban} & \textipa{rapto:l} & \textipa{rabnak} & \textipa{rab}\\
\textipa{ka:d} & \textipa{ka:dban} & \textipa{ka:tto:l} & \textipa{ka:dnak} & \textipa{ka:d}\\
\textipa{meleg} & \textipa{melegben} & \textipa{melekt\"o:l} & \textipa{melegnek} & \textipa{meleg}\\
\textipa{vi:z} & \textipa{vi:zben} & \textipa{vi:st\"o:l} & \textipa{vi:znek} & \textipa{vi:z}\\
\textipa{vara:Z} & \textipa{vara:Zban} & \textipa{vara:Sto:l} & \textipa{vara:Znak} &  -- \\
\textipa{a:g\super j} & \textipa{a:g\super jban} & \textipa{a:k\super jto:l} & \textipa{a:g\super jnak} &  -- \\
\textipa{sem} & \textipa{semben} & \textipa{semt\"o:l} & \textipa{semnek} &  -- \\
\textipa{b\"un} & \textipa{b\"unben} & \textipa{b\"unt\"o:l} & \textipa{b\"unnek} &  -- \\
\textipa{toro\~n} & \textipa{toro\~nban} & \textipa{toro\~nto:l} & \textipa{toro\~nnak} &  -- \\
\textipa{fal} & \textipa{falban} & \textipa{falto:l} & \textipa{falnak} &  -- \\
\textipa{\"o:r} & \textipa{\"o:rben} & \textipa{\"o:rt\"o:l} & \textipa{\"o:rnek} &  -- \\
\textipa{sa:j} & \textipa{sa:jban} & \textipa{sa:jto:l} & \textipa{sa:jnak} &  -- \\
\bottomrule\end{longtable}
\begin{tabular}{l}\emph{Rules: }\\
\verb|[-sonorant]| $\to$ \verb|[+voice]| /  \verb|_| \verb|[-sonorant +voice]|\\\verb|[-sonorant]| $\to$ \verb|[-voice]| /  \verb|_| \verb|[-voice]|\\\verb|[+vowel]| $\to$ \verb|[-back -low]| / \verb|[-back +vowel]| \verb|[ ]|* \verb|_| 
\end{tabular}

\pagebreak

\emph{Kikuria:}\\\begin{longtable}{ll|l}\toprule
$\varnothing$ $+$stem$+$ \textipa{a} & $\varnothing$ $+$stem$+$ \textipa{era} & UR
\\ \midrule
\textipa{suraaNga} & \textipa{suraaNgera} & \textipa{suraaNg}\\
\textipa{taaNgata} & \textipa{taaNgatera} & \textipa{taaNgat}\\
\textipa{baamba} & \textipa{baambera} & \textipa{baamb}\\
\textipa{reenda} & \textipa{reendera} & \textipa{riend}\\
\textipa{rema} & \textipa{remera} & \textipa{rem}\\
\textipa{hoora} & \textipa{hoorera} & \textipa{hoor}\\
\textipa{roma} & \textipa{romera} & \textipa{rom}\\
\textipa{sooka} & \textipa{sookera} & \textipa{sook}\\
\textipa{tat\super Sora} & \textipa{tat\super Sorera} & \textipa{tat\super Sor}\\
\textipa{siika} & \textipa{seekera} & \textipa{siik}\\
\textipa{tiga} & \textipa{tegera} & \textipa{tig}\\
\textipa{ruga} & \textipa{rogera} & \textipa{rug}\\
\textipa{suka} & \textipa{sokera} &  -- \\
\textipa{huuta} & \textipa{hootera} &  -- \\
\textipa{riiNga} & \textipa{reeNgera} &  -- \\
\textipa{siinda} & \textipa{seendera} &  -- \\
\bottomrule\end{longtable}
\begin{tabular}{l}\emph{Rules: }\\
\verb|[+tense]| $\to$ \verb|[-high]| /  \verb|_| \verb|[-liquid]|* \textipa{e}
\end{tabular}

\pagebreak

\emph{Farsi:}\\\begin{longtable}{ll|l}\toprule
$\varnothing$ $+$stem$+$ $\varnothing$ & $\varnothing$ $+$stem$+$ \textipa{an} & UR
\\ \midrule
\textipa{z\ae n} & \textipa{z\ae nan} & \textipa{z\ae n}\\
\textipa{l\ae b} & \textipa{l\ae ban} & \textipa{l\ae b}\\
\textipa{h\ae sud} & \textipa{h\ae sudan} & \textipa{h\ae sud}\\
\textipa{b\ae rad\ae r} & \textipa{b\ae rad\ae ran} & \textipa{b\ae rad\ae r}\\
\textipa{bozorg} & \textipa{bozorgan} & \textipa{bozorg}\\
\textipa{m\ae leke} & \textipa{m\ae lekean} & \textipa{m\ae leke}\\
\textipa{valede} & \textipa{valedean} & \textipa{valede}\\
\textipa{k\ae bire} & \textipa{k\ae birean} & \textipa{k\ae bire}\\
\textipa{ahu} & \textipa{ahuan} & \textipa{ahu}\\
\textipa{hamele} & \textipa{hamelean} & \textipa{hamele}\\
\textipa{bat\super St\super Se} & \textipa{bat\super St\super Segan} & \textipa{bat\super St\super Seg}\\
\textipa{setare} & \textipa{setaregan} & \textipa{setareg}\\
\textipa{b\ae nde} & \textipa{b\ae ndegan} & \textipa{b\ae ndeg}\\
\textipa{azade} & \textipa{azadegan} &  -- \\
\textipa{divane} & \textipa{divanegan} &  -- \\
\bottomrule\end{longtable}
\begin{tabular}{l}\emph{Rules: }\\
\verb|[ ]| $\to$ $\varnothing$ / \textipa{e} \verb|_| \#
\end{tabular}

\pagebreak

\emph{Tibetan:}\\\begin{longtable}{ll}\toprule
Number & Surface form
\\ \midrule
10 & \textipa{d\super Zu}\\
1 & \textipa{d\super Zig}\\
11 & \textipa{d\super Zugd\super Zig}\\
4 & \textipa{Si}\\
14 & \textipa{d\super ZubSi}\\
40 & \textipa{Sibd\super Zu}\\
9 & \textipa{gu}\\
19 & \textipa{d\super Zurgu}\\
90 & \textipa{gubd\super Zu}\\
5 & \textipa{Na}\\
15 & \textipa{d\super ZuNa}\\
50 & \textipa{Nabd\super Zu}\\
\bottomrule\end{longtable}
\begin{tabular}{l}\emph{Rules: }\\
\verb|[-nasal]| $\to$ $\varnothing$ / \# \verb|_| 
\end{tabular}

\pagebreak

\emph{Makonde:}\\\begin{longtable}{lll|l}\toprule
$\varnothing$ $+$stem$+$ \textipa{\'{a}Nga} & $\varnothing$ $+$stem$+$ \textipa{\'{i}le} & $\varnothing$ $+$stem$+$ \textipa{a} & UR
\\ \midrule
\textipa{am\'{a}Nga} & \textipa{am\'{i}le} & \textipa{\'{a}ma} & \textipa{\'{a}m}\\
\textipa{tav\'{a}Nga} & \textipa{tav\'{i}le} & \textipa{t\'{a}va} & \textipa{tav}\\
\textipa{ak\'{a}Nga} & \textipa{ak\'{i}le} & \textipa{\'{a}ka} & \textipa{\'{a}k}\\
\textipa{pat\'{a}Nga} & \textipa{pat\'{i}le} & \textipa{p\'{o}ta} & \textipa{p\'{o}t}\\
\textipa{tat\'{a}Nga} & \textipa{tat\'{i}le} & \textipa{t\'{o}ta} & \textipa{t\'{o}t}\\
\textipa{dab\'{a}Nga} & \textipa{dab\'{i}le} & \textipa{d\'{o}ba} & \textipa{d\'{o}b}\\
\textipa{av\'{a}Nga} & \textipa{av\'{i}le} & \textipa{\'{o}va} & \textipa{\'{o}v}\\
\textipa{am\'{a}Nga} & \textipa{am\'{i}le} & \textipa{\'{o}ma} & \textipa{\'{o}m}\\
\textipa{tap\'{a}Nga} & \textipa{tap\'{i}le} & \textipa{t\'{e}pa} & \textipa{t\'{e}p}\\
\textipa{pat\'{a}Nga} & \textipa{pat\'{i}le} & \textipa{p\'{e}ta} & \textipa{p\'{e}t}\\
\textipa{av\'{a}Nga} & \textipa{av\'{i}le} & \textipa{\'{e}va} & \textipa{\'{e}v}\\
\textipa{bab\'{a}Nga} & \textipa{bab\'{i}le} & \textipa{b\'{e}ba} & \textipa{b\'{e}b}\\
\textipa{ut\'{a}Nga} & \textipa{ut\'{i}le} & \textipa{\'{u}ta} & \textipa{ut}\\
\textipa{luk\'{a}Nga} & \textipa{luk\'{i}le} & \textipa{l\'{u}ka} & \textipa{luk}\\
\textipa{lum\'{a}Nga} & \textipa{lum\'{i}le} & \textipa{l\'{u}ma} & \textipa{lum}\\
\textipa{uNg\'{a}Nga} & \textipa{uNg\'{i}le} & \textipa{\'{u}Nga} & \textipa{uNg}\\
\textipa{iv\'{a}Nga} & \textipa{iv\'{i}le} & \textipa{\'{i}va} & \textipa{iv}\\
\textipa{pit\'{a}Nga} & \textipa{pit\'{i}le} & \textipa{p\'{i}ta} & \textipa{pit}\\
\textipa{imb\'{a}Nga} & \textipa{imb\'{i}le} & \textipa{\'{i}mba} &  -- \\
\textipa{lim\'{a}Nga} & \textipa{lim\'{i}le} & \textipa{l\'{i}ma} &  -- \\
\bottomrule\end{longtable}
\begin{tabular}{l}\emph{Rules: }\\
\verb|[+vowel]| $\to$ \verb|[+highTone]| /  \verb|_| \verb|[-highTone]|* \verb|[+vowel]| \#\\\verb|[+highTone]| $\to$ \textipa{a} /  \verb|_| \verb|[ ]| \verb|[+highTone]|
\end{tabular}

\pagebreak

\emph{North Saami:}\\\begin{longtable}{ll|l}\toprule
$\varnothing$ $+$stem$+$ $\varnothing$ & $\varnothing$ $+$stem$+$ \textipa{in} & UR
\\ \midrule
\textipa{varit} & \textipa{varihin} & \textipa{varih}\\
\textipa{oahpis} & \textipa{oahpisin} & \textipa{oahpis}\\
\textipa{t\super SoarvuS} & \textipa{t\super SoarvuSin} & \textipa{t\super SoarvuS}\\
\textipa{lottaaS} & \textipa{lottaad\super Zin} & \textipa{lottaad\super Z}\\
\textipa{t\super Suoivvat} & \textipa{t\super Suoivvagin} & \textipa{t\super Suoivvag}\\
\textipa{ahhkut} & \textipa{ahhkubin} & \textipa{ahhkub}\\
\textipa{suohkat} & \textipa{suohkaDin} & \textipa{suohkaD}\\
\textipa{heed\super ZoS} & \textipa{heed\super Zod\super Zin} & \textipa{heed\super Zod\super Z}\\
\textipa{aad\super Zd\super Zut} & \textipa{aad\super Zd\super Zubin} & \textipa{aad\super Zd\super Zub}\\
\textipa{bissobeahtset} & \textipa{bissobeahtsehin} & \textipa{bissobeahtseh}\\
\textipa{t\super Seahtsit} & \textipa{t\super Seahtsibin} & \textipa{t\super Seahtsib}\\
\textipa{jaaPmin} & \textipa{jaaPmimin} & \textipa{jaaPmim}\\
\textipa{t\super Suoivat} & \textipa{t\super Suoivagin} & \textipa{t\super Suoivag}\\
\textipa{laageS} & \textipa{laaged\super Zin} & \textipa{laaged\super Z}\\
\textipa{gahpir} & \textipa{gahpirin} & \textipa{gahpir}\\
\textipa{gaauhtsis} & \textipa{gaauhtsisin} & \textipa{gaauhtsis}\\
\textipa{aaslat} & \textipa{aaslagin} & \textipa{aaslag}\\
\textipa{baDooSgaattset} & \textipa{baDooSgaattsebin} &  -- \\
\textipa{ahhkit} & \textipa{ahhkiDin} &  -- \\
\textipa{bahaanaalat} & \textipa{bahaanaalagin} &  -- \\
\textipa{beStor} & \textipa{beStorin} &  -- \\
\textipa{heevemeahhtun} & \textipa{heevemeahhtunin} &  -- \\
\textipa{beed\super Zot} & \textipa{beed\super Zohin} &  -- \\
\textipa{bissomeahtun} & \textipa{bissomeahtumin} &  -- \\
\textipa{laDas} & \textipa{laDasin} &  -- \\
\textipa{heaijusmielat} & \textipa{heaijusmielagin} &  -- \\
\textipa{heaNkkan} & \textipa{heaNkkanin} &  -- \\
\textipa{jaman} & \textipa{jamanin} &  -- \\
\bottomrule\end{longtable}
\begin{tabular}{l}\emph{Rules: }\\
\verb|[-liquid -nasal -sibilant]| $\to$ \textipa{t} /  \verb|_| \#\\\verb|[ ]| $\to$ \verb|[+coronal]| /  \verb|_| \#\\\verb|[-sonorant +voice]| $\to$ \textipa{S} /  \verb|_| \#
\end{tabular}

\pagebreak

\emph{Samoan:}\\\begin{longtable}{ll|l}\toprule
$\varnothing$ $+$stem$+$ $\varnothing$ & $\varnothing$ $+$stem$+$ \textipa{ia} & UR
\\ \midrule
\textipa{olo} & \textipa{oloia} & \textipa{olo}\\
\textipa{lafo} & \textipa{lafoia} & \textipa{lafo}\\
\textipa{usu} & \textipa{usuia} & \textipa{usu}\\
\textipa{tau} & \textipa{tauia} & \textipa{tau}\\
\textipa{taui} & \textipa{tauia} & \textipa{taui}\\
\textipa{lele} & \textipa{lelea} & \textipa{lele}\\
\textipa{tafe} & \textipa{tafea} & \textipa{tafe}\\
\textipa{tau} & \textipa{taulia} & \textipa{taul}\\
\textipa{oso} & \textipa{osofia} & \textipa{osof}\\
\textipa{valu} & \textipa{valusia} & \textipa{valus}\\
\textipa{u:} & \textipa{u:tia} &  -- \\
\bottomrule\end{longtable}
\begin{tabular}{l}\emph{Rules: }\\
\verb|[-tense]| $\to$ $\varnothing$ /  \verb|_| \#\\\verb|[+vowel]| $\to$ $\varnothing$ / \verb|[-back +tense]| \verb|_| \verb|[+vowel]|
\end{tabular}

\pagebreak

\emph{Russian:}\\\begin{longtable}{ll|l}\toprule
$\varnothing$ $+$stem$+$ $\varnothing$ & $\varnothing$ $+$stem$+$ \textipa{a} & UR
\\ \midrule
\textipa{vagon} & \textipa{vagona} & \textipa{vagon}\\
\textipa{glas} & \textipa{glaza} & \textipa{glaz}\\
\textipa{golos} & \textipa{golosa} & \textipa{golos}\\
\textipa{ras} & \textipa{raza} & \textipa{raz}\\
\textipa{les} & \textipa{lesa} & \textipa{les}\\
\textipa{porok} & \textipa{poroga} & \textipa{porog}\\
\textipa{vrak} & \textipa{vraga} & \textipa{vrag}\\
\textipa{urok} & \textipa{uroka} & \textipa{urok}\\
\textipa{tvet} & \textipa{tveta} &  -- \\
\textipa{prut} & \textipa{pruda} &  -- \\
\textipa{soldat} & \textipa{soldata} &  -- \\
\textipa{zavot} & \textipa{zavoda} &  -- \\
\textipa{xlep} & \textipa{xleba} &  -- \\
\textipa{grip} & \textipa{griba} &  -- \\
\textipa{trup} & \textipa{trupa} &  -- \\
\bottomrule\end{longtable}
\begin{tabular}{l}\emph{Rules: }\\
\verb|[-sonorant]| $\to$ \verb|[-voice]| /  \verb|_| \#
\end{tabular}

\pagebreak

\emph{English:}\\\begin{longtable}{lll|l}\toprule
$\varnothing$ $+$stem$+$ $\varnothing$ & $\varnothing$ $+$stem$+$ \textipa{d} & $\varnothing$ $+$stem$+$ \textipa{@z} & UR
\\ \midrule
\textipa{ro} & \textipa{rod} & \textipa{roz} & \textipa{ro}\\
\textipa{gr\ae b} & \textipa{gr\ae bd} & \textipa{gr\ae bz} & \textipa{gr\ae b}\\
\textipa{sim} & \textipa{simd} & \textipa{simz} & \textipa{sim}\\
\textipa{liv} & \textipa{livd} & \textipa{livz} & \textipa{liv}\\
\textipa{h@g} & \textipa{h@gd} & \textipa{h@gz} & \textipa{h@g}\\
\textipa{lUk} & \textipa{lUkt} & \textipa{lUks} & \textipa{lUk}\\
\textipa{\ae sk} & \textipa{\ae skt} & \textipa{\ae sks} & \textipa{\ae sk}\\
\textipa{wErk} & \textipa{wErkt} & \textipa{wErks} & \textipa{wErk}\\
\textipa{k\textiota s} & \textipa{k\textiota st} & \textipa{k\textiota s@z} & \textipa{k\textiota s}\\
\textipa{f\textiota S} & \textipa{f\textiota St} & \textipa{f\textiota S@z} & \textipa{f\textiota S}\\
\textipa{kw\textiota z} & \textipa{kw\textiota zd} & \textipa{kw\textiota z@z} & \textipa{kw\textiota z}\\
\textipa{b2z} & \textipa{b2zd} & \textipa{b2z@z} & \textipa{b2z}\\
\textipa{wet} & \textipa{wet@d} & \textipa{wets} & \textipa{wet}\\
\textipa{wed} & \textipa{wed@d} & \textipa{wedz} & \textipa{wed}\\
\textipa{l\textiota ft} & \textipa{l\textiota ft@d} & \textipa{l\textiota fts} & \textipa{l\textiota ft}\\
\bottomrule\end{longtable}
\begin{tabular}{l}\emph{Rules: }\\
\verb|[+vowel]| $\to$ $\varnothing$ / \verb|[+continuant]| \verb|[-sibilant]| \verb|_| \\$\varnothing$ $\to$ \textipa{@} / \verb|[-continuant +coronal]| \verb|_| \verb|[-continuant]|\\\verb|[-sonorant]| $\to$ \verb|[-voice]| / \verb|[-voice]| \verb|_| 
\end{tabular}

\pagebreak

\emph{Finnish:}\\\begin{longtable}{ll|l}\toprule
$\varnothing$ $+$stem$+$ $\varnothing$ & $\varnothing$ $+$stem$+$ \textipa{\ae } & UR
\\ \midrule
\textipa{aamu} & \textipa{aamua} & \textipa{aamu}\\
\textipa{hopea} & \textipa{hopeaa} & \textipa{hopea}\\
\textipa{katto} & \textipa{kattoa} & \textipa{katto}\\
\textipa{kello} & \textipa{kelloa} & \textipa{kello}\\
\textipa{k\"ulm\ae } & \textipa{k\"ulm\ae \ae } & \textipa{k\"ulm\ae }\\
\textipa{koulu} & \textipa{koulua} & \textipa{koulu}\\
\textipa{lintu} & \textipa{lintua} & \textipa{lintu}\\
\textipa{h\"ull\"u} & \textipa{h\"ull\"u\ae } & \textipa{h\"ull\"u}\\
\textipa{k\"ompel\"o} & \textipa{k\"ompel\"o\ae } & \textipa{k\"ompel\"o}\\
\textipa{n\ae k\"o} & \textipa{n\ae k\"o\ae } & \textipa{n\ae k\"o}\\
\textipa{joki} & \textipa{jokea} & \textipa{joke}\\
\textipa{kivi} & \textipa{kive\ae } & \textipa{kive}\\
\textipa{muuri} & \textipa{muuria} & \textipa{muuri}\\
\textipa{naapuri} & \textipa{naapuria} &  -- \\
\textipa{nimi} & \textipa{nime\ae } &  -- \\
\textipa{kiirehti} & \textipa{kiirehti\ae } &  -- \\
\bottomrule\end{longtable}
\begin{tabular}{l}\emph{Rules: }\\
\verb|[+low]| $\to$ \verb|[+back]| / \verb|[+rounded +back]| \verb|[ ]|* \verb|_| \\\textipa{e} $\to$ \verb|[+high]| /  \verb|_| \#
\end{tabular}

\pagebreak

\emph{Lithuanian:}\\\begin{longtable}{l|l}\toprule
\textipa{at} $+$stem$+$ \textipa{ti} & UR
\\ \midrule
\textipa{ateiti} & \textipa{ei}\\
\textipa{atimti} & \textipa{im}\\
\textipa{atleisti} & \textipa{leis}\\
\textipa{atlikti} & \textipa{lik}\\
\textipa{atko:pti} & \textipa{ko:p}\\
\textipa{atkurti} & \textipa{kur}\\
\textipa{adbekti} & \textipa{bek}\\
\textipa{adgauti} & \textipa{gau}\\
\textipa{adbukti} & \textipa{buk}\\
\textipa{adgimti} & \textipa{gim}\\
\textipa{atiduoti} & \textipa{iduo}\\
\textipa{atidari:ti} & \textipa{idari:}\\
\textipa{atideti} &  -- \\
\textipa{atiteisti} &  -- \\
\bottomrule\end{longtable}
\begin{tabular}{l}\emph{Rules: }\\
\verb|[-vowel]| $\to$ \textipa{d} /  \verb|_| \verb|[-sonorant +voice]|
\end{tabular}

\pagebreak

\emph{Kerewe:}\\\begin{longtable}{llllllll|l}\toprule
\textipa{ku} $+$stem$+$ \textipa{a} & \textipa{ku} $+$stem$+$ \textipa{ana} & \textipa{ku} $+$stem$+$ \textipa{ila} & \textipa{ku} $+$stem$+$ \textipa{ilana} & \textipa{kut\'{u}} $+$stem$+$ \textipa{a} & \textipa{kuk\'{i}} $+$stem$+$ \textipa{a} & \textipa{kut\'{u}} $+$stem$+$ \textipa{ila} & \textipa{kuk\'{i}t\'{u}} $+$stem$+$ \textipa{ila} & UR
\\ \midrule
\textipa{kubala} & \textipa{kubalana} & \textipa{kubalila} & \textipa{kubalilana} & \textipa{kut\'{u}b\'{a}la} & \textipa{kuk\'{i}b\'{a}la} & \textipa{kut\'{u}b\'{a}lila} & \textipa{kuk\'{i}t\'{u}balila} & \textipa{bal}\\
\textipa{kugaja} & \textipa{kugajana} & \textipa{kugajila} & \textipa{kugajilana} & \textipa{kut\'{u}g\'{a}ja} & \textipa{kuk\'{i}g\'{a}ja} & \textipa{kut\'{u}g\'{a}jila} & \textipa{kuk\'{i}t\'{u}gajila} & \textipa{gaj}\\
\textipa{kugula} & \textipa{kugulana} & \textipa{kugulila} & \textipa{kugulilana} & \textipa{kut\'{u}g\'{u}la} & \textipa{kuk\'{i}g\'{u}la} & \textipa{kut\'{u}g\'{u}lila} & \textipa{kuk\'{i}t\'{u}gulila} & \textipa{gul}\\
\textipa{kub\'{a}la} & \textipa{kub\'{a}l\'{a}na} & \textipa{kub\'{a}l\'{i}la} & \textipa{kub\'{a}l\'{i}lana} & \textipa{kut\'{u}b\'{a}la} & \textipa{kuk\'{i}b\'{a}la} & \textipa{kut\'{u}b\'{a}lila} & \textipa{kuk\'{i}t\'{u}balila} & \textipa{b\'{a}l}\\
\textipa{kul\'{u}ma} & \textipa{kul\'{u}m\'{a}na} & \textipa{kul\'{u}m\'{i}la} & \textipa{kul\'{u}m\'{i}lana} & \textipa{kut\'{u}l\'{u}ma} & \textipa{kuk\'{i}l\'{u}ma} & \textipa{kut\'{u}l\'{u}mila} & \textipa{kuk\'{i}t\'{u}lumila} &  -- \\
\textipa{kus\'{u}na} & \textipa{kus\'{u}n\'{a}na} & \textipa{kus\'{u}n\'{i}la} & \textipa{kus\'{u}n\'{i}lana} & \textipa{kut\'{u}s\'{u}na} & \textipa{kuk\'{i}s\'{u}na} & \textipa{kut\'{u}s\'{u}nila} & \textipa{kuk\'{i}t\'{u}sunila} &  -- \\
\textipa{kul\'{a}ba} & \textipa{kul\'{a}b\'{a}na} & \textipa{kul\'{a}b\'{i}la} & \textipa{kul\'{a}b\'{i}lana} & \textipa{kut\'{u}l\'{a}ba} & \textipa{kuk\'{i}l\'{a}ba} & \textipa{kut\'{u}l\'{a}bila} & \textipa{kuk\'{i}t\'{u}labila} &  -- \\
\bottomrule\end{longtable}
\begin{tabular}{l}\emph{Rules: }\\
\verb|[ ]| $\to$ \verb|[-highTone]| / \verb|[+highTone]| \verb|[ ]| \verb|_| \\\verb|[ ]| $\to$ \verb|[+highTone]| / \verb|[+highTone]| \verb|[ ]| \verb|_| \verb|[-vowel]| \verb|[+vowel]|
\end{tabular}

\pagebreak

\emph{Polish:}\\\begin{longtable}{ll|l}\toprule
$\varnothing$ $+$stem$+$ $\varnothing$ & $\varnothing$ $+$stem$+$ \textipa{i} & UR
\\ \midrule
\textipa{klup} & \textipa{klubi} & \textipa{klub}\\
\textipa{trup} & \textipa{trupi} & \textipa{trup}\\
\textipa{dom} & \textipa{domi} & \textipa{dom}\\
\textipa{snop} & \textipa{snopi} & \textipa{snp}\\
\textipa{Zwup} & \textipa{Zwobi} & \textipa{Zwob}\\
\textipa{trut} & \textipa{trudi} & \textipa{trud}\\
\textipa{dzvon} & \textipa{dzvoni} & \textipa{dzvon}\\
\textipa{kot} & \textipa{koti} & \textipa{kt}\\
\textipa{lut} & \textipa{lodi} & \textipa{lod}\\
\textipa{grus} & \textipa{gruzi} & \textipa{gruz}\\
\textipa{nos} & \textipa{nosi} & \textipa{ns}\\
\textipa{vus} & \textipa{vozi} & \textipa{voz}\\
\textipa{wuk} & \textipa{wugi} & \textipa{wug}\\
\textipa{wuk} & \textipa{wuki} & \textipa{wuk}\\
\textipa{sok} & \textipa{soki} & \textipa{sk}\\
\textipa{ruk} & \textipa{rogi} & \textipa{rog}\\
\textipa{bur} & \textipa{bori} & \textipa{bor}\\
\textipa{vuw} & \textipa{vowi} &  -- \\
\textipa{sul} & \textipa{soli} &  -- \\
\textipa{buj} & \textipa{boji} &  -- \\
\textipa{Sum} & \textipa{Sumi} &  -- \\
\textipa{Zur} & \textipa{Zuri} &  -- \\
\bottomrule\end{longtable}
\begin{tabular}{l}\emph{Rules: }\\
\verb|[-sonorant]| $\to$ \verb|[-voice]| /  \verb|_| \#\\\verb|[+vowel]| $\to$ \textipa{u} /  \verb|_| \verb|[-nasal]| \#\\$\varnothing$ $\to$ \textipa{o} / \verb|[-vowel]| \verb|_| \verb|[-voice]|
\end{tabular}

\pagebreak

\emph{Ancient Greek:}\\\begin{longtable}{llll|l}\toprule
$\varnothing$ $+$stem$+$ \textipa{s} & $\varnothing$ $+$stem$+$ \textipa{os} & $\varnothing$ $+$stem$+$ \textipa{i} & $\varnothing$ $+$stem$+$ \textipa{si} & UR
\\ \midrule
\textipa{hals} & \textipa{halos} & \textipa{hali} & \textipa{halsi} & \textipa{hal}\\
\textipa{ojs} & \textipa{ojos} & \textipa{oji} & \textipa{ojsi} & \textipa{oj}\\
\textipa{sus} & \textipa{suos} & \textipa{sui} & \textipa{susi} & \textipa{su}\\
\textipa{klo:ps} & \textipa{klo:pos} & \textipa{klo:pi} & \textipa{klo:psi} & \textipa{klo:p}\\
\textipa{p\super hle:ps} & \textipa{p\super hle:bos} & \textipa{p\super hle:bi} & \textipa{p\super hle:psi} & \textipa{p\super hle:b}\\
\textipa{kate:lips} & \textipa{kate:lip\super hos} & \textipa{kate:lip\super hi} & \textipa{kate:lipsi} & \textipa{kate:lip\super h}\\
\textipa{p\super hulaks} & \textipa{p\super hulakos} & \textipa{p\super hulaki} & \textipa{p\super hulaksi} & \textipa{p\super hulak}\\
\textipa{ajks} & \textipa{ajgos} & \textipa{ajgi} & \textipa{ajksi} & \textipa{ajg}\\
\textipa{salpiNks} & \textipa{salpiNgos} & \textipa{salpiNgi} & \textipa{salpiNksi} & \textipa{salpiNg}\\
\textipa{onuks} & \textipa{onuk\super hos} & \textipa{onuk\super hi} & \textipa{onuksi} & \textipa{onuk\super h}\\
\textipa{t\super he:s} & \textipa{t\super he:tos} & \textipa{t\super he:ti} & \textipa{t\super he:si} & \textipa{t\super he:t}\\
\textipa{k\super haris} & \textipa{k\super haritos} & \textipa{k\super hariti} & \textipa{k\super harisi} & \textipa{k\super harit}\\
\textipa{elpis} & \textipa{elpidos} & \textipa{elpidi} & \textipa{elpisi} & \textipa{elpid}\\
\textipa{korus} & \textipa{korut\super hos} & \textipa{korut\super hi} & \textipa{korusi} & \textipa{korut\super h}\\
\textipa{ri:s} & \textipa{ri:nos} & \textipa{ri:ni} & \textipa{ri:si} & \textipa{ri:n}\\
\textipa{delp\super hi:s} & \textipa{delp\super hi:nos} & \textipa{delp\super hi:ni} & \textipa{delp\super hi:si} & \textipa{delp\super hi:n}\\
\bottomrule\end{longtable}
\begin{tabular}{l}\emph{Rules: }\\
\verb|[-lateral +coronal]| $\to$ $\varnothing$ /  \verb|_| \verb|[-vowel]|\\\verb|[-sonorant]| $\to$ \verb|[-aspirated -voice]| /  \verb|_| \verb|[-voice]|
\end{tabular}

\pagebreak

\emph{Catalan:}\\\begin{longtable}{ll|l}\toprule
$\varnothing$ $+$stem$+$ $\varnothing$ & $\varnothing$ $+$stem$+$ \textipa{@} & UR
\\ \midrule
\textipa{@kelj} & \textipa{@kelj@} & \textipa{@kelj}\\
\textipa{mal} & \textipa{mal@} & \textipa{mal}\\
\textipa{siBil} & \textipa{siBil@} & \textipa{siBil}\\
\textipa{@skerp} & \textipa{@skerp@} & \textipa{@skerp}\\
\textipa{Sop} & \textipa{Sop@} & \textipa{Sop}\\
\textipa{sEk} & \textipa{sEk@} & \textipa{sEk}\\
\textipa{@spEs} & \textipa{@spEs@} & \textipa{@spEs}\\
\textipa{gros} & \textipa{gros@} & \textipa{gros}\\
\textipa{baS} & \textipa{baS@} & \textipa{baS}\\
\textipa{koS} & \textipa{koS@} & \textipa{koS}\\
\textipa{tot} & \textipa{tot@} & \textipa{tot}\\
\textipa{brut} & \textipa{brut@} & \textipa{brut}\\
\textipa{pOk} & \textipa{pOk@} & \textipa{pOk}\\
\textipa{pr@sis} & \textipa{pr@siz@} & \textipa{pr@siz}\\
\textipa{fr@nses} & \textipa{fr@nsez@} & \textipa{fr@nsez}\\
\textipa{gris} & \textipa{griz@} & \textipa{griz}\\
\textipa{k@zat} & \textipa{k@zaD@} & \textipa{k@zad}\\
\textipa{bwit} & \textipa{bwiD@} & \textipa{bwid}\\
\textipa{rOt\super S} & \textipa{rOZ@} & \textipa{rOd\super Z}\\
\textipa{bot\super S} & \textipa{boZ@} & \textipa{bod\super Z}\\
\textipa{orp} & \textipa{orB@} & \textipa{orb}\\
\textipa{ljark} & \textipa{ljarG@} & \textipa{ljarg}\\
\textipa{sek} & \textipa{seG@} & \textipa{seg}\\
\textipa{f@Suk} & \textipa{f@SuG@} & \textipa{f@Sug}\\
\textipa{grok} & \textipa{groG@} & \textipa{grog}\\
\textipa{puruk} & \textipa{puruG@} & \textipa{purug}\\
\textipa{kandit} & \textipa{kandiD@} & \textipa{kandid}\\
\textipa{frEt} & \textipa{frED@} & \textipa{frEd}\\
\textipa{s@Gu} & \textipa{s@Gur@} & \textipa{s@Gur}\\
\textipa{du} & \textipa{dur@} & \textipa{dur}\\
\textipa{s@G@Do} & \textipa{s@G@Dor@} & \textipa{s@g@Dor}\\
\textipa{kla} & \textipa{klar@} & \textipa{klar}\\
\textipa{nu} & \textipa{nu@} & \textipa{nu}\\
\textipa{kru} & \textipa{kru@} & \textipa{kru}\\
\textipa{flO\~nd\super Zu} & \textipa{flO\~nd\super Z@} & \textipa{flO\~nd\super Zu}\\
\textipa{dropu} & \textipa{drop@} & \textipa{dropu}\\
\textipa{@gzakt@} & \textipa{@gzakt@} & \textipa{@gzakt@}\\
\textipa{@lBi} & \textipa{@lBin@} & \textipa{@lBin}\\
\textipa{sa} & \textipa{san@} & \textipa{san}\\
\textipa{pla} & \textipa{plan@} & \textipa{plan}\\
\textipa{bo} & \textipa{bon@} & \textipa{bon}\\
\textipa{s@rE} & \textipa{s@rEn@} & \textipa{s@rEn}\\
\textipa{suBlim} & \textipa{suBlim@} & \textipa{suBlim}\\
\textipa{al} & \textipa{alt@} & \textipa{alt}\\
\textipa{fOr} & \textipa{fOrt@} & \textipa{fOrt}\\
\textipa{kur} & \textipa{kurt@} & \textipa{kurt}\\
\textipa{sor} & \textipa{sorD@} & \textipa{sord}\\
\textipa{bEr} & \textipa{bErD@} & \textipa{bErd}\\
\textipa{san} & \textipa{sant@} & \textipa{sant}\\
\textipa{k@lEn} & \textipa{k@lEnt@} & \textipa{k@lEnt}\\
\textipa{prufun} & \textipa{prufund@} & \textipa{prufund}\\
\textipa{f@kun} & \textipa{f@kund@} & \textipa{f@kund}\\
\textipa{d@sen} & \textipa{d@sent@} & \textipa{d@sent}\\
\textipa{dulen} & \textipa{dulent@} & \textipa{dulent}\\
\textipa{@stuDian} & \textipa{@stuDiant@} & \textipa{@stuDEiant}\\
\textipa{blaN} & \textipa{blaNk@} & \textipa{blaN\~n}\\
\bottomrule\end{longtable}
\begin{tabular}{l}\emph{Rules: }\\
\verb|[-lateral +sonorant +coronal]| $\to$ $\varnothing$ /  \verb|_| \#\\\verb|[-sonorant]| $\to$ \verb|[-voice]| /  \verb|_| \#\\\verb|[-vowel]| $\to$ \textipa{k} / \textipa{N} \verb|_| \\\verb|[-nasal +voice]| $\to$ \verb|[+continuant]| / \verb|[-nasal]| \verb|_| \verb|[+vowel]|\\\verb|[+vowel]| $\to$ $\varnothing$ / \verb|[-sonorant]| \verb|_| \verb|[+vowel]|\\\textipa{t} $\to$ $\varnothing$ / \verb|[-vowel]| \verb|_| \#
\end{tabular}

\end{document}

### 2018-02-20 topics.tex
initial commit

--- 2018-02-20 outline.tex
@@
 \usepackage{amssymb}
+\usepackage{dsfont}
 \usepackage{tikz}
@@
 \usepackage{hyperref}
+\usepackage{amsfonts}       % blackboard math symbols
 
@@
 
-\title{Discovering the Rules of language using Bayesian Program Learning}
+\title{Automatically generating alternative hypotheses and controls}
 %\author{Kevin Ellis, Timothy O'Donnell}
@@
 \DeclareMathOperator*{\argmin}{arg\,min}
+\newcommand{\Expect}{\mathds{E}} %{{\rm I\kern-.3em E}}
+\newcommand{\indicator}{\mathds{1}} %{{\rm I\kern-.3em E}}
+\newcommand{\expect}{\mathds{E}} %{{\rm I\kern-.3em E}}
+\newcommand{\probability}{\mathds{P}} %{{\rm I\kern-.3em P}}
 
@@
 
-%% How do linguists come up with phonological rules, how do kids learn
-%% artificial grammars, and how does one acquire pig latin?  The
-%% solutions to these problems share a common representation, which we
-%% show can be modeled as a program, and the corresponding learning
-%% problems modeled as program induction.  This framing lets us apply
-%% ideas from Bayesian Program Learning to induce grammars, which
-%% combines program synthesis techniques with a compression-based inductive
-%% bias. This lets the models capture phonological phenomena like vowel
-%% harmony or stress patterns and learn synthetic grammars used in prior
-%% studies of artificial grammar learning.  Going beyond individual
-%% grammar learning problems, we consider the problem of jointly
-%% inferring many related rule systems. By solving many textbook
-%% phonology problems, we can ask the model what kind of inductive bias
-%% best explains the attested phenomena.
+\section{Automatically generating grammars with Bayesian program learning}
 
-\begin{abstract}
-    %% How do linguists come up with phonological rules? How do kids
-    %% learn pig latin?  We develop computational models of these and
-    %% other phenomena in language.  The unifying element of the models
-    %% is to treat grammars as programs, which lets us apply ideas from
-    %% the field of program synthesis to learn grammars. This lets the
-    %% models capture phonological phenomena like vowel harmony or stress
-    %% patterns and learn synthetic grammars used in prior studies of
-    %% rule learning.  Going beyond individual grammar learning problems,
-    %% we consider the problem of jointly inferring many related rule
-    %% systems. By solving many textbook phonology problems, we can
-    %% ask the model what kind of inductive bias best explains the
-  %% attested phenomena.
-Both children and linguists confront a similar problem of inference:
-given utterances produced by speakers, together with aspects of the
-meaning of those utterances, discover the grammatical principles that
-relate form to meaning. We study this abstract computational
-problem within the domain of morphophonology, contributing a computational model
-that learns phenomena from many natural languages and generalizes in
-humanlike ways from data used in behavioral studies of artificial grammar
-learning.
+\subsection{A Probabilistic Framing}
 
-Our work draws on two analogies. The \emph{child-as-linguist analogy}
-holds that both children and linguists must solve the same abstract
-inductive reasoning problem, even though the nature of the input data and underlying mental algorithms
-are surely different in precise detail. Accordingly we isolate the
-problem of learning morphophonological systems, and show that a single
-solution to this problem can capture both linguistic analyses from
-natural languages and infant rule learning of artificial languages.
-We adopt the framework of ``Bayesian Program learning'' (BPL) -- in
-which learning is formulated a synthesizing a program which compactly
-describes the input data. This \emph{learning-as-programming analogy}
-lets us exploit recent techniques from the field of program synthesis to
-induce morphophonological rules from data.
-While \emph{child-as-linguist} poses the
-computational problem, \emph{learning-as-programming} offers a
-solution.
-  \end{abstract}
+The goal of the grammar learner, whether it is a child, linguist, or computer program, is to
+infer the grammar that gave rise to
+a collection of utterances.
+One way of
+modeling this inference problem is to (1) place a prior distribution over grammars, for example, a distribution that puts more probability mass on shorter or simpler grammars; (2) equip each grammar with a
+\emph{likelihood model} that specifies exactly how likely a grammar is to produce a given utterance;
+and then (3) use Bayes's rule to work backwards from the
+utterances to the grammar that was likely to produce them:
+\begin{equation}
+  \probability\left[\text{grammar}|\text{utterances} \right]\propto\underbrace{\probability\left[\text{utterances}|\text{grammar} \right]}_{\text{Likelihood}}\times\underbrace{\probability\left[\text{grammar} \right]}_{\text{Prior}}\label{posterior}
+\end{equation}
+We can think of the above equation as a recipe for
+scoring how well a grammar explains a
+collection of utterances.
+Once we have a prior and a likelihood,
+we know how to evaluate how well a grammar explains the data, modulo the assumptions implicit in the
+prior and likelihood.
 
-\section{Introduction}
+A simple and intuitive prior over grammars is one which penalizes longer grammars and favors parsimonious grammars, for example defining $\probability\left[\text{grammar} \right]\propto\exp\left(- \text{\# symbols in grammar} \right)$.
+An example of a likelihood model,
+and the one we use here,
+is just to count the number of extra bits or symbols needed to encode an utterance given the grammar\footnote{This is equivalent to a minimum description length (MDL) interpretation}.
 
-Both children and linguists confront a similar problem of inference:
-given utterances produced by speakers, together with aspects of the
-meaning of those utterances, discover the grammatical principles that
-relate form to meaning. Here, we study this abstract computational
-problem within the domain of morphophonology, the part of language
-that builds up words from simpler parts such as stems, prefixes, and
-suffixes (morphology) and then specifies how the words are realized as
-sound patterns (phonology).  Our contribution is a computational model
-that learns phenomena from 26 natural languages and generalizes in
-humanlike ways from data used in infant studies of artificial grammar
-learning.
+Now in theory this probabilistic framing
+has now bought us a procedure for automatically discovering
+grammars that do a good job explaining the data.
+In practice,
+the space of all grammars is infinite and combinatorial,
+and so cashing out this framing requires a
+efficient procedure for searching the space of grammars and honing in on those that maximize Equation~\ref{posterior}.
+A subfield of computer science, called \textbf{program synthesis},
+has developed efficient techniques for
+solving seemingly intractable combinatorial search problems
+like those that arise in grammar learning.
+We use a program synthesis tool called Sketch~\cite{solar2008program}
+to search for grammars maximizing~\ref{posterior},
+which was previously applied to language learning problems in~\cite{ellis2015unsupervised}.
+For more detail on these program synthesis techniques we refer the reader to~\cite{solar2008program}.
+This cashing out of the framing makes
+our approach an instance of  \textbf{Bayesian Program Learning (BPL)}; see~\cite{lake2015human}.
 
-Our work draws on two analogies. The \emph{child-as-linguist analogy}
-holds that both children and linguists must solve the same abstract
-inductive reasoning problem, even though the nature of the input data,
-the process of theory construction, and underlying mental algorithms
-are surely different in precise detail. Accordingly we isolate the
-problem of learning morphophonological systems, and show that a single
-solution to this problem can capture both linguistic analyses from
-natural languages and infant rule learning of artificial languages.
-
-We adopt the framework of ``Bayesian Program learning'' (BPL) -- in
-which learning is formulated a synthesizing a program which compactly
-describes the input data. This \emph{learning-as-programming analogy}
-explains the flexibility and richness of a grammar
-as a consequence of 
-While \emph{child-as-linguist} poses the
-computational problem, \emph{learning-as-programming} offers a
-solution.
-
-We develop a programming language tailored to expressing
-morphophonological generalizations.
-For children this programming language is innate and is
-commonly called ``Universal Grammar'' (UG). Linguists instead must
-uncover this programming language by looking at many languages and
-searching for commonalities. Learning what makes up UG unfolded over
-evolutionary time for children and is ongoing within linguistics.  BPL
-reinterprets the problem of discovering UG.  In BPL, UG is the
-programming language combined with an inductive bias over programs.
-Here we estimate this inductive bias empirically from naturally
-occurring morphophonological systems as a way of learning to learn
-grammars.
-
-\section{Morphophonological phenomena}
-
-What is phonology supposed to explain? Familiar phenomena from
-English help motivate phonological grammars.  The plural of \emph{cat}
-is pronounced \textipa{k\ae ts} (``cats''), while the plural of
-\emph{dog} is pronounced \textipa{dagz} (``dogs''). A phonological
-rule accounts for the alternation between \textipa{s} and \textipa{z}:
-\textipa{z} becomes \textipa{s} at the end of words whenever the
-preceding sound is unvoiced (does not involve vibrating the vocal
-cords). This generalization holds throughout the entirety of English.
-The count system of Tibetan presents another puzzle.
-``1'' in Tibetan is pronounced
-\textipa{\|x{j}ig}, ``10'' is \textipa{\|x{j}u},
-yet ``11'' ($ = 10 + 1$) is \textipa{\|x{j}ug\|x{j}ig},
-rather than \textipa{\|x{j}u\|x{j}ig}.
-The explanation lies in Tibetan phonology,
-which deletes word initial consonants followed by another consonant (Table~\ref{Tibetan}).
-In this analysis of Tibetan,
-``1'' has a latent ``underlying'' representation as \textipa{g\|x{j}ig}
-but is pronounced as \textipa{\|x{j}ig}.
-
-The aim of phonology is to explain pronunciations of words in terms of
-latent underlying representations (eg, \textipa{g\|x{j}ig} for Tibetan
-``1'') and the rules that give observed surface pronunciations from
-underlying forms (eg, deleting word initial consonants).
-Context-sensitive rewrite systems are a Turing complete representation
-that also naturally express phonological
-rules~\cite{chomsky1968sound}.  We use context-sensitive rewrites as a
-representation for programs that implement a language's phonology.
-For example, the rewrite in Table~\ref{Tibetan},
-\verb|C|$\to\varnothing /$\verb|#_C|, deletes ($\to\varnothing $)
-consonants (\verb|C|) at word boundaries (\verb|#|) when followed by
-another consonant (\verb|C|).
-Thus phonological rule learners, whether they be children, linguists, or machines, must solve a program induction problem.
-
-\subsection{Inducing Programs}
-We model the program learning problem as Bayesian inference, treating the pronunciations of words as observed data and the phonology as a latent variable whose values are programs.
-So the model's beliefs about possible programs are:
-\begin{equation}
-  P(\text{program}|\text{words})\propto P(\text{program}|\text{UG})\prod_{w\in\text{words}}\sum_{u\in\text{underlying forms}} P(u)P(w|u,\text{program})
-  \end{equation}
-Here, Universal Grammar (UG) is an inductive bias over programs.
-This inductive bias encodes knowledge of the program space (what rules are possible in human language and what rules are impossible), imparting a bias towards more naturally structured rules, like rules written in terms of syllables and consonants.
-
-\subsection{The role and importance of universal grammar}
-                We claim any plausible theory of linguistic rule learning needs a universal grammar.
-Studies of infant grammar learning suggests that babies come equipped with prior knowledge of what language should look like.
-This prior knowledge constrains the space of allowed generalizations, accounting for the astonishing pace at which children acquire language.
-This line of reasoning is a ``poverty-of-the-stimulus'' style argument: if children can induce so much grammar from such meager data, something else must bridge the gap.
-Because universal grammar also constrains the space of allowed grammars, we would expect that humans do not learn certain generalizations, even when the generalizations seem to hold in the data.
-Indeed this is the case: this ``surfeit-of-the-stimulus'' style argument  explains phenomena in natural languages (for example, statistical regularities that seem to hold in Turkish are not actually used by speakers) and in infant artificial language learning (for example, infants acquire certain phonological generalizations from just a few examples, while never acquiring other generalizations that are seemingly just a simple).
-While previous accounts of universal grammar were abstract, our model is concrete and computational.
-We define universal grammar as a probability mass function that puts zero mass on impossible or unattested generalizations, places more mass on programs licensing the generalizations that people learn quickly and naturally, and then smoothly falls off in probability mass as the programs become less natural.
-
-\subsection{Learning natural language phonology}                
-Our model learns a diverse set of phonological generalizations found in natural languages.
-We compiled a multilingual data set of words from standard linguistic textbooks.
-In these data sets, our model recovers allophonic relationships (e.g., l vs.
-r in Japanese) and systems that form different inflections of words (e.g., English verb inflections, or the Tibetan count system).
-Systems like these have received intense study within computational linguistics.
-But no prior model could explain the diversity of phonological phenomena found even in a textbook.
-To explain many diverse phenomena we need a rich, flexible representation -- programs -- coupled to a linguisticly informed inductive bias -- universal grammar.
-Our model learns simple generalizations like English verbal inflections, as well as phenomena like vowel harmony (in the language Kikuria, the word tiga becomes tegera in a different inflection, or ruga becomes rogera), or tonal rules in languages with tone (in the language Kerewe, different inflections cause tones to appear, disappear, or move to different parts of a word).
-
-For these results, the learning-as-programming analogy does the work for us.
-BPL moves beyond the impoverished representations of prior learners (neural networks as in, or rigid classes of symbolic rules as in) instead represented generalizations as programs.
-We claim that programs are a sufficiently general representation to support the kinds of rules found in phonology.
-
-\begin{table}\centering
-\begin{tabular*}{10cm}{lcl}
-  \toprule
-  Example data&Phonology\\\midrule
-\begin{tabular}{ll}
-1&\textipa{\|x{j}ig}\\
-4&\textipa{\|x{s}i}\\
-5&\textipa{Na}\\
-9&\textipa{gu}\\
-10&\textipa{\|x{j}u}\\
-11 ($= 10 + 1$) & \textipa{\|x{j}ug\|x{j}ig}\\
-14 ($= 10 + 4$) & \textipa{\|x{j}ub\|x{s}i}\\
-15 ($= 10 + 5$) & \textipa{\|x{j}uNa}\\
-19 ($= 10 + 9$) & \textipa{\|x{j}urgu}\\
-40 ($= 4 + 10$) & \textipa{\|x{s}ib\|x{j}u}\\
-50 ($= 5 + 10$) & \textipa{Nab\|x{j}u}\\
-90 ($= 9 + 10$) & \textipa{gub\|x{j}u}
-\end{tabular}
-&
-\begin{tabular}{l}
-  \emph{Consonant cluster reduction:}\\
-  \verb|C|$\to\varnothing $ $/$ \verb|# _ C|\\
-  \end{tabular}
-
-  \\
-  \bottomrule  \end{tabular*}
-\caption{Tibetan count system.}
-\label{Tibetan}
-  \end{table}
-
-\pagebreak
-
-%% 2 insights: analogy between children and linguists. analogy between
-%% grammars and programs.
-
-%% The core problem we're pushing on is \textbf{structure learning in
-%%   language}. It makes sense that you could learn parameters and
-%% lexical items, but could you learn something with more the structure
-%% of a grammar? Here we answer that question in the affirmative within a
-%% restricted domain (phonology) on restricted problems (smaller data
-%% sets and simpler sets of rules).
-
-%% Why focus on some problems that feel more like computational linguistics and other problems that feel more like developmental psychology?
-Connection to child-as-linguist~\cite{chomsky1988current}, child-as-scientist, child-as-hacker\cite{Rule2015RepresentingAL}.
-
-%% What is the connection between what children do and what linguists do? From the point of view of this research project,
-%% the difference lies in the nature and role of universal grammar.
-%% For a linguist, universal grammar is an unknown to be discovered.
-%% When analyzing a new data set, a linguist might need to entertain a broader or more expressive universal grammar.
-%% For the child, universal grammar is an innate part of our genetic endowment.
-
-\section{The Model}
-
-\begin{figure}[h!]\centering
-  \includegraphics[width = 0.9\textwidth]{generativeModel.pdf}
-  \caption{{\color{red}Red}: Both the morphology $\mathcal{M}$ and
-    phonology $\mathcal{P}$ are programs either build through
-    concatenation or function composition
-    (Tables~\ref{morphologyGrammar} \&~\ref{phonologyGrammar}).
-These programs explain how words are built in both natural and artificial grammars ({\color{blue}blue}).
-Learners observe pronunciations of words ({\color{magenta}}) in order to recover these programs, but with the aid of a strong inductive bias (Universal Grammar, {\color{purple}purple})}.
-  \label{beautiful}
-\end{figure}
-
-\section{Artificial Grammar Learning}
+\subsection{Modeling solutions to AGL problems}
 
@@
 instead studying AGL in controlled experiments.
-
-Our contribution to AGL research is the demonstration that
-a single BPL model can explain (1) that infants quickly acquire
-certain simple grammars very easily, and at the same time (2)
-those same infants go on to learn
-the very complicated grammar of their first language.
+Our contribution here is the demonstration that
+a single BPL model can explain how a learner can quickly infer a grammar represented as SPE--style rules.
 We collected grammars from AGL datasets~\cite{gerken2010infants,marcus1999rule,frank2011three}
 and presented our BPL learner with data drawn from these grammars (Table~\ref{artificialGrammarTable}),
-finding that our model recovers the generative processes that
- underlies these data sets.
+finding that our model automatically discovers the grammars that
+ underly these data sets.
 \begin{table}[h!]
@@
 
+\subsection{Automatically exploring the space of alternative hypotheses}
+
 Just like in first language acquisition,
 these AGL learning setups introduce a trade-off between
-grammars that are likely and therefore small (``parsimony'')  and grammars that
-best fit the data.
+grammars that are likely under the prior (and are therefore small; ``parsimony'')  and grammars that
+assign high likelihood to the data (and therefore closely ``fit the data'', having high likelihood).
 For example, a learner could infer a grammar that just memorized to the data (perfect fit but poor parsimony)
@@
   \caption{Pareto fronts for ABB  (Marcus~1999; top two) \& AAX  (Gerken~2006; bottom two) learning problems fo either one example word (left column) or three example words  (right column).
+Rightward on x-axis corresponds to more parsimonious grammars and upward on y-axis corresponds to grammars that best fit the data, so the best grammars live in the upper right corners of the graphs.
 {\color{red}Red shade}: ground truth grammar.
@@
 
-A successful model of
-artificial grammar learning should also explain at least some aspects of first language acquisition.
-As a first step in this direction,
-we take the same model that we applied to
-these AGL learning problems and use it to solve phonology textbook exercises.
-Each of these textbook exercises is an inductive reasoning
-problem where the learner must
-explain surface pronunciations
-in terms of a latent generative model (a program).
-Figure~\ref{naturalLanguageExample}
-shows our model solving a
-representative textbook problem.
+A successful model of artificial grammar learning should also explain
+at least some aspects of first language acquisition.  As a first step
+in this direction, we take the same model that we applied to these AGL
+learning problems and use it to solve phonology textbook exercises.
+Each of these textbook exercises is an inductive reasoning problem
+where the learner must explain surface pronunciations in terms of a
+latent generative model (a grammar).
+Figure~\ref{naturalLanguageExample} shows our model solving a
+representative textbook problem.  We believe this extension two
+natural language is an important next step for models of artificial
+grammar learning,
+and helps deliver on the basic promises and presuppositions of AGL research.
 
-\textbf{TIM: Should I talk about UG? We needed it for Tibetan. We probably don't have space.}
-
-\begin{figure}\centering
+\begin{figure}[H]\centering
   \begin{minipage}{7cm}\centering
@@
   \end{minipage}
-  \caption{Example textbook phonology exercise (left) and the solution our model discovers for it (right).
+  \caption{Example textbook phonology exercise (left, taken from~\cite{9780511808869}) and the solution our model discovers for it (right).
 To solve this problem the model must jointly infer unobserved underlying representations for each of the Tibetan numbers, along with a phonological rule that explains their surface pronunciations.
@@
 
-\pagebreak
-\section{Appendix}
-
-\begin{table*}[h]
-\centering
-\begin{tabular*}{\textwidth}{ll}
-\toprule
-Grammar rule  & English description \\\midrule
-$\mathcal{P}\to \mathcal{R}\circ \mathcal{R}\circ \cdots \circ \mathcal{R}$&Phonology $\mathcal{P}$ is a sequence of compositions of rewrites $\mathcal{R}$\\
-$\mathcal{R}\to \mathcal{F}\longrightarrow \mathcal{C} / \mathcal{T}\_\mathcal{T}$& Rewrite focus $\mathcal{F}$ to structural change $\mathcal{C}$ between triggers $\mathcal{T}$\\
-$\mathcal{T}\to \# \mathcal{T}' | \mathcal{T}'$ & Triggers optionally match end of string\\
-$\mathcal{T}'\to \epsilon | \mathcal{X} \mathcal{T}' | \mathcal{X}^* \mathcal{T}'$ & Triggers are sequences of matrices $\mathcal{X}$ possibly with Kleene star\\
-$\mathcal{X}\to a|t|s|\cdots$&Matrices can be constant phonemes\\
-$\mathcal{X}\to [\pm \mathcal{G} \; \pm \mathcal{G} \; \cdots \; \pm \mathcal{G}]$& Matrices check features $\mathcal{G}$\\
-$\mathcal{X}\to \sigma$&Matrices can be whole syllables\\
-$\mathcal{G}\to \text{voice}|\text{nasal}|\cdots$&Standard phonological features\\
-$\mathcal{F}\to \mathcal{X}$&Focus can be a feature matrix\\
-$\mathcal{F}\to \varnothing$&Focus can be empty (insertion rule)\\
-$\mathcal{F}\to\mathbb{Z}$&Focus can be one of the triggers (copies it)\\
-$\mathcal{C}\to\mathcal{X}$&Structural change can be a feature matrix\\
-$\mathcal{C}\to\varnothing$&Deletion rule\\
-\bottomrule
-\end{tabular*}
-\caption{Grammar over phonologies. The non-terminal $\mathcal{P}$ is the start symbol for the grammar.}
-\label{phonologyGrammar}
-\end{table*}
-
-\begin{table*}[h]
-\centering
-\begin{tabular*}{\textwidth}{ll}
-\toprule
-Grammar rule  & English description \\\midrule
-$\mathcal{W}\to (\mathcal{M}) \mathcal{M} (\mathcal{M})$&Morphology builds words $\mathcal{W}$ by concatenating morphemes $\mathcal{M}$. Prefix/Suffix optional (in parentheses)\\
-$\mathcal{M}\to$ sequence of phonemes\bottomrule
-\end{tabular*}
-\caption{Grammar over morphologies. The non-terminal $\mathcal{W}$ is the start symbol for the grammar.}
-\label{morphologyGrammar}
-\end{table*}
-
-\pagebreak
-\subsection{Inference}
-
-The solver takes as input:
-\begin{itemize}
-\item A set of inflected surface forms, $\{x_n\}$.
-Here the variable $x_n$ refers to all of the inflections of a word.
-So an example set of 3 inflected surface forms might be $\{(\text{fish},\text{fished}), (\text{walk},\text{walked}), (\text{eat},\text{ate})\}$.
-\item A bound on the number of rules to consider synthesizing, $B_R$.
-\end{itemize}
-It produces as output:
-\begin{itemize}
-\item A set of underlying forms, $\{u_n\}$.
-\item A morphology consisting of a prefix ($\{p_i\}$) and a suffix ($\{s_i\}$)  for each inflection
-\item  A sequence of rules, $\{r_j\}_{j = 1}^{B_R}$
-\end{itemize}
-It is \emph{guaranteed} to discover the rules, morphology, and underlying forms minimizing:
-\begin{equation}
-  \text{solver}(\{x_n\},B_R) = \argmin_{\{r_j\}_{j = 1}^{B_R}, \{u_n\}, \{p_i\}, \{s_i\}} \sum_j \text{cost}(r_j) + \sum_i (\text{cost}(p_i) + \text{cost}(s_i) ) + \sum_n \text{cost}(u_n)
-\end{equation}
-In exchange for this strong guarantee, however, it has only a very
-weak guarantee on the amount of time it will take to discover the
-solution: in the worst case, it will take time exponential size of the
-problem. In practice it takes close to a day for it to do exact search over 3 rules.
-For easy problems, like the AGL experiments, I can get away with 4 rules if I wait for a few hours.
-
-One can also ask the solver for the top $K$ solutions minimizing the cost function, which I will write:
-\begin{equation}
-  \text{solver}_K(\{x_n\},B_R) = \text{the }K\text{ minimum cost solutions.}
-\end{equation}
-One can also provide additional constraints to the solver, for example telling it what the underlying forms are,
-or telling it what some of the rules are. I will write these constraints by adding an extra argument to the solver
-that looks like conditioning in Bayesian inference. For example,
-\begin{equation}
-  \text{solver}_K(\{x_n\},B_R|u_1 = \text{fish}) = \text{the }K\text{ minimum cost solutions where }u_1\text{ has to be fish.}
-  \end{equation}
-
-So now the game is to leverage this solver as a ``black box'' inside of an incomplete search algorithm.
-A key subroutine in my attempts at building such a search algorithm is
-\texttt{solveIncrementalChange}, which takes as input an existing set of rules ($\{r_j^e\}_{j = 1}^{B_R}$)
-along with a \textbf{search radius} ($R$),
-and solves for a solution that changes at most \textbf{search radius} rules.
-\begin{equation}
-  \text{solveIncrementalChange}_K (\{x_n\}, \{r_j^e\}_{j = 1}^{B_R}, R)
-   = \text{solver}_K (\{x_n\}, B_R+1 | r_j\not= r_j^e \text{ at most }R\text{ times})
-\end{equation}
-So \texttt{solveIncrementalChange} can add a rule (see $B_R+1$).
-I didn't build in any explicit mechanism for reordering rules -- if you set the search radius $R$ to 2, then it could effectively swap 2 rules.
-
-In addition to synthesizing large programs, we also have to deal with
-large data sets. Armando already came up with a nice trick for this called CEGIS\footnote{counterexample
-  guided inductive synthesis}.
-The idea behind CEGIS is that you synthesize a program from one example,
-and then you check to see if that program is inconsistent with any other examples.
-If not then you are done.
-Otherwise, you have found a \textbf{counterexample}. You add the counterexample to your
-training data set and repeat.
-
-Combining \texttt{solveIncrementalChange} with CEGIS gives the
-skeleton of the inference algorithm that I have been experimenting
-with, called \texttt{incrementalCEGIS}. See the pseudocode at the top of the next page:
-\pagebreak
-\begin{algorithmic}
-  \STATE \textbf{\texttt{incrementalCEGIS}}:
-  \STATE {\bfseries Input:} Matrix of inflected forms for $N$ lexemes, $X$
-  \STATE {\bfseries Output:} underlying forms, $\{u_n\}$; prefixes $\{p_i\}$; suffixes $\{s_i\}$; rules $\{r_j\}$
-  \STATE // Initialize training set $T$ with a single example
-  \STATE Set $T = $ inflections for any lexeme in $X$
-  \STATE // Find a solution for this single example, using as many rules as you need ($\infty$)
-  \STATE Set $\{r_j\}_{j = 1}^{B_R}, \{p_i\}, \{s_i\} = \text{solver}(T,\infty)$
-  \REPEAT
-  \STATE // Find the set of counterexamples $C$ inconsistent with our current guess as to what the solution is
-  \STATE Set $C = \{x | x\in X \text{ and } x \text{ inconsistent with } \{r_j\}_{j = 1}^{B_R},\{p_i\}, \{s_i\}\}$
-  \STATE // If there are no counterexamples as we are done
-  \STATE \textbf{if} $C = \varnothing$ \textbf{ then return }$\{r_j\}_{j = 1}^{B_R},\{p_i\}, \{s_i\}$
-
-  \STATE // Start out with a search radius of $1$ and increase it until we find a way of accommodating another example
-  \FOR{$R = 1,2,3,4,...$}
-  \FOR {$c\in C$}
-  \STATE Set synthesisResult$ = \text{solveIncrementalChange}(T\cup\{c\},\{r_j\},R)$
-  \IF{synthesisResult$\not= \bot$}
-  \STATE Set $\{r_j\}_{j = 1}^{B_R}, \{p_i\}, \{s_i\} = $synthesisResult
-  \STATE Add $c$ to the set of training examples, $T$
-  \STATE \textbf{break out of loop over }$R$ // Successfully updated both our solution and added another example to the training set
-\end{algorithmic}
-\vspace{1cm}
-
-One important feature of this approach is that it does not reuse
-previous inferences about the morphology or the underlying forms.
-I think we want to add this.
-For example, we might allow it to change up to $R$ underlying forms,
-or disallow changes in the morphology once the training set is big enough.
-
-Variations of \textbf{incrementalCEGIS} that I have tried:
-\begin{itemize}
-\item[] Maintaining a population of solutions.
-Instead of taking the top 1 solution every time you synthesize, try taking the top $K$.
-There is some trade-off here between the cost of solving for more solutions and
-  the benefit of not needing as many counterexamples. I have not done any systematic experiments,
-  but my general sense is that this trade-off favored $K = 1$ in practice (so not maintaining a population of solutions).
-\item[] Letting the solver insert the new rule anywhere vs constraining it to either append or prepended the new rule.
-Sketch uses up a lot of memory whenever I let it insert the rule anywhere -- not sure why this is.
-\item[] Giving sketch more problems to solve but making each problem smaller.
-Here what I do is have an outer loop that iterates over all of the different ways you could edit an existing set of rules, and then invoke the solver for each of these.
-So for example I invoke the solver telling it to change only the first rule,
-  then I invoke it telling it to change only the second rule, etc, finally telling it to keep all the rules the same but add a new rule.
-\end{itemize}
-
-\section{Related work}
-
-Minimum generalization learner: \cite{Albright03rulesvs}.
-The interesting similarity here is that we both learned the structure of the rules.
-But it's restriction to the supervised case hampers its ability to really capture what a grammar is about (a generative model learned without supervision) and it misses out on lots of cool phenomena.
-
-Maximum entropy learner:~\cite{goldwater2003learning}.
-Whether you take an OT perspective or a generative perspective, the structure of the rules (or constraints) has to come from somewhere.
-How do you learn that structure?
-This is a program induction problem.
-
-Unsupervised phonological rule learners:
-~\cite{cotterell-peng-eisner-2015} learns weighted fst's while
-\cite{rulebased} learns context-sensitive rewrites. Here we show that
-these results and more can be reformulated as a program induction
-problem, and that doing so buys you (1) connections to probabilistic
-programming and Bayesian program learning; (2) theoretical guarantees;
-(3) a bunch more phenomena modeled, like supervised cases (pig latin),
-UG, and infant studies.
-
-References from a discussion with Adam Albright:
-
-LISA (an acronym standing for something) babbling
-
-Alex Cristia: Can infants learn phonology in the lab?
-
-David Stampe: Spontaneous Devoicing. Evidence of UG
-
-Chapter 8 of SPE~\cite{chomsky1968sound}: What is the right evaluation metric for rules?
-
-Collin Wilson: paletization generalizations as more evidence of UG in phonology
-
-Degree of articulatory effort as an inductive bias
-
-Rules shouldn't be too disruptive of the underlying form. Should also minimize confuseability. Final devoicing is a good example of this. PMAP. ``weak contrast''
-
-\section{Experiments}
-
-\subsection{Phonological Rule Induction}
-
-Several metrics of success here: (1) how well does it compress the
-data? (2) how well does it predict held out forms? (3) how natural or
-interpretable are the resulting solutions?
-
-Here are cases where I think it learns something cool:
-
-\noindent\begin{tabular*}{\textwidth}{lclll}
-  \toprule
-  Language&Example data&Underlying forms&Phonology&Phenomena\\\midrule
-Gen&
-\begin{tabular}{l}
-\textipa{sra}\\
-\textipa{agble}\\
-\textipa{drE}\\
-\textipa{hlE}
-\end{tabular}
-&\textipa{l} underlying
-  & \textipa{l}$\to$\textipa{r}$/$\verb|[+coronal] _|
-  &
-\textipa{l}$\sim$\textipa{r} alternation
-  \\
-  \bottomrule  \end{tabular*}
-
-\noindent\begin{tabular*}{\textwidth}{lclll}
-  \toprule
-  Language&Example data&Phonology&Morphology&Phenomena\\\midrule
-  Tibetan&
-\begin{tabular}{ll}
-1&\textipa{\|x{j}ig}\\
-4&\textipa{\|x{s}i}\\
-5&\textipa{Na}\\
-9&\textipa{gu}\\
-10&\textipa{\|x{j}u}\\
-11 ($= 10 + 1$) & \textipa{\|x{j}ug\|x{j}ig}\\
-14 ($= 10 + 4$) & \textipa{\|x{j}ub\|x{s}i}\\
-15 ($= 10 + 5$) & \textipa{\|x{j}uNa}\\
-19 ($= 10 + 9$) & \textipa{\|x{j}urgu}\\
-40 ($= 4 + 10$) & \textipa{\|x{s}ib\|x{j}u}\\
-50 ($= 5 + 10$) & \textipa{Nab\|x{j}u}\\
-90 ($= 9 + 10$) & \textipa{gub\|x{j}u}
-\end{tabular}
-&
-\begin{tabular}{l}
-  \verb|[-nasal]|$\to\varnothing $ $/$ \verb|# _|\\
-  \emph{or,}\\
-  \verb|C|$\to\varnothing $ $/$ \verb|# _ C|\\
-  \end{tabular}
-  &
-\emph{given in problem}
-  &
-Consonant cluster reduction
-  \\
-  \bottomrule  \end{tabular*}
-
-\begin{tabular*}{\textwidth}{lclll}
-  \toprule
-  Language&Example data&Phonology&Morphology&Phenomena\\\midrule
-  Kerewe&
-\begin{tabular}{llll}
-\textipa{kubala} & \textipa{kubalana} & \textipa{kubalila} & \textipa{kubalilana}\\
-\textipa{kugaya} & \textipa{kugayana} & \textipa{kugayila} & \textipa{kugayilana}\\
-\textipa{kub\'ala} & \textipa{kub\'al\'ana} & \textipa{kub\'al\'ila} & \textipa{kub\'al\'ilana}\\\hline
-\textipa{kut\'ub\'ala} & \textipa{kuk\'ib\'ala} & \textipa{kut\'ub\'alila} & \textipa{kuk\'it\'ubalila}\\
-\textipa{kut\'ug\'aya} & \textipa{kuk\'ig\'aya} & \textipa{kut\'ug\'ayila} & \textipa{kuk\'it\'ugayila}\\
-\textipa{kut\'ub\'ala} & \textipa{kuk\'ib\'ala} & \textipa{kut\'ub\'alila} & \textipa{kuk\'it\'ubalila}
-\end{tabular}
-&
-\begin{tabular}{l}
-  \verb|[]|$\to$\verb|[-H]|$/$\verb|[+H] [] _| \\
-  \verb|[]|$\to$\verb|[+H]|$/$\verb|[+H] [] _ []|
-  \end{tabular}
-  &
-\begin{tabular}{ll}
-\textipa{ku}$+$stem$+$\textipa{a}\\
-\textipa{ku}$+$stem$+$\textipa{ana}\\
-\textipa{ku}$+$stem$+$\textipa{ila}\\
-\textipa{ku}$+$stem$+$\textipa{ilana}\\
-\textipa{kut\'u}$+$stem$+$\textipa{a}\\
-\textipa{kuk\'i}$+$stem$+$\textipa{a}\\
-\textipa{kut\'u}$+$stem$+$\textipa{ila}\\
-\textipa{kuk\'it\'u}$+$stem$+$\textipa{ila}
-  \end{tabular}
-  &
-\begin{tabular}{ll}
-  Tone spreading\\
-  Rule ordering
-  \end{tabular}
-  \\
-  \bottomrule  \end{tabular*}
-
-\noindent\begin{tabular*}{\textwidth}{lclll}
-  \toprule
-  Language&Example data&Phonology&Morphology&Phenomena\\\midrule
-  Polish&
-\begin{tabular}{ll}
-\textipa{dom} & \textipa{domi}\\
-\textipa{kot} & \textipa{koti}\\
-\textipa{lut} & \textipa{lodi}\\
-\textipa{vus} & \textipa{vozi}\\
-\textipa{wuk} & \textipa{wugi}\\
-\textipa{ruk} & \textipa{rogi}\\
-\textipa{bur} & \textipa{bori}\\
-\textipa{\|x{s}um} & \textipa{\|x{s}umi}
-\end{tabular}
-&\begin{tabular}{l}
-   \textipa{o}$\to$\textipa{u}$/$\verb|_[-nasal +voice]#|\\%
-\verb|[-sonorant]|$\to$\verb|[-voice]|$/$\verb|_#|
-   \end{tabular}
-  &
-\begin{tabular}{ll}
-  stem\\
-  stem$ + $\textipa{i}
-  \end{tabular}
-  &
-\begin{tabular}{ll}
-  Final devoicing\\
-  Rule ordering
-  \end{tabular}
-  \\
-  \bottomrule  \end{tabular*}
-
-\noindent\begin{tabular*}{\textwidth}{lclll}
-  \toprule
-  Language&Example data&Phonology&Morphology&Phenomena\\\midrule
-Makonde  &
-\begin{tabular}{lll}
-\textipa{am\'aNga} & \textipa{am\'ile} & \textipa{\'ama}\\
-\textipa{ak\'aNga} & \textipa{ak\'ile} & \textipa{\'aka}\\
-\textipa{av\'aNga} & \textipa{av\'ile} & \textipa{\'ova}\\
-\textipa{am\'aNga} & \textipa{am\'ile} & \textipa{\'oma}\\
-\textipa{ut\'aNga} & \textipa{ut\'ile} & \textipa{\'uta}\\
-\textipa{av\'aNga} & \textipa{av\'ile} & \textipa{\'eva}\\
-\textipa{tav\'aNga} & \textipa{tav\'ile} & \textipa{t\'ava}\\
-\textipa{uNg\'aNga} & \textipa{uNg\'ile} & \textipa{\'uNga}\\
-\textipa{pat\'aNga} & \textipa{pat\'ile} & \textipa{p\'ota}
-\end{tabular}
-&
-\begin{tabular}{l}
-\verb|[]|$\to$\verb|[-stress]| $/$ \verb| _ [ ]* [ +stress]|\\
-\verb|[+middle -stress]|$\to$\textipa{a}$/$\verb| _ [ ]|
-  \end{tabular}
-  &
-\begin{tabular}{ll}
-  stem$ + $\textipa{\'aNga}\\
-  stem$ + $\textipa{\'ile}\\
-  stem$ + $\textipa{a}
-  \end{tabular}
-  &
-\begin{tabular}{ll}
-Stress patterns\\
-Neutralizing rule
-  \end{tabular}
-  \\
-  \bottomrule  \end{tabular*}
-
-\noindent\begin{tabular*}{\textwidth}{lclll}
-  \toprule
-  Language&Example data&Phonology&Morphology&Phenomena\\\midrule
-  Kikuria&
-\begin{tabular}{ll}
-\textipa{rema} & \textipa{remera}\\
-\textipa{roma} & \textipa{romera}\\
-\textipa{tiga} & \textipa{tegera}\\
-\textipa{ruga} & \textipa{rogera}\\
-\textipa{hoora} & \textipa{hoorera}\\
-\textipa{siika} & \textipa{seekera}\\
-\textipa{huuta} & \textipa{hootera}\\
-\textipa{suraaNga} & \textipa{suraaNgera}
-\end{tabular}
-&
-\verb|[+high]|$\to$\verb|[+middle]|$ /$\verb| _ [-low]* [+middle]|
-  &
-\begin{tabular}{ll}
-  stem$ + $\textipa{a}\\
-  stem$ + $\textipa{era}
-  \end{tabular}
-  &
-Vowel harmony
-  \\
-  \bottomrule  \end{tabular*}
-
-\noindent\begin{tabular*}{\textwidth}{lclll}
-  \toprule
-  Language&Example data&Phonology&Morphology&Phenomena\\\midrule
-  Hungarian&
-  \begin{tabular}{llll}
-\textipa{a:g\super y} & \textipa{a:g\super yban} & \textipa{a:k\super yto:l} & \textipa{a:g\super ynak}\\
-\textipa{\"o:r} & \textipa{\"o:rben} & \textipa{\"o:rt\"o:l} & \textipa{\"o:rnek}\\
-\textipa{ku:t} & \textipa{ku:dban} & \textipa{ku:tto:l} & \textipa{ku:tnak}\\
-\textipa{re:s} & \textipa{re:zben} & \textipa{re:st\"o:l} & \textipa{re:snek}\\
-\textipa{rab} & \textipa{rabban} & \textipa{rapto:l} & \textipa{rabnak}\\
-\textipa{vi:z} & \textipa{vi:zben} & \textipa{vi:st\"o:l} & \textipa{vi:znek}\\
-\textipa{fal} & \textipa{falban} & \textipa{falto:l} & \textipa{falnak}\\
-\textipa{test} & \textipa{tezdben} & \textipa{testt\"o:l} & \textipa{testnek}
-  \end{tabular}&
-  \begin{tabular}{l}
-    \verb|V|$\to$\verb|[+mid +tense +front]|$ /$\\\verb|  [ +front ] [ ]* _ |\\
-         \verb|[ ]|$\to$\verb|[+voice]|$ / $\\\verb| _ [+bilabial]|\\
-         \verb|[-sonorant]|$\to$\verb|[-voice]|$ / $\\\verb| _ [-voice]|
-    \end{tabular}
-  &
-  \begin{tabular}{l}
-    stem\\
-    stem$ + $\textipa{ban}\\
-    stem$ + $\textipa{to:l}\\
-    stem$ + $\textipa{nak}
-  \end{tabular}
-  &
-  \begin{tabular}{l}
-    Vowel harmony\\
-    Voicing assimilation\\
-    Rule ordering
-    \end{tabular}
-  \\
-\bottomrule  \end{tabular*}
-
-\pagebreak
-These problems are solved:
-\begin{itemize}
-\item Kikurai: fricatives alternate with stops after nasals
-\item Greek: Velar stops alternate with palletized versions before front vowels
-\item Farsi: Trills alternate with flaps
-\item Osage: coronal stops become dentals before central vowels
-\item Amharic: alternation between \textipa{@} \& \textipa{E}
-\item Gen: l/r alternation
-\item Kishambaa: voiced/unvoiced nasals alternate
-\item  Thai: stops are unreleased word finally
-\item Palauan: a word initial neutralizing rule
-\item Quechua: Velar becomes uvular when followed by uvular (spreading)
-\item Lhasa Tibetan: no contrast between velar/uvular, or voiced/voiceless stops or fricatives
-\item Axininca Campa: stops become glides
-\item Kikuyu: infinitive prefix can surface as either k or \textipa{G}
-\item Korean: vowel harmony, aspiration only surfaces in certain contexts
-\item Hungarian: vowel harmony and voicing assimilation
-\item Kikuria: vowel harmony
-\item Farsi: a deletion rule explains singular/plural
-\item Tibetan: initial consonant cluster reduction explains counting system
-\item Makonde: stress patterns; unstressed vowels are neutralized
-\item North Saami: 3 neutralization rules explain nominative sg essive
-\item Samoan: 2 deletion rules that explain words that sound the same in one inflection being different in another
-\item     Russian: devoicing of word final obstruent
-\item English: verbal inflections (voicing assimilation, epenthesis)
-\item Finnish: nominative/partive explained by vowel raising and vowel harmony
-\item Kerewe: Interacting tone rules
-\item Polish: vowel alternations interacting with devoicing of word final obstruent in singular/plural
-\item Ancient Greek: voicing assimilation, deaspiration, deletion rule. Order matters.
-\item Serbo-Croatian: predictable stress, devoicing, neutralizing, epenthesis
-\end{itemize}
-17 textbook problems are not solved.
-2 of these need a tier-based representation.
-A couple more of these are due to limitations in the program learner's representation of rules.
-The rest are because they involve scaling to more than 3 rules, which is about the practical limit for exact search with these methods.
-\pagebreak
-
-\subsubsection{An approximation to UG: Program fragments}
-
-Within-rule reuse of structure:
-\begin{tabular}{ll}
-  Tibetan: & {\color{red}\texttt{C}}$\to\varnothing /$\verb|#_|{\color{red}\texttt{C}} \\
-  English: &          \verb|[-sonorant]|$\to${\color{red}\texttt{[-voice]}}$ / ${\color{red}\texttt{[-voice]}}\verb| _ |\\
-Kikuria: &\verb|[ +high ]|$\to$ {\color{red}\texttt{[ +middle ]}} $/$ \verb| _ [ -low ]*| {\color{red}\texttt{[ +middle ]}}
-\end{tabular}
-
-\vspace{0.5cm}\noindent Within-language reuse of structure:
-\begin{tabular}{ll}
-  Axininca Campa: &
-  \begin{tabular}{l}
-    \textipa{p}{\color{red}$\to$}\textipa{w} {\color{red}$/$ \texttt{[ ] \_}}\\
-    \textipa{k}{\color{red}$\to$}\textipa{y}\, {\color{red}$/$ \texttt{[ ] \_}}
-  \end{tabular}\\
-  English:&
-  \begin{tabular}{l}
-    {\color{red}$\varnothing\to$\textipa{@} $/$} \verb|[+sib] _ [+sib]|\\
-    {\color{red}$\varnothing\to$\textipa{@} $/$} \verb|[+cor +stop] _ [+cor +stop]|
-  \end{tabular}
-\end{tabular}
-
-\vspace{0.5cm}\noindent Cross-language reuse of structure:
-\begin{tabular}{ll}
-  Russian/Polish:& \verb|[-sonorant]|$\to$\verb|[-voice]|$/$\verb|_#|\\
-  Samoan/Tibetan:&
-  \begin{tabular}{l}
-    {\color{red}\texttt{C}$\to\varnothing /$}\verb|_#|\\
-    {\color{red}\texttt{C}$\to\varnothing /$}\verb|#_C|
-  \end{tabular}\\
-  Ancient Greek/Hungarian:&
-  \begin{tabular}{l}
-    {\color{red}\texttt{[-sonorant]}$\to$\texttt{[-voice}} \texttt{-aspirated]} {\color{red} $/$ \texttt{ \_ [-voice]}}\\
-    {\color{red}\texttt{[-sonorant]}$\to$\texttt{[-voice}} \texttt{]} {\color{red} $/$ \texttt{ \_ [-voice]}}
-    \end{tabular}
-  \end{tabular}
-
-\begin{figure}[h]
-  \includegraphics[width = \textwidth]{featureFrequencies.png}
-  \caption{Which features are frequently used in correct rules?
-Voicing, sonorant, nasality are very frequent, and are also near the top of the feature geometry hierarchy.
-Feature geometry was not given to the system, but you can see at least this coarse level of structure emerge from the data.
-Vowel features also frequent due to vowel harmony occurring in lots of languages.}
-\end{figure}
-
-\begin{figure}[h]
-  \includegraphics[width = \textwidth]{skeletonFrequencies.png}
-  \caption{What is the general structure of rules that occur in phonology, ignoring the particular constants that occur in the rule?
-Here $K$ stands in for an arbitrary phoneme; $\pm$features stands in for an arbitrary feature matrix.
-From this plot you can glean that (1) neutralizing rules are very common (top two most frequent rules); (2) harmony/assimilation rules are also frequent; (4) deletion rules tend to occur at word boundaries.
-There are about $5\times 10^4$ possible syntactic structures for rules, but only 15 are actually attested (relative frequency$ > 1$).
-Here I think that the sparsity of the problem sets is hurting us: these results come from 26 problems, so the fact that you see only 15 general schemas isn't too crazy if you don't believe in strong innate constraints on the structure of rules.
-But looking at this data you also see a generalization that holds across all of the rules learned: the structures tend to be pretty small, so you never see a rule that checks feature values of more than two adjacent segments.}
-\end{figure}
-
-%% \subsection{Historical productivity}
-
-%% There is an interesting tension between discovering regularities
-%% that are psychologically real (like things that are used productively)
-%% and discovering regularities that are in the data but only due to
-%% historical accidents. Our system cannot distinguish them.
-
-%% Are these deep analyses real? We don't answer this question, but we do
-%% raise it. We are agnostic. These are the kinds of regularities that a
-%% good linguist would notice, and our system notices them too.
-
-\subsection{Artificial Phonologies}
-
-It would be cool to model all of the grammars in~\cite{frank2011three}.
-Right now we can model AAB/ABA patterns like in Gary Marcus's famous infant experiments.
-Other infant studies with artificial grammars:~\cite{gerken2006decisions}.
-MNPQ problem:~\cite{smith1966grammatical}.
-
-Analogs in real languages: agreement morphology (eg gender in Spanish
-- there is a constraint that two things have to be the same);
-reduplication (this could literally be modeled using the same ABA
-system); obligatory contour principle (opposite of harmony - two
-things have to be different, just like the ssd/sds distinction).
-
-For any of these problems, there is a trade-off between how
-complicated of a program you are willing to use and how badly you want
-to compress the data. Here, compressing the data means representing
-the surface forms with short underlying representations. This
-trade-off is of course present in the phonology problem sets as well.
-But the problem sets are complicated and have lots of data, precluding
-any in-depth analysis of the trade-off except quantitatively. For
-these synthetic grammar learning problems, we can dig in and see
-exactly what the trade-off looks like for different amounts of data.
-See figure~\ref{fronts} for an analysis of this trade-off along with
-examples of what the programs look like.
-
-\section{Applications}
-The point of this is to motivate why you would want to solve exactly
-this problem even though it doesn't do anything more immediately
-useful than solve your phonology homework. Practical applications that could come from this line of work:
-
-\begin{itemize}
-\item Low-resource language acquisition: \url{http://sapience.dec.ens.fr/bootphon/} for some languages there is very little data and very few speakers
-\item Mathematica for Phonology: a lot of the time physicists don't do
-  integrals; they just punch them into Mathematica. This is a tool
-  that provably finds the ``best'' grammar for some data, sort of like
-  how Mathematica finds the correct integral of some expression. And
-  just like some integrals are beyond the reach of computer algebra
-  systems, some data sets might be beyond the reach of this
-  hypothetical tool. Analogy to the work
-  in~\cite{schmidt2009distilling}. According to Tim it would be more
-  practical for the field for knowledge is to have a system that can
-  work out the consequences of a given rule set, maybe pointing out
-  places where it's inconsistent with the data or suggesting
-  underlying forms. These alternative uses could also be accommodated.
-\item A probe for UG: the prior over phonology is a parameter of the
-  algorithm which you could in principle estimate from the data or
-  compare different settings of the prior to see how well they
-  compress the data. Or, come up with the right constraints so that
-  the system doesn't generalize in the same places where humans don't
-  generalize (surfeit of the stimulus:~\cite{becker2011surfeit}). You
-  can approach universal grammar from two directions: what do I need
-  to build then so that it will learn the same things that humans
-  learn? And what do I need to take out in order to stop it from
-  learning things that humans never learn? 
-\end{itemize}
-
-\section{Discussion}
-
-Deep learnability issue: what is the actual computational complexity
-of learning these grammars? The actual hardness is an empirical
-question which you can upper bound by coming up with better heuristics
-for inducing grammars. Attempts at \emph{lower bounding} the
-complexity of learning has generally been way too conservative. We
-want to know the actual hardness of the learning problem, and
-constructing a practical program learner shows that it might not be
-too hard with the right UG.
-
-We must emphasize that there is \textbf{no general-purpose
-  learner}. So we need UG. This tool lets us dig in and study it.
-There's always going to be a trade-off between how much you build in and how much data/computation is needed to recover the correct generalizations.
-Our compromise in the space of this trade-off is to:
-\begin{itemize}
-\item  Not play in the space of arbitrary turing machines but instead search through a finite program space (bounded program size runtime and memory consumption; program structure given by a grammar)
-  \item Make contentfull proposals about what the structure of
-    universal grammar might be, and test them empirically on many
-    languages
-\end{itemize}
-
-\textbf{Comparison with deep learning:} In a sense this project is like
-the opposite of deep learning. It's about the kind of learning that
-yields explanations of the causal processes behind data sets; where
-the learner comes equipped with explicit background knowledge (UG)
-that we can tweak finely and systematically.
-
-One similarity though is that some of the phenomena we model, like
-infant learning of synthetic grammars, are the sort of things that
-humans pick up unconsciously. A lot of deep learning might be
-analogous to humans rote learning some motor or perceptual task.
-
 \end{document}

### 2018-02-22 topics.tex
topics Pareto frontier writing

@@
 
+As an example of how the Pareto front offers a lens on the
+space of possible hypotheses,
+consider the upper left Pareto front in figure~\ref{quadrupleFrontier} (\emph{aab, 1 example}).
+Here the learner has seen a single word, [\textipa{vEfefe}].
+Some grammars living on the Pareto frontier are:
+\begin{itemize}
+\item \textbf{A grammar that does nothing:} In the lower right-hand corner of~\ref{quadrupleFrontier}
+  is the grammar labeled ``surface=underlying'',
+  which just says that every word (a ``surface pronunciation'')
+  is represented (``underlyingly'') literally how it is pronounced.
+  The observed word [\textipa{vEfefe}] is represented as /\textipa{vEfefe}/, which has 6 symbols,
+  giving a fit to the data of -6.
+\item \textbf{A grammar that duplicates syllables:} Highlighted in dark red in~\ref{quadrupleFrontier}
+  is a grammar with the single rule $\varnothing\to\sigma_i/\_\sigma_i$\texttt{\#},
+  which inserts a copy of the last syllable syllable.
+  This rule generates the word [\textipa{vEfefe}] by starting with the stem (i.e., underlying form) /\textipa{vEfe}/ (which has 4 symbols, fit to data of -4) and then applying the rule in the grammar,
+  which copies the last syllable and makes [\textipa{vEfefe}].
+\item \textbf{A grammar that duplicates syllables and appends
+  morphemes:} Highlighted in pink in~\ref{quadrupleFrontier} are
+  grammars that duplicate a syllable but also append or prepend extra
+  morphemes.  These correspond to generalizations where the learner
+  believes that different parts of [\textipa{vEfefe}] correspond to
+  prefixes or suffixes in the language.  For example, the grammar in
+  the upper left corner incorporates the suffix /\textipa{fe}/ as well
+  as the prefix /\textipa{vE}/, and explains the observation using an
+  underlying form that is completely empty (fit to data of 0) -- this
+  grammar has memorized the observed data, and so maximally compresses
+  it, but at the cost of having many symbols inside of the grammar (22 symbols, vs 6 symbols in the grammar that just duplicates a syllable).  
+  \end{itemize}
+
 \begin{figure}[H]
@@
 latent generative model (a grammar).
-Figure~\ref{naturalLanguageExample} shows our model solving a
+
+These natural language results are still pulmonary and
+ongoing. Figure~\ref{naturalLanguageExample} shows our model solving a
 representative textbook problem.  We believe this extension two
 natural language is an important next step for models of artificial
-grammar learning,
-and helps deliver on the basic promises and presuppositions of AGL research.
+grammar learning, and helps deliver on the basic promises and
+presuppositions of AGL research.
 

### 2018-02-22 topics.tex
small amount of writing

@@
 
-\title{Automatically generating alternative hypotheses and controls}
+\title{Automatically Exploring Hypothesis Spaces}
 %\author{Kevin Ellis, Timothy O'Donnell}
@@
 prior and likelihood.
+
+In this work we will use this probabilistic framing
+to explore spaces of grammars
+that an AGL learner could use to explain
+the pronunciations of words.
+Here the grammars will be SPE-style rewrite rules~\cite{chomsky1968sound},
+commonly used in generative phonology,
+but at a high level this framing can apply to
+learning other aspects of grammar like semantics~\cite{piantadosi2011learning}, syntax~\cite{perfors2011learnability}, and morphology~\cite{tim}.
 
@@
   \toprule
-  Grammar&Example input to learner&MAP grammar&Natural language analogues\\\midrule
+  Grammar&Example input to learner&Grammar we infer&Natural language analogues\\\midrule
   \begin{tabular}{l}

### 2018-03-18 topics.tex
prettier rules

@@
 \usepackage{amsfonts}       % blackboard math symbols
-
+\usepackage{phonrule} 
  % For algorithms
@@
 
-The goal of the grammar learner, whether it is a child, linguist, or computer program, is to
+A goal of the grammar learner, whether it is a child, linguist, or computer program, is to
 infer the grammar that gave rise to
@@
 prior and likelihood.
-
 In this work we will use this probabilistic framing
@@
 
-A simple and intuitive prior over grammars is one which penalizes longer grammars and favors parsimonious grammars, for example defining $\probability\left[\text{grammar} \right]\propto\exp\left(- \text{\# symbols in grammar} \right)$.
+\noindent\textbf{SPE-style rewrite notation:} Each SPE-style rewrite rule is a function that both inputs and outputs a sequence of phonemes.
+These rewrites are written as ``\phonb{input}{output}{left}{right}'' %$\text{input}\to\text{output }/\text{ left } \_ \text{ right}$'',
+which means that ``input'' gets rewritten to ``output'' whenever ``left'' is to the left and ``right'' is to the right.
+Some special symbols used in these rules are \# to mean the start or end of the input;
+C/V to mean a consonant/vowel; $\sigma $ to mean a syllable;
+and $\varnothing$ to mean the empty string.
+For example, the rule
+\phonb{$\sigma $}{$\varnothing$}{\#}{C}
+deletes a syllable at the start of a word if it is followed by a consonant.
+Rules can refer to sets of phonemes by writing down vectors of phonological features -- for example,
+the rule \phon{[-son]}{[+voice]} rewrites segments that are -sonorant (e.g., \textipa{k,d,v,S,...}) to their voiced counterparts (e.g., \textipa{g,d,v,Z,...}).
+Rules can also bind variables to phonemes or syllables,
+which are written using subscripts;
+for example,
+\phonr{C$_i$}{$\varnothing$}{C$_i$}
+%$\text{\texttt{C}}_i\to\varnothing / \_ \text{\texttt{C}}_i$
+deletes a consonant if it is followed by the same consonant,
+and
+\phonb{$ \varnothing $}{$\sigma _i$}{#}{$\sigma _i$}
+%$\varnothing\to\sigma_i / \text{\texttt{\#}}\_\sigma_i$
+inserts a copy of a word initial syllable.
+The special subscript ``0" means zero or more occurrences of the subscripted segment,
+so for example \phonl{V}{[+hi]}{[+hi][ ]$_0$} means that vowels become +hi whenever there exists a +hi phoneme anywhere to the left.
+
+\noindent \textbf{Prior \& Likelihood:} A simple and intuitive prior over grammars is one which penalizes longer grammars and favors parsimonious grammars, for example defining $\probability\left[\text{grammar} \right]\propto\exp\left(- \text{\# symbols in grammar} \right)$.
 An example of a likelihood model,
@@
 instead studying AGL in controlled experiments.
-Our contribution here is the demonstration that
-a single BPL model can explain how a learner can quickly infer a grammar represented as SPE--style rules.
+Our demonstration here is that
+a BPL model can explain how a learner can quickly (e.g., from very little data) infer a grammar represented as SPE--style rules.
 We collected grammars from AGL datasets~\cite{gerken2010infants,marcus1999rule,frank2011three}
@@
   \toprule
-  Grammar&Example input to learner&Grammar we infer&Natural language analogues\\\midrule
+  Grammar&\begin{tabular}{c}
+    Example input \\to learner
+    \end{tabular}&Inferred grammar&Natural language analogues\\\midrule
   \begin{tabular}{l}
-    ABA (same/different/same)\\
+    ABA \\
   Marcus et al. 1999.     
@@
   \begin{tabular}{l}
-$\varnothing\to \sigma_i$\verb| / #_|$\sigma \sigma_i$
+    \phonb{$\varnothing $}{$\sigma _i$}{\#}{$\sigma \sigma _i$}
+%$\varnothing\to \sigma_i/$\texttt{\#\_}$\sigma \sigma_i$
   \end{tabular}& Reduplication (eg, Tagalog)
@@
   \begin{tabular}{l}
-      ABB (different/same/same)\\
+      ABB\\
   Marcus et al. 1999. 
@@
   \begin{tabular}{l}
-$\varnothing\to \sigma_i$\verb| / |$\sigma_i$\verb|_#|
+    \phonb{$\varnothing $}{$\sigma _i$}{$\sigma _i$}{\#}
+    %$\varnothing\to \sigma_i/\sigma_i$\texttt{\_\#}
   \end{tabular}&
@@
   \begin{tabular}{l}
-      ABx (different/different/constant)
+      ABx 
     \end{tabular}&
@@
   \begin{tabular}{l}
-stem$ + x$
+stem$+x$
   \end{tabular}&concatenative morphology
@@
   \begin{tabular}{l}
-    AAx (same/same/constant)\\Gerken 2006. %~\cite{gerken2010infants}
+    AAx \\Gerken 2006. %~\cite{gerken2010infants}
     \end{tabular}&
@@
   \begin{tabular}{l}
-    stem + $x$\\
-$\varnothing\to \sigma_i$\verb| / #_|$\sigma_i$
+    stem+$x$\\
+    \phonb{$\varnothing $}{$\sigma _i$}{\#}{$\sigma _i$}
+%$\varnothing\to \sigma_i/$\texttt{\#\_}$\sigma_i$
   \end{tabular}&
@@
   \begin{tabular}{l}
-    AxA (same/constant/same)\\Gerken 2006. %~\cite{gerken2010infants}
+    AxA \\Gerken 2006. %~\cite{gerken2010infants}
     \end{tabular}&
@@
   \begin{tabular}{l}
-$x + $ stem\\
-    $\varnothing$ $\to\sigma_i$  / \verb|# _| $\sigma$ $\sigma_i$
+    $x+$stem\\
+    \phonb{$\varnothing $}{$\sigma _i$}{\#}{$\sigma \sigma _i$}
+    %$\varnothing$ $\to\sigma_i/$\texttt{\#\_}  $\sigma$ $\sigma_i$
   \end{tabular}&
@@
   \begin{tabular}{l}
-      $\varnothing\to$\verb|C|$_i$ $/$ \verb|#C|$_i$\verb| [ ]* _ #|\\%
-      $\varnothing\to$ \textipa{e} $/$ \verb| _ #|\\%
- \verb|C|$\to\varnothing$ $/$ \verb|# _|
+    \phonb{$\varnothing $}{C$_i$}{\#C$_i$[ ]$_0$}{\#}\\
+    \phonr{$\varnothing $}{\textipa{e}}{\#}\\
+    \phonl{C}{$\varnothing $}{\#}
+ %%      $\varnothing\to$\texttt{C}$_i$ $/$\texttt{\#C}$_i$\texttt{[ ]*\_\#}\\%
+ %%      $\varnothing\to$ \textipa{e} $/$\texttt{\_\#}\\%
+ %% \texttt{C}$\to\varnothing$ $/$ \texttt{\#\_}
   \end{tabular}  & \begin{tabular}{l}
-    \textbf{Tim, is there an  analog?}\\
-    \textbf{word initial segments moves to  }\\
-    \textbf{word final?}
     \end{tabular}
-  \bottomrule  \end{tabular}
+\\  \bottomrule  \end{tabular}
 \caption{Using BPL to model artificial grammar learning. Learner given 5 examples.}\label{artificialGrammarTable}
@@
 \item \textbf{A grammar that duplicates syllables:} Highlighted in dark red in~\ref{quadrupleFrontier}
-  is a grammar with the single rule $\varnothing\to\sigma_i/\_\sigma_i$\texttt{\#},
+  is a grammar with the single rule
+  \phonr{$\varnothing $}{  $\sigma _i$}{$\sigma _i$\#}
+  %$\varnothing\to\sigma_i/\_\sigma_i$\texttt{\#},
   which inserts a copy of the last syllable syllable.
@@
 
-\includegraphics[width = 0.5\textwidth]{abb1.png} \includegraphics[width = 0.5\textwidth]{abb3.png}
-\includegraphics[width = 0.5\textwidth]{aax1.png} \includegraphics[width = 0.5\textwidth]{aax3.png}
+\includegraphics[width = 0.5\textwidth]{alternativeHypothesesFigures/abb1.png} \includegraphics[width = 0.5\textwidth]{alternativeHypothesesFigures/abb3.png}
+\includegraphics[width = 0.5\textwidth]{alternativeHypothesesFigures/aax1.png} \includegraphics[width = 0.5\textwidth]{alternativeHypothesesFigures/aax3.png}
 
@@
 
-These natural language results are still pulmonary and
+These natural language results are still preliminary and
 ongoing. Figure~\ref{naturalLanguageExample} shows our model solving a
-representative textbook problem.  We believe this extension two
+representative textbook problem.  We believe this extension to
 natural language is an important next step for models of artificial
@@
   \end{minipage}%
-  \hspace{0.5cm}%
-  \begin{minipage}{6cm}\centering
+  \hspace{0.25cm}
+  \begin{minipage}{7cm}\centering
     \textbf{Our model's solution:}
 
-    Induced phonological rule: \texttt{C}$\to\varnothing /$\texttt{\#\_C} \\
+    Induced phonological rule: \phonb{C}{$\varnothing $}{\#}{C} \\ %\texttt{C}$\to\varnothing /$\texttt{\#\_C} \\
     \hspace{0.5cm}\emph{(Reduce word initial consonant clusters)}
@@
 
-%\pagebreak
-%% \begin{figure}
-%%   \begin{tikzpicture}
-%%     %  \draw[help lines] (0,0) grid (10,-10);
-%%     \node(UG)[purple,align = center,rotate = 90] at (1,-1) {\textbf{Universal}\\\textbf{Grammar}};
-%%     %\draw[ultra thick] (2,0) rectangle (13,-2.3);
-
-%%     \node(sounds) at (7,-0.3) {Phoneme Inventory};
-%%     \node(soundExamples)[draw,align = center] at (7,-1.3) {\textipa{@, p, E, 2, k}\\\textipa{a, 3, A, O, d}\\$\cdots$};
-
-%%     \node at (11,-0.3) {Program space};
-%%     \node(programSpace)[draw,align = left] at (11,-1.3) {$\mathcal{M}\to \mathcal{A} | $\verb|stem|$|\mathcal{M}+\mathcal{M}$\\$\mathcal{P}\to \mathcal{R}\circ \mathcal{R}\circ \cdots \circ \mathcal{R}$};
-
-%%     \node(featureSpace) at (3.5,-0.3) {Feature space};
-%%     \node(featureExamples)[draw,align = center] at (3.5,-1.3) {\verb|voiced|, \verb|vowel|\\\verb|obstruent|, \verb|nasal|\\$\cdots$};
-
-%%     \draw[thick,->] (featureExamples.east) -- (soundExamples.west);
-%%     \draw[thick,->] (soundExamples.east) -- (programSpace.west);
-
-%%     \node(universalBox)[purple,draw,ultra thick, fit = (sounds) (featureSpace) (featureExamples) (soundExamples) (programSpace)] {};
-%%     \draw(lowerCausalArrow)[thick,->] (featureExamples.south) .. controls (7,-2.5) ..  (programSpace.south);
-
-%%     \node(morphology)[blue,align = above,rotate = 90] at (1,-5) {\textbf{Programs}};
-%%     \draw[ultra thick,->] (UG.west) -- (morphology.east);
-
-%%     \node(ABAmorphologyLabel) at (10.5,-3.5) {Pig Latin Morphology};
-%%     \node(ABAmorphology) [draw,align = left] at (10.5,-4){ \verb|observation|$\to$\verb|stem|};
-
-%%     \node(ABAphonologyLabel) at (10.5,-5.3) {Pig Latin Phonology};
-%%     \node(ABAphonology) [draw,align = left] at (10.5,-6.3) {%
-%% $r_1$:      $\varnothing\to$\verb|C|$_i$ $/$ \verb|#C|$_i$\verb| [ ]* _ #|\\%
-%%       $r_2$:      $\varnothing\to$ \textipa{e} $/$ \verb| _ #|\\%
-%%     $r_3$: \verb|C|$\to\varnothing$ $/$ \verb|# _|};
-%%     \node(ABAmorphologyBox)[blue,draw,ultra thick, fit = (ABAmorphologyLabel) (ABAphonology)] {};
-
-%%     %  \draw[ultra thick] (2,-3) rectangle (7,-7);
-
-%%     \node(polishMorphologyLabel) at (4.5,-3.5) {Polish Morphology};
-%%     \node(polishMorphology) [draw,align = left] at (4.5,-4.3) {\verb|singular|$\to$ \verb|stem|\\\verb|plural|$\to$ \verb|stem|$ + $ /\textipa{i}/};
-
-%%     \node at (4.5,-5.5) {Polish Phonology};
-%%     \node(polishPhonology) [draw,align = left] at (4.5,-6.3) {%
-%%       $r_1$: \textipa{o}$\to$\textipa{u}$/$\verb|_[-nasal +voice]#|\\%
-%%       $r_2$: \verb|[-sonorant]|$\to$\verb|[-voice]|$/$\verb|_#|};
-%%     \node(polishMorphologyBox)[blue,draw,ultra thick, fit = (polishMorphologyLabel) (polishPhonology)] {};
-
-%%     \draw[ultra thick,->] (universalBox.south) -- (ABAmorphologyBox.north);
-%%     \draw[ultra thick,->] (universalBox.south) -- (polishMorphologyBox.north);
-
-%%     \node(lexicon)[magenta,align = center,rotate = 90] at (1,-10) {\textbf{Words}};
-%%     \draw[ultra thick,->] (morphology.west)-- (lexicon.east);
-
-%%     \node(p1) [align = left] at (9.5,-8) {\verb|LATIN:| \textipa{l\ae tIn}};
-%%     \node(p2) [draw,align = left] at (9.5,-9.5) {\textipa{l\ae tIn}\nextForm{r_1}\\\textipa{l\ae tInl}\nextForm{r_2}\\\textipa{l\ae tInle}\nextForm{r_3}\\\textipa{\ae tInle}}; %\\\textipa{l\ae tInle}}};
-%%     \draw [thick,->] (p1.south) -- (p2.north);
-
-%%     \node(w1) [align = left] at (12,-8) {\verb|ASK:| \textipa{\ae sk}};
-%%     \node(w2) [draw,align = left] at (12,-9.5) {\textipa{\ae sk}\nextForm{r_1}\\\textipa{\ae sk}\nextForm{r_2}\\\textipa{\ae ske}\nextForm{r_3}\\\textipa{\ae ske}};
-%%     \draw [thick,->] (w1.south) -- (w2.north);
-
-%%     \node[magenta,draw,align= left](ABAvocabulary) at (10.5,-11.5){\textipa{\ae tInle}, \textipa{\ae ske}, $\cdots$ };
-%%     \draw [thick,->] (p2.south) -- (ABAvocabulary.north);
-%%     \draw [thick,->] (w2.south) -- (ABAvocabulary.north);
-%%     \node(ABAlexiconBox)[draw,ultra thick, fit = (p1) (w1) (ABAvocabulary)] {};
-%%     \draw [ultra thick,->] (ABAmorphologyBox.south) -- (ABAlexiconBox.north);
-
-%%     \node(crib) at (4.5,-8) {\verb|horn:| \textipa{rog}};
-%%     \node(Cingular)[draw,align = left] at (3.5,-9.5) {\verb|singular:|\\ \textipa{rog}\nextForm{r_1}\\\textipa{rug}\nextForm{r_2}\\\textipa{ruk}};
-%%     \node(plural)[draw,align = left] at (5.5,-9.5) {\verb|plural:|\\ \textipa{rogi}\nextForm{r_1}\\\textipa{rogi}\nextForm{r_2}\\\textipa{rogi}};
-%%     \draw [thick,->] (crib.south) -- (Cingular.north);
-%%     \draw [thick,->] (crib.south) -- (plural.north);
-
-%%     \node[magenta,draw,align = left] (polishObservation) at (4.5,-11.5) {\verb|horn, singular:| \textipa{\v{z}wop}\\\verb|horn, plural:| \textipa{\v{z}wubi}};
-%%     \draw [thick,->] (Cingular.south) -- (polishObservation.north);
-%%     \draw [thick,->] (plural.south) -- (polishObservation.north);
-
-%%     \node(polishLexicon)[draw,ultra thick, fit = (crib) (Cingular) (plural) (polishObservation)] {};
-%%     \draw [ultra thick,->] (polishMorphologyBox.south) -- (polishLexicon.north);
-
-%%   \end{tikzpicture}
-%% \end{figure}
-
-%\bibliographystyle{named}
 \bibliographystyle{unsrt} \bibliography{main}

### 2018-04-29 test.tex
some more problem solutions

@@
 \end{tabular}
+\pagebreak
+\emph{Korean:}\\\begin{longtable}{lllllllllllll|l}\toprule
+$\varnothing$ $+$stem$+$ \textipa{man} & $\varnothing$ $+$stem$+$ \textipa{maNk\super h1m} & $\varnothing$ $+$stem$+$ \textipa{nar1m} & $\varnothing$ $+$stem$+$ $\varnothing$ & $\varnothing$ $+$stem$+$ \textipa{tero} & $\varnothing$ $+$stem$+$ \textipa{kwa} & $\varnothing$ $+$stem$+$ \textipa{pota} & $\varnothing$ $+$stem$+$ \textipa{kat\super Si} & $\varnothing$ $+$stem$+$ \textipa{i} & $\varnothing$ $+$stem$+$ \textipa{1n} & $\varnothing$ $+$stem$+$ \textipa{e} & $\varnothing$ $+$stem$+$ \textipa{ita} & $\varnothing$ $+$stem$+$ \textipa{1ro} & UR
+\\ \midrule
+\textipa{pamman} & \textipa{pammaNk\super h1m} & \textipa{pamnar1m} & \textipa{pap} & \textipa{paptero} & \textipa{papkwa} & \textipa{pappota} & \textipa{papkat\super Si} & \textipa{papi} & \textipa{pap1n} & \textipa{pape} & \textipa{papita} & \textipa{pap1ro} &  -- \\
+\textipa{summan} & \textipa{summaNk\super h1m} & \textipa{sumnar1m} & \textipa{sup} & \textipa{suptero} & \textipa{supkwa} & \textipa{suppota} & \textipa{supkat\super Si} & \textipa{sup\super hi} & \textipa{sup\super h1n} & \textipa{sup\super he} & \textipa{sup\super hita} & \textipa{sup\super h1ro} &  -- \\
+\textipa{pamman} & \textipa{pammaNk\super h1m} & \textipa{pamnar1m} & \textipa{pam} & \textipa{pamtero} & \textipa{pamkwa} & \textipa{pampota} & \textipa{pamkat\super Si} & \textipa{pami} & \textipa{pam1n} & \textipa{pame} & \textipa{pamita} & \textipa{pam1ro} &  -- \\
+\textipa{pamman} & \textipa{pammaNk\super h1m} & \textipa{pannar1m} & \textipa{pat} & \textipa{pattero} & \textipa{pakkwa} & \textipa{pappota} & \textipa{pakkat\super Si} & \textipa{pat\super S\super hi} & \textipa{pat\super h1n} & \textipa{pat\super he} & \textipa{pat\super S\super hita} & \textipa{pat\super h1ro} &  -- \\
+\textipa{namman} & \textipa{nammaNk\super h1m} & \textipa{nannar1m} & \textipa{nat} & \textipa{nattero} & \textipa{nakkwa} & \textipa{nappota} & \textipa{nakkat\super Si} & \textipa{nasi} & \textipa{nas1n} & \textipa{nase} & \textipa{nasita} & \textipa{nas1ro} &  -- \\
+\textipa{namman} & \textipa{nammaNk\super h1m} & \textipa{nannar1m} & \textipa{nat} & \textipa{nattero} & \textipa{nakkwa} & \textipa{nappota} & \textipa{nakkat\super Si} & \textipa{nat\super Si} & \textipa{nat\super S1n} & \textipa{nat\super Se} & \textipa{nat\super Sita} & \textipa{nat\super S1ro} &  -- \\
+\textipa{namman} & \textipa{nammaNk\super h1m} & \textipa{nannar1m} & \textipa{nat} & \textipa{nattero} & \textipa{nakkwa} & \textipa{nappota} & \textipa{nakkat\super Si} & \textipa{nat\super S\super hi} & \textipa{nat\super S\super h1n} & \textipa{nat\super S\super he} & \textipa{nat\super S\super hita} & \textipa{nat\super S\super h1ro} &  -- \\
+\textipa{pamman} & \textipa{pammaNk\super h1m} & \textipa{pannar1m} & \textipa{pan} & \textipa{pantero} & \textipa{paNkwa} & \textipa{pampota} & \textipa{paNkat\super Si} & \textipa{pani} & \textipa{pan1n} & \textipa{pane} & \textipa{panita} & \textipa{pan1ro} &  -- \\
+\bottomrule\end{longtable}
+\begin{tabular}{l}\emph{Rules: }\\
+\textipa{t\super h} $\to$ \textipa{t\super S\super h} /  \verb|_| \textipa{i}\\\verb|[+coronal]| $\to$ $1$ /  \verb|_| \verb|[-continuant]|\\\verb|[ ]| $\to$ \verb|[-aspirated]| /  \verb|_| \{\verb|[-continuant]|,\#\}\\\textipa{p} $\to$ \textipa{m} /  \verb|_| \verb|[+nasal]|\\\verb|[+sibilant]| $\to$ \textipa{t} /  \verb|_| \#
+\end{tabular}
+\pagebreak
+
+\emph{Serbo--Croatian:}\\\begin{longtable}{llllllll|l}\toprule
+$\varnothing$ $+$stem$+$ $\varnothing$ & $\varnothing$ $+$stem$+$ \textipa{a} & $\varnothing$ $+$stem$+$ \textipa{o} & $\varnothing$ $+$stem$+$ \textipa{i} & $\varnothing$ $+$stem$+$ \textipa{\'{e}m} & $\varnothing$ $+$stem$+$ \textipa{l} & $\varnothing$ $+$stem$+$ \textipa{l\'{a}} & $\varnothing$ $+$stem$+$ \textipa{lo} & UR
+\\ \midrule
+\textipa{ml\'{a}d} & \textipa{mlad\'{a}} & \textipa{mlad\'{o}} & \textipa{mlad\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{t\'{u}p} & \textipa{tup\'{a}} & \textipa{tup\'{o}} & \textipa{tup\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{bl\'{a}g} & \textipa{blag\'{a}} & \textipa{blag\'{o}} & \textipa{blag\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{gr\'{u}b} & \textipa{grub\'{a}} & \textipa{grub\'{o}} & \textipa{grub\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{kriZan} & \textipa{kriZana} & \textipa{kriZano} & \textipa{kriZani} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{sunt\super San} & \textipa{sunt\super Sana} & \textipa{sunt\super Sano} & \textipa{sunt\super Sani} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{svet\super San} & \textipa{svet\super Sana} & \textipa{svet\super Sano} & \textipa{svet\super Sani} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{b\'{e}o} & \textipa{bel\'{a}} & \textipa{bel\'{o}} & \textipa{bel\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{veseo} & \textipa{vesela} & \textipa{veselo} & \textipa{veseli} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{deb\'{e}o} & \textipa{debel\'{a}} & \textipa{debel\'{o}} & \textipa{debel\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{m\'{i}o} & \textipa{mil\'{a}} & \textipa{mil\'{o}} & \textipa{mil\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{zel\'{e}n} & \textipa{zelen\'{a}} & \textipa{zelen\'{o}} & \textipa{zelen\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{krad\'{e}n} & \textipa{kraden\'{a}} & \textipa{kraden\'{o}} & \textipa{kraden\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{dal\'{e}k} & \textipa{dalek\'{a}} & \textipa{dalek\'{o}} & \textipa{dalek\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{vis\'{o}k} & \textipa{visok\'{a}} & \textipa{visok\'{o}} & \textipa{visok\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{dub\'{o}k} & \textipa{dubok\'{a}} & \textipa{dubok\'{o}} & \textipa{dubok\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{bogat} & \textipa{bogata} & \textipa{bogato} & \textipa{bogati} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{rapav} & \textipa{rapava} & \textipa{rapavo} & \textipa{rapavi} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{j\'{a}san} & \textipa{jasn\'{a}} & \textipa{jasn\'{o}} & \textipa{jasn\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{v\'{a}Zan} & \textipa{vaZn\'{a}} & \textipa{vaZn\'{o}} & \textipa{vaZn\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{s\'{i}tan} & \textipa{sitn\'{a}} & \textipa{sitn\'{o}} & \textipa{sitn\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{ledan} & \textipa{ledna} & \textipa{ledno} & \textipa{ledni} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{t\'{a}nak} & \textipa{tank\'{a}} & \textipa{tank\'{o}} & \textipa{tank\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{kr\'{a}tak} & \textipa{kratk\'{a}} & \textipa{kratk\'{o}} & \textipa{kratk\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{bl\'{i}zak} & \textipa{blisk\'{a}} & \textipa{blisk\'{o}} & \textipa{blisk\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{\'{u}zak} & \textipa{usk\'{a}} & \textipa{usk\'{o}} & \textipa{usk\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{d\'{o}bar} & \textipa{dobr\'{a}} & \textipa{dobr\'{o}} & \textipa{dobr\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{\'{o}Star} & \textipa{oStr\'{a}} & \textipa{oStr\'{o}} & \textipa{oStr\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{bodar} & \textipa{bodra} & \textipa{bodro} & \textipa{bodri} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{ustao} & \textipa{ustala} & \textipa{ustalo} & \textipa{ustali} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{m\'{u}kao} & \textipa{mukl\'{a}} & \textipa{mukl\'{o}} & \textipa{mukl\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{\'{o}bao} & \textipa{obl\'{a}} & \textipa{obl\'{o}} & \textipa{obl\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+\textipa{p\'{o}dao} & \textipa{podl\'{a}} & \textipa{podl\'{o}} & \textipa{podl\'{i}} &  --  &  --  &  --  &  --  &  -- \\
+ --  &  --  &  --  &  --  & \textipa{tep\'{e}m} & \textipa{t\'{e}pao} & \textipa{tepl\'{a}} & \textipa{tepl\'{o}} &  -- \\
+ --  &  --  &  --  &  --  & \textipa{skub\'{e}m} & \textipa{sk\'{u}bao} & \textipa{skubl\'{a}} & \textipa{skubl\'{o}} &  -- \\
+ --  &  --  &  --  &  --  & \textipa{tres\'{e}m} & \textipa{tr\'{e}sao} & \textipa{tresl\'{a}} & \textipa{tresl\'{o}} &  -- \\
+ --  &  --  &  --  &  --  & \textipa{vez\'{e}m} & \textipa{v\'{e}zao} & \textipa{vezl\'{a}} & \textipa{vezl\'{o}} &  -- \\
+\bottomrule\end{longtable}
+\begin{tabular}{l}\emph{Rules: }\\
+\verb|[+vowel]| $\to$ \verb|[+highTone]| / \verb|[+highTone]| \verb|[ ]|* \verb|_| \\$\varnothing$ $\to$ \textipa{a} / \verb|[-vowel]| \verb|_| \verb|[-vowel]| \#\\\verb|[+vowel]| $\to$ \verb|[-highTone]| /  \verb|_| \verb|[ ]|* \verb|[+highTone]|\\\textipa{l} $\to$ \textipa{o} /  \verb|_| \#\\\verb|[-sonorant]| $\to$ \verb|[-voice]| /  \verb|_| \verb|[-voice]|
+\end{tabular}
 

### 2018-04-29 outline.tex
heuristic description

@@
   \STATE {\bfseries Input:} Matrix of inflected forms for $N$ lexemes, $X$
+  \STATE {\bfseries Hyperparameter:} Batch size $B$
   \STATE {\bfseries Output:} underlying forms, $\{u_n\}$; prefixes $\{p_i\}$; suffixes $\{s_i\}$; rules $\{r_j\}$
-  \STATE // Initialize training set $T$ with a single example
-  \STATE Set $T = $ inflections for any lexeme in $X$
-  \STATE // Find a solution for this single example, using as many rules as you need ($\infty$)
+  \STATE // Initialize training set $T$ with $B$ examples
+  \STATE Set $T = $ inflections for $B$ lexemes in $X$
+  \STATE // Find a solution for examples, using as many rules as you need ($\infty$)
   \STATE Set $\{r_j\}_{j = 1}^{B_R}, \{p_i\}, \{s_i\} = \text{solver}(T,\infty)$
@@
   \STATE \textbf{if} $C = \varnothing$ \textbf{ then return }$\{r_j\}_{j = 1}^{B_R},\{p_i\}, \{s_i\}$
-
+  \STATE // Add a batch of counterexamples to the training set
+  \STATE $T\gets T\cup \{C_j\}_{j = 1}^B$
   \STATE // Start out with a search radius of $1$ and increase it until we find a way of accommodating another example
   \FOR{$R = 1,2,3,4,...$}
-  \FOR {$c\in C$}
-  \STATE Set synthesisResult$ = \text{solveIncrementalChange}(T\cup\{c\},\{r_j\},R)$
+%  \FOR {$c\in C$}
+  \STATE Set synthesisResult$ = \text{solveIncrementalChange}(T,\{r_j\},R)$
   \IF{synthesisResult$\not= \bot$}
   \STATE Set $\{r_j\}_{j = 1}^{B_R}, \{p_i\}, \{s_i\} = $synthesisResult
-  \STATE Add $c$ to the set of training examples, $T$
-  \STATE \textbf{break out of loop over }$R$ // Successfully updated both our solution and added another example to the training set
+  \STATE \textbf{break out of loop over }$R$ // Successfully updated both our solution and the training set
 \end{algorithmic}
@@
 \end{itemize}
+
+\subsection{Search heuristics}
+
+\subsubsection{Guessing underlying forms}
+
+\textbf{incrementalCEGIS} is a generic way of learning large programs that explain large data sets,
+and does not encode any domain specific knowledge about phonology.
+Empirically, the search problem is \emph{much} easier if you
+either know what the underlying forms are or have some constraints on what the underlying forms are.
+If I introspect on my own reasoning process while solving these problems,
+I usually guess at most a few plausible underlying forms for each lexeme.
+I encoded a simple heuristic for guessing parts of underlying representations:
+given prefixes/suffixes $\left\{p_i \right\}$ \& $\left\{s_i \right\}$
+and observed inflections for a single lexeme $\left\{x_i \right\}$,
+this heuristic guesses that the underlying form begins with the longest shared prefix
+of the surface forms with the prefixes chopped off,
+and that it ends with the longest shared suffix of the surface forms with the suffixes chopped off.
+For example, if we have surface forms $(\text{test},\text{testing},\text{tests})$,
+and have prefixes $(\text{t},\text{t},\text{t})$ and suffixes $(\text{t},\text{ting},\text{ts})$,
+this heuristic constrains the underlying form to start with $\text{es}$ and end with $\text{es}$.
 

### 2018-04-30 test.tex
updated problem solutions

@@
 \end{tabular}
+\pagebreak
+
+\emph{Standard Ukrainian:}\\\begin{longtable}{llllllllll|l}\toprule
+$\varnothing$ $+$stem$+$ $\varnothing$ & $\varnothing$ $+$stem$+$ \textipa{am} & $\varnothing$ $+$stem$+$ \textipa{ov\super ji} & $\varnothing$ $+$stem$+$ \textipa{i} & $\varnothing$ $+$stem$+$ \textipa{ov\super ji} & $\varnothing$ $+$stem$+$ \textipa{o} & $\varnothing$ $+$stem$+$ \textipa{a} & $\varnothing$ $+$stem$+$ \textipa{u} & $\varnothing$ $+$stem$+$ \textipa{i} & $\varnothing$ $+$stem$+$ $\varnothing$ & UR
+\\ \midrule
+\textipa{zub} & \textipa{zubam} & \textipa{zubov\super ji} & \textipa{zub\super ji} &  --  &  --  &  --  &  --  &  --  &  --  &  -- \\
+\textipa{sv\super jit} & \textipa{sv\super jitam} & \textipa{sv\super jitov\super ji} & \textipa{sv\super jit\super ji} &  --  &  --  &  --  &  --  &  --  &  --  &  -- \\
+\textipa{z\super jat\super j} & \textipa{z\super jat\super jam} & \textipa{z\super jatev\super ji} &  --  & \textipa{z\super jatev\super ji} &  --  &  --  &  --  &  --  &  --  &  -- \\
+\textipa{koS\super jil\super j} & \textipa{koSel\super jam} & \textipa{koSelev\super ji} & \textipa{koSel\super ji} &  --  &  --  &  --  &  --  &  --  &  --  &  -- \\
+\textipa{zlod\super jij} & \textipa{zlod\super jijam} & \textipa{zlod\super jijev\super ji} &  --  & \textipa{zlod\super jijev\super ji} &  --  &  --  &  --  &  --  &  --  &  -- \\
+\textipa{m\super jis\super jar\super s \super j} & \textipa{m\super jis\super jar\super s \super jam} & \textipa{m\super jis\super jat\super sev\super ji} & \textipa{m\super jis\super jar\super s \super ji} &  --  &  --  &  --  &  --  &  --  &  --  &  -- \\
+\textipa{korovaj} & \textipa{korovajam} & \textipa{korovajev\super ji} & \textipa{korovaji} &  --  &  --  &  --  &  --  &  --  &  --  &  -- \\
+\textipa{kam\super jin\super j} & \textipa{kamen\super jam} & \textipa{kamenev\super ji} & \textipa{kamen\super ji} &  --  &  --  &  --  &  --  &  --  &  --  &  -- \\
+\textipa{m\super jid\super j} & \textipa{m\super jid\super jam} & \textipa{m\super jidev\super ji} & \textipa{m\super jid\super ji} &  --  &  --  &  --  &  --  &  --  &  --  &  -- \\
+\textipa{xl\super jiw} & \textipa{xl\super jivam} & \textipa{xl\super jivov\super ji} & \textipa{xl\super jiv\super ji} &  --  &  --  &  --  &  --  &  --  &  --  &  -- \\
+\textipa{holub} & \textipa{holubam} & \textipa{holubov\super ji} & \textipa{holub\super ji} &  --  &  --  &  --  &  --  &  --  &  --  &  -- \\
+\textipa{s\super jin} & \textipa{s\super jinam} & \textipa{s\super jinov\super ji} & \textipa{s\super jin\super ji} &  --  &  --  &  --  &  --  &  --  &  --  &  -- \\
+\textipa{leb\super jid\super j} & \textipa{lebed\super jam} & \textipa{lebedev\super ji} & \textipa{lebed\super ji} &  --  &  --  &  --  &  --  &  --  &  --  &  -- \\
+\textipa{sus\super jid} & \textipa{sus\super jidam} & \textipa{sus\super jidov\super ji} &  --  & \textipa{sus\super jidov\super ji} &  --  &  --  &  --  &  --  &  --  &  -- \\
+\textipa{t\super Solov\super jik} & \textipa{t\super Solov\super jikam} & \textipa{t\super Solov\super jikov\super ji} &  --  & \textipa{t\super Solov\super jikov\super ji} &  --  &  --  &  --  &  --  &  --  &  -- \\
+\textipa{l\super jid} & \textipa{ledam} & \textipa{ledov\super ji} & \textipa{led\super ji} &  --  &  --  &  --  &  --  &  --  &  --  &  -- \\
+\textipa{bil\super j} & \textipa{bol\super jam} & \textipa{bolev\super ji} & \textipa{bol\super ji} &  --  &  --  &  --  &  --  &  --  &  --  &  -- \\
+\textipa{riw} & \textipa{rovam} & \textipa{rovov\super ji} & \textipa{rov\super ji} &  --  &  --  &  --  &  --  &  --  &  --  &  -- \\
+\textipa{stiw} & \textipa{stolam} & \textipa{stolov\super ji} & \textipa{stol\super ji} &  --  &  --  &  --  &  --  &  --  &  --  &  -- \\
+\textipa{d\super jid} & \textipa{d\super jidam} & \textipa{d\super jidov\super ji} &  --  & \textipa{d\super jidov\super ji} &  --  &  --  &  --  &  --  &  --  &  -- \\
+\textipa{l\super jit} & \textipa{l\super jotam} & \textipa{l\super jotov\super ji} & \textipa{l\super jot\super ji} &  --  &  --  &  --  &  --  &  --  &  --  &  -- \\
+\textipa{mist} & \textipa{mostam} & \textipa{mostov\super ji} & \textipa{most\super ji} &  --  &  --  &  --  &  --  &  --  &  --  &  -- \\
+\textipa{vet\super Sir} & \textipa{vet\super Soram} & \textipa{vet\super Sorov\super ji} & \textipa{vet\super Sor\super ji} &  --  &  --  &  --  &  --  &  --  &  --  &  -- \\
+ --  &  --  &  --  &  --  &  --  & \textipa{t\super jilo} & \textipa{t\super jila} & \textipa{t\super jilu} & \textipa{t\super jil\super ji} & \textipa{t\super jiw} &  -- \\
+ --  &  --  &  --  &  --  &  --  & \textipa{koleso} & \textipa{kolesa} & \textipa{kolesu} & \textipa{koles\super ji} & \textipa{kol\super jis} &  -- \\
+ --  &  --  &  --  &  --  &  --  & \textipa{ozero} & \textipa{ozera} & \textipa{ozeru} & \textipa{ozer\super ji} & \textipa{oz\super jir} &  -- \\
+ --  &  --  &  --  &  --  &  --  & \textipa{selo} & \textipa{sela} & \textipa{selu} & \textipa{sel\super ji} & \textipa{s\super jiw} &  -- \\
+ --  &  --  &  --  &  --  &  --  & \textipa{pole} & \textipa{pol\super ja} & \textipa{pol\super ju} & \textipa{pol\super ji} & \textipa{pil\super j} &  -- \\
+ --  &  --  &  --  &  --  &  --  & \textipa{slovo} & \textipa{slova} & \textipa{slovu} & \textipa{slov\super ji} & \textipa{sliw} &  -- \\
+ --  &  --  &  --  &  --  &  --  & \textipa{more} & \textipa{mor\super ja} & \textipa{mor\super ju} & \textipa{mor\super ji} & \textipa{mir\super j} &  -- \\
+\bottomrule\end{longtable}
+\begin{tabular}{l}\emph{Rules: }\\
+\verb|[ ]| $\to$ $-2$ / \verb|[ ]| \verb|[ ]| \verb|_| \verb|[+sibilant]| \#\\\verb|[ ]| $\to$ \textipa{e} / \verb|[+palletized]| \verb|_| \verb|[+voice]| \verb|[+vowel]|\\\verb|[-glide]| $\to$ \verb|[-palletized]| /  \verb|_| \verb|[-back]|\\\textipa{w} $\to$ \textipa{v} / \verb|[+vowel]| \verb|_| \verb|[+vowel]|\\\verb|[ ]| $\to$ \verb|[+palletized]| /  \verb|_| \textipa{i}\\\textipa{o} $\to$ \textipa{i} /  \verb|_| \verb|[-vowel]|* \verb|[-vowel]| \#\\\textipa{l} $\to$ \textipa{w} /  \verb|_| \#
+\end{tabular}
+
+\pagebreak
+
+\emph{Latin:}\\\begin{longtable}{lll|l}\toprule
+$\varnothing$ $+$stem$+$ \textipa{s} & $\varnothing$ $+$stem$+$ \textipa{is} & $\varnothing$ $+$stem$+$ \textipa{i:} & UR
+\\ \midrule
+\textipa{arks} & \textipa{arkis} &  --  &  -- \\
+\textipa{duks} & \textipa{dukis} &  --  &  -- \\
+\textipa{daps} & \textipa{dapis} &  --  &  -- \\
+\textipa{re:ks} & \textipa{re:gis} &  --  &  -- \\
+\textipa{falanks} & \textipa{falangis} &  --  &  -- \\
+\textipa{filiks} & \textipa{filikis} &  --  &  -- \\
+\textipa{lapis} & \textipa{lapidis} &  --  &  -- \\
+\textipa{li:s} & \textipa{li:tis} &  --  &  -- \\
+\textipa{fraws} & \textipa{frawdis} &  --  &  -- \\
+\textipa{noks} & \textipa{noktis} &  --  &  -- \\
+\textipa{frons} & \textipa{frontis} &  --  &  -- \\
+\textipa{frons} & \textipa{frondis} &  --  &  -- \\
+\textipa{inku:s} & \textipa{inku:dis} &  --  &  -- \\
+\textipa{sors} & \textipa{sortis} &  --  &  -- \\
+\textipa{fu:r} & \textipa{fu:ris} &  --  &  -- \\
+\textipa{murmur} & \textipa{murmuris} &  --  &  -- \\
+\textipa{augur} & \textipa{auguris} &  --  &  -- \\
+\textipa{arbor} & \textipa{arboris} &  --  &  -- \\
+\textipa{pugil} & \textipa{pugilis} &  --  &  -- \\
+\textipa{sal} & \textipa{salis} &  --  &  -- \\
+\textipa{adeps} & \textipa{adipis} &  --  &  -- \\
+\textipa{apeks} & \textipa{apikis} &  --  &  -- \\
+\textipa{pri:nkeps} & \textipa{pri:nkipis} &  --  &  -- \\
+\textipa{ekwes} & \textipa{ekwitis} &  --  &  -- \\
+\textipa{miles} & \textipa{militis} &  --  &  -- \\
+\textipa{no:men} & \textipa{no:minis} &  --  &  -- \\
+\textipa{karmen} & \textipa{karminis} &  --  &  -- \\
+\textipa{lu:men} & \textipa{lu:minis} &  --  &  -- \\
+\textipa{wenter} & \textipa{wentris} &  --  &  -- \\
+\textipa{pater} & \textipa{patris} &  --  &  -- \\
+\textipa{kada:wer} & \textipa{kada:weris} &  --  &  -- \\
+\textipa{tu:ber} & \textipa{tu:beris} &  --  &  -- \\
+\textipa{piper} & \textipa{piperis} &  --  &  -- \\
+\textipa{karker} & \textipa{karkeris} &  --  &  -- \\
+\textipa{die:s} &  --  & \textipa{die:i:} &  -- \\
+\textipa{li:ber} &  --  & \textipa{li:beri:} &  -- \\
+\textipa{miser} &  --  & \textipa{miseri:} &  -- \\
+\textipa{ager} &  --  & \textipa{agri:} &  -- \\
+\textipa{sinister} &  --  & \textipa{sinistri:} &  -- \\
+\textipa{liber} &  --  & \textipa{libri:} &  -- \\
+\textipa{as} & \textipa{assis} &  --  &  -- \\
+\textipa{os} & \textipa{ossis} &  --  &  -- \\
+\textipa{far} & \textipa{farris} &  --  &  -- \\
+\textipa{mel} & \textipa{mellis} &  --  &  -- \\
+\textipa{o:s} & \textipa{o:ris} &  --  &  -- \\
+\textipa{flo:s} & \textipa{flo:ris} &  --  &  -- \\
+\textipa{mu:s} & \textipa{mu:ris} &  --  &  -- \\
+\textipa{cru:s} & \textipa{cru:ris} &  --  &  -- \\
+\textipa{kinis} & \textipa{kineris} &  --  &  -- \\
+\textipa{pulvis} & \textipa{pulveris} &  --  &  -- \\
+\bottomrule\end{longtable}
+\begin{tabular}{l}\emph{Rules: }\\
+\verb|[-sonorant]| $\to$ \verb|[-voice]| /  \verb|_| \verb|[-voice]|\\\textipa{e} $\to$ $2$ /  \verb|_| \verb|[-liquid]| \verb|[+vowel]|\\\verb|[-vowel]| $\to$ $\varnothing$ / \verb|[+coronal]| \verb|_| \#\\\verb|[-sonorant]| $\to$ \textipa{s} /  \verb|_| \#\\$\varnothing$ $\to$ \textipa{e} / \verb|[-sonorant]| \verb|_| \verb|[+voice]| \#\\$\varnothing$ $\to$ $2$ / \verb|[+sibilant]| \verb|_| \verb|[+vowel]| \verb|[-voice]|\\$\varnothing$ $\to$ $1$ / \# \verb|[-coronal]| \verb|[-high]| \verb|_| \verb|[+voice]| \textipa{i}
+\end{tabular}
 

### 2018-05-01 test.tex
textbook solutions

@@
 \end{tabular}
+\begin{tabular}{l}\emph{Textbook solution rules: }\\
+\verb|[+vowel]| $\to$ \verb|[+highTone]| / \verb|[+highTone]| \verb|[-vowel]|* \verb|_| \\\verb|[+vowel]| $\to$ \verb|[-highTone]| /  \verb|_| \verb|[-vowel]|* \verb|[+highTone]|\\$\varnothing$ $\to$ \textipa{a} / \verb|[-vowel]| \verb|_| \verb|[-vowel]| \#\\\textipa{l} $\to$ \textipa{o} /  \verb|_| \#\\\verb|[-sonorant]| $\to$ \verb|[-voice]| /  \verb|_| \verb|[-voice]|
+\end{tabular}
+
 \pagebreak
@@
 \verb|[ ]| $\to$ $-2$ / \verb|[ ]| \verb|[ ]| \verb|_| \verb|[+sibilant]| \#\\\verb|[ ]| $\to$ \textipa{e} / \verb|[+palletized]| \verb|_| \verb|[+voice]| \verb|[+vowel]|\\\verb|[-glide]| $\to$ \verb|[-palletized]| /  \verb|_| \verb|[-back]|\\\textipa{w} $\to$ \textipa{v} / \verb|[+vowel]| \verb|_| \verb|[+vowel]|\\\verb|[ ]| $\to$ \verb|[+palletized]| /  \verb|_| \textipa{i}\\\textipa{o} $\to$ \textipa{i} /  \verb|_| \verb|[-vowel]|* \verb|[-vowel]| \#\\\textipa{l} $\to$ \textipa{w} /  \verb|_| \#
+\end{tabular}
+\begin{tabular}{l}\emph{Textbook solution rules: }\\
+\verb|[-vowel]| $\to$ \verb|[+palletized]| /  \verb|_| \textipa{i}\\\textipa{o} $\to$ \textipa{e} / \verb|[+palletized]| $+$ \verb|_| \\\verb|[+vowel -low -high]| $\to$ \textipa{i} /  \verb|_| \verb|[-vowel]|* \verb|[-vowel]| \#\\\verb|[-glide -vowel]| $\to$ \verb|[-palletized]| /  \verb|_| \textipa{e}\\\textipa{l} $\to$ \textipa{w} /  \verb|_| \#\\\textipa{v} $\to$ \textipa{w} /  \verb|_| \#
 \end{tabular}
@@
 \end{tabular}
+\begin{tabular}{l}\emph{Textbook solution rules: }\\
+\verb|[-sonorant]| $\to$ \verb|[-voice]| /  \verb|_| \verb|[-voice]|\\\textipa{s} $\to$ $\varnothing$ / \verb|[+coronal +sonorant]| \verb|_| \\\verb|[+coronal -sonorant]| $\to$ $\varnothing$ /  \verb|_| \textipa{s}\\\textipa{e} $\to$ \textipa{i} /  \verb|_| \verb|[-vowel]| \verb|[+vowel]|\\$\varnothing$ $\to$ \textipa{e} / \verb|[-vowel]| \verb|_| \verb|[+coronal +sonorant]| \#\\\textipa{s} $\to$ \textipa{r} / \verb|[+vowel]| \verb|_| \verb|[+vowel]|\\\verb|[+vowel -back]| $\to$ \verb|[-high]| /  \verb|_| \textipa{r}\\$1$ $\to$ $\varnothing$ /  \verb|_| \verb|[-vowel]| \#
+\end{tabular}
 

### 2018-06-16 manuscript.tex
science paper progress

% Use only LaTeX2e, calling the article.cls class and 12-point type.

\documentclass[12pt]{article}

% Users of the {thebibliography} environment or BibTeX should use the
% scicite.sty package, downloadable from *Science* at
% http://www.sciencemag.org/authors/preparing-manuscripts-using-latex 
% This package should properly format in-text
% reference calls and reference-list numbers.

\usepackage{scicite}

\usepackage{times}

% The preamble here sets up a lot of new/revised commands and
% environments.  It's annoying, but please do *not* try to strip these
% out into a separate .sty file (which could lead to the loss of some
% information when we convert the file to other formats).  Instead, keep
% them in the preamble of your main LaTeX source file.

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{tipa}
\usepackage{graphicx}
\usepackage{booktabs}
% The following parameters seem to provide a reasonable page setup.

\topmargin 0.0cm
\oddsidemargin 0.2cm
\textwidth 16cm 
\textheight 21cm
\footskip 1.0cm

\usepackage{phonrule}

%The next command sets up an environment for the abstract to your paper.

\newenvironment{sciabstract}{%
\begin{quote} \bf}
{\end{quote}}

% Include your paper's title here

\title{Synthesizing Theories of Human Language with Bayesian Program Induction}

% Place the author information here.  Please hand-code the contact
% information and notecalls; do *not* use \footnote commands.  Let the
% author contact information appear immediately below the author names
% as shown.  We would also prefer that you don't change the type-size
% settings shown here.

\author
{John Smith,$^{1\ast}$ Jane Doe,$^{1}$ Joe Scientist$^{2}$\\
\\
\normalsize{$^{1}$Department of Chemistry, University of Wherever,}\\
\normalsize{An Unknown Address, Wherever, ST 00000, USA}\\
\normalsize{$^{2}$Another Unknown Address, Palookaville, ST 99999, USA}\\
\\
\normalsize{$^\ast$To whom correspondence should be addressed; E-mail:  jsmith@wherever.edu.}
}

% Include the date command, but leave its argument blank.

\date{}

%%%%%%%%%%%%%%%%% END OF PREAMBLE %%%%%%%%%%%%%%%%

\begin{document}

% Double-space the manuscript.

\baselineskip24pt

% Make the title.

\maketitle

% Place your abstract within the special {sciabstract} environment.

\begin{sciabstract}
\end{sciabstract}

% In setting up this template for *Science* papers, we've used both
% the \section* command and the \paragraph* command for topical
% divisions.  Which you use will of course depend on the type of paper
% you're writing.  Review Articles tend to have displayed headings, for
% which \section* is more appropriate; Research Articles, when they have
% formal topical divisions at all, tend to signal them with bold text
% that runs into the paragraph, for which \paragraph* is the right
% choice.  Either way, use the asterisk (*) modifier, as shown, to
% suppress numbering.

\section*{Introduction}

An age-old aspiration within artificial intelligence research is to build a machine that automates the scientific process~\cite{paul1990autonomous,langley1987scientific}, either by designing experiments~\cite{sparkes2010towards} or
synthesizing theories and models~\cite{schmidt2009distilling}.
However, gaining traction on the problem of theory induction requires
a spectrum of tractable theory synthesis problems:
while early work attempted to recapitulate
classic discoveries like Kepler's laws~\cite{BACON},
these results proved brittle and difficult to generalize.
Despite these early works,
theory induction has never attained the same level of success as
machine vision or natural language processing.
How can the artificial intelligence community get theory induction off the ground?
This is an especially difficult question,
because the current mainstream in machine intelligence focuses on
learning to solve a qualitatively different class of problems (e.g., prediction tasks like classification and regression),
whereas theory induction requires
synthesizing human-understandable causal models of real-world phenomena~\cite{pearl2009causality},
so that human scientists can understand and learn from the AI's outputs.

We propose theory induction research start with theories of human
language, and scope out a collection of problems and techniques for a
key module of natural language: \emph{morphophonology}, the
relationship between pronunciations of words and their meaning.
Acquiring the morphophonology of a language involve solving a basic
problem that confronts both linguists and children: given a collection
of utterances, together with aspects of their meaning, what is the
causal relationship between form and meaning?  Morphophonology, and
natural language more broadly, presents a tractable challenge to
theory induction, for three reasons.  First, every natural language is
a different theory induction problem, so there is a wide spectrum of
induction tasks to evaluate on.  Second, linguistics as a field
traditionally formalizes theories in computational terms, and so there
is a pre-existing body of representations and algorithms for the AI
researcher to draw upon.  Third, we know that acquiring theories of
language (e.g., grammars) is tractable, because children 
acquire language, and field linguists can be trained to build theories of grammar.

Our contribution to theory induction is
an algorithm for synthesizing
theories of natural language morphophonology.
Like the linguist,
the algorithm starts with
a collection of utterances 
annotated with their meanings,
and then constructs a
causal, interpretable model explaining how
those meanings gave rise to the utterances.
An important aspect of a scientific theory is
that it makes predictions; here,
the theory predicts pronunciations of unobserved words conditioned on their meaning.
We evaluate our algorithm on \# data sets spanning \# languages,
automatically finding theories that can model a wide swath of a core component of human natural language.

\begin{table}\centering
\begin{tabular}{lcl}
  \toprule
  Example data&Morphophonology\\\midrule
\begin{tabular}{ll}
1&\textipa{\|x{j}ig}\\
4&\textipa{\|x{s}i}\\
5&\textipa{Na}\\
9&\textipa{gu}\\
10&\textipa{\|x{j}u}\\
11 ($= 10 + 1$) & \textipa{\|x{j}ug\|x{j}ig}\\
14 ($= 10 + 4$) & \textipa{\|x{j}ub\|x{s}i}\\
15 ($= 10 + 5$) & \textipa{\|x{j}uNa}\\
19 ($= 10 + 9$) & \textipa{\|x{j}urgu}\\
40 ($= 4 + 10$) & \textipa{\|x{s}ib\|x{j}u}\\
50 ($= 5 + 10$) & \textipa{Nab\|x{j}u}\\
90 ($= 9 + 10$) & \textipa{gub\|x{j}u}
\end{tabular}
&
\begin{tabular}{l}
  \emph{Consonant cluster reduction:}\\
  \phonb{C}{$\varnothing$}{\#}{C}\\
  (\emph{Upon encountering a consonant (C),}\\
  \emph{ delete it ($\to\varnothing$) at the beginning of}\\
  \emph{ a word (\#\_) if followed by another}\\
  \emph{ consonant (\_C)})
  \end{tabular}

  \\
  \bottomrule  \end{tabular}
\caption{Tibetan count system.}
\label{Tibetan}
\end{table}

\begin{figure}
  \includegraphics[width = \textwidth]{naturalLanguageGenerativeModel.pdf}
  \caption{Theory learner induces grammars (blue) for a range of languages, given observed form/meaning pairs (pink). Grammars are expressed as programs drawn from a universal grammar, or programming language. }\label{generativeModel}
  \end{figure}

\section*{Discovering Theories by Synthesizing Programs}

We frame our approach as Bayesian Program Learning (BPL: see~\cite{lake2015human}),
where the agent explains a
set of utterance/meaning pairs $\left\{(u_n,m_n) \right\}_{n = 1}^N$
by inferring a theory $T$, which we model as a program.
Formalizing grammars (theories) as generative programs has a long history in linguistics~\cite{chomsky1968sound}.
Written as a probabilistic inference problem, the agent seeks the $T$ maximizing $\left[\prod_{n = 1}^N P(u_n,m_n | T) \right] P(T|UG)$,
where $UG$ is a ``universal grammar'' (\textbf{TIM}: what should I cite here?)
constraining the space of allowed theories and imparting an inductive bias over the space of grammars.
In this BPL setting we model $UG$ as a prior distribution over theories (programs).
To model the space of programs
use Context-Sensitive Rewrites,
a Turing-complete program representation,
but restrict the rewrites
in such a way as to make them equivalent to finite state transducers --- this insight comes from the computational linguistics literature~\cite{beesley2003finite}.
For morphophonology,
the relationship between meaning $m_n$ and utterance $u_n$
involves a latent \emph{lexicon},
which is a mapping between lexemes and the pronunciation of their stem.
For children, linguists, and our model, the lexicon
must also be inferred (Fig.~\ref{generativeModel}).

Although this framing captures the problem a BPL theory inductor needs to solve,
it offers no guidance on how to solve that problem:
the space of all programs (theories) is infinitely large and
sharply discontinuous, lacking the local smoothness that
enables local optimization algorithms (e.g., gradient descent; MCMC)
to succeed.
We adopt a strategy currently unconventional to the
AI community: constraint--based program synthesis,
where the optimization problem is translated
into a combinatorial constraint satisfaction 
problem and solved using
a SMT solver~\cite{solar2006combinatorial}.

\section*{Discussion}

\bibliography{main}

\bibliographystyle{Science}

%% \section*{Acknowledgments}
%% Include acknowledgments of funding, any patents pending, where raw data for the paper are deposited, etc.

%% %Here you should list the contents of your Supplementary Materials -- below is an example. 
%% %You should include a list of Supplementary figures, Tables, and any references that appear only in the SM. 
%% %Note that the reference numbering continues from the main text to the SM.
%% % In the example below, Refs. 4-10 were cited only in the SM.     
%% \section*{Supplementary materials}
%% Materials and Methods\\
%% Supplementary Text\\
%% Figs. S1 to S3\\
%% Tables S1 to S4\\
%% References \textit{(4-10)}

%% % For your review copy (i.e., the file you initially send in for
%% % evaluation), you can use the {figure} environment and the
%% % \includegraphics command to stream your figures into the text, placing
%% % all figures at the end.  For the final, revised manuscript for
%% % acceptance and production, however, PostScript or other graphics
%% % should not be streamed into your compliled file.  Instead, set
%% % captions as simple paragraphs (with a \noindent tag), setting them
%% % off from the rest of the text with a \clearpage as shown  below, and
%% % submit figures as separate files according to the Art Department's
%% % instructions.

%% \clearpage

%% \noindent {\bf Fig. 1.} Please do not use figure environments to set
%% up your figures in the final (post-peer-review) draft, do not include graphics in your
%% source code, and do not cite figures in the text using \LaTeX\
%% \verb+\ref+ commands.  Instead, simply refer to the figure numbers in
%% the text per {\it Science\/} style, and include the list of captions at
%% the end of the document, coded as ordinary paragraphs as shown in the
%% \texttt{scifile.tex} template file.  Your actual figure files should
%% be submitted separately.

\end{document}

### 2018-06-16 outline.tex
science paper progress

@@
 \subsection{Search heuristics}
+
+\subsubsection{Bounding the size of the rules}
+Sketch only works with finite program spaces, because it unrolls anything that loops or is recursive up to some depth bound.
+A necessary consequence of this is that we cannot learn rules with an arbitrary number of feature matrices.
+I bound the number of feature matrices to be 2 on each side.
+
+\subsubsection{``Freezing'' rules and underlying forms}
+\textbf{incrementalCEGIS} constructs a sequence of models that explain
+more and more of the data.  If in this sequence we see that a certain
+rule always occurs, or that a certain surface form always has a
+certain underlying representation, then we might guess that rule or
+underlying representation is really part of the correct solution and
+that we shouldn't go back and try to revise that rule or underlying
+form.  Concretely what this means is that I keep track of how many
+times a rule or underlying form is found by \textbf{incrementalCEGIS};
+once it has occurred at least 5 times I ``freeze'' the rule or
+underlying form and do not allow it to change in the future.
+
+For example, after a few rounds of \textbf{incrementalCEGIS}
+we might find that we always need to devoice word final segments.
+If this is the case then we will stop trying to change the rule that devoices
+word final segments.
 
@@
 \item Serbo-Croatian: predictable stress, devoicing, neutralizing, epenthesis
+\item Catalan: Spritization, devoicing, several different kinds of cluster reduction
+\item Korean: place assimilation and nasalization
+\item Latin: assimilation, devoicing, different kinds of cluster reduction, vowel harmony
 \end{itemize}
-17 textbook problems are not solved.
-2 of these need a tier-based representation.
-A couple more of these are due to limitations in the program learner's representation of rules.
-The rest are because they involve scaling to more than 3 rules, which is about the practical limit for exact search with these methods.
+These problems it solves most of:
+\begin{itemize}
+\item Somali:  Spritization, copying rules
+\end{itemize}
+12 textbook problems are not solved, along with two other problems that involve tier--based representations.
+3 of the above problems are not actually exercises but are instead examples that I included because they are neat problems. This means that we solve $\frac{28}{40} = 70\%$ of the textbook currently,
+using a slightly liberal interpretation of what it means to ``solve a problem'' (e.g., our solution for Latin is slightly different from the textbook answer, but to me it looks fine and is at least as compressive).
 \pagebreak

### 2018-06-16 test.tex
science paper progress

--- 2018-04-30 test.tex
@@
 \usepackage[margin = 0.1cm]{geometry}
+\usepackage{phonrule}
 \usepackage{tipa}
@@
 \begin{tabular}{l}\emph{Rules: }\\
-\verb|[+vowel]| $\to$ \verb|[+highTone]| / \verb|[+highTone]| \verb|[ ]|* \verb|_| \\$\varnothing$ $\to$ \textipa{a} / \verb|[-vowel]| \verb|_| \verb|[-vowel]| \#\\\verb|[+vowel]| $\to$ \verb|[-highTone]| /  \verb|_| \verb|[ ]|* \verb|[+highTone]|\\\textipa{l} $\to$ \textipa{o} /  \verb|_| \#\\\verb|[-sonorant]| $\to$ \verb|[-voice]| /  \verb|_| \verb|[-voice]|
-\end{tabular}
+\phonl{V}{[ +highTone ]}{[ +highTone ] [ ]\textsubscript{*}}\\\phonb{$\varnothing$}{\textipa{a}}{C}{C \#}\\\phonr{V}{[ -highTone ]}{[ ]\textsubscript{*} [ +highTone ]}\\\phonr{\textipa{l}}{\textipa{o}}{\#}\\\phonr{[ -sonorant ]}{[ -voice ]}{[ -voice ]}
+\end{tabular}
+\begin{tabular}{l}\emph{Textbook solution rules: }\\
+\phonl{V}{[ +highTone ]}{[ +highTone ] C\textsubscript{*}}\\\phonr{V}{[ -highTone ]}{C\textsubscript{*} [ +highTone ]}\\\phonb{$\varnothing$}{\textipa{a}}{C}{C \#}\\\phonr{\textipa{l}}{\textipa{o}}{\#}\\\phonr{[ -sonorant ]}{[ -voice ]}{[ -voice ]}
+\end{tabular}
+
 \pagebreak
@@
 \begin{tabular}{l}\emph{Rules: }\\
-\verb|[ ]| $\to$ $-2$ / \verb|[ ]| \verb|[ ]| \verb|_| \verb|[+sibilant]| \#\\\verb|[ ]| $\to$ \textipa{e} / \verb|[+palletized]| \verb|_| \verb|[+voice]| \verb|[+vowel]|\\\verb|[-glide]| $\to$ \verb|[-palletized]| /  \verb|_| \verb|[-back]|\\\textipa{w} $\to$ \textipa{v} / \verb|[+vowel]| \verb|_| \verb|[+vowel]|\\\verb|[ ]| $\to$ \verb|[+palletized]| /  \verb|_| \textipa{i}\\\textipa{o} $\to$ \textipa{i} /  \verb|_| \verb|[-vowel]|* \verb|[-vowel]| \#\\\textipa{l} $\to$ \textipa{w} /  \verb|_| \#
+\phonb{[ ]}{[ ]\textsubscript{i}}{[ ]\textsubscript{i} [ ]}{[ +sibilant ] \#}\\\phonb{[ ]}{\textipa{e}}{[ +palletized ]}{[ +voice ] V}\\\phonr{[ -glide ]}{[ -palletized ]}{[ -back ]}\\\phonb{\textipa{w}}{\textipa{v}}{V}{V}\\\phonr{[ ]}{[ +palletized ]}{\textipa{i}}\\\phonr{\textipa{o}}{\textipa{i}}{C\textsubscript{*} C \#}\\\phonr{\textipa{l}}{\textipa{w}}{\#}
+\end{tabular}
+\begin{tabular}{l}\emph{Textbook solution rules: }\\
+\phonr{C}{[ +palletized ]}{\textipa{i}}\\\phonl{\textipa{o}}{\textipa{e}}{[ +palletized ] $+$}\\\phonr{[ +vowel -low -high ]}{\textipa{i}}{C\textsubscript{*} C \#}\\\phonr{[ -glide -vowel ]}{[ -palletized ]}{\textipa{e}}\\\phonr{\textipa{l}}{\textipa{w}}{\#}\\\phonr{\textipa{v}}{\textipa{w}}{\#}
 \end{tabular}
@@
 \begin{tabular}{l}\emph{Rules: }\\
-\verb|[-sonorant]| $\to$ \verb|[-voice]| /  \verb|_| \verb|[-voice]|\\\textipa{e} $\to$ $2$ /  \verb|_| \verb|[-liquid]| \verb|[+vowel]|\\\verb|[-vowel]| $\to$ $\varnothing$ / \verb|[+coronal]| \verb|_| \#\\\verb|[-sonorant]| $\to$ \textipa{s} /  \verb|_| \#\\$\varnothing$ $\to$ \textipa{e} / \verb|[-sonorant]| \verb|_| \verb|[+voice]| \#\\$\varnothing$ $\to$ $2$ / \verb|[+sibilant]| \verb|_| \verb|[+vowel]| \verb|[-voice]|\\$\varnothing$ $\to$ $1$ / \# \verb|[-coronal]| \verb|[-high]| \verb|_| \verb|[+voice]| \textipa{i}
+\phonr{[ -sonorant ]}{[ -voice ]}{[ -voice ]}\\\phonr{\textipa{e}}{V\textsubscript{i}}{[ -liquid ] V\textsubscript{i}}\\\phonb{C}{$\varnothing$}{[ +coronal ]}{\#}\\\phonr{[ -sonorant ]}{\textipa{s}}{\#}\\\phonb{$\varnothing$}{\textipa{e}}{[ -sonorant ]}{[ +voice ] \#}\\\phonb{$\varnothing$}{[ -voice ]\textsubscript{i}}{[ +sibilant ]}{V [ -voice ]\textsubscript{i}}\\\phonb{$\varnothing$}{[ +voice ]\textsubscript{i}}{\# [ -coronal ] [ -high ]}{[ +voice ]\textsubscript{i} \textipa{i}}
+\end{tabular}
+\begin{tabular}{l}\emph{Textbook solution rules: }\\
+\phonr{[ -sonorant ]}{[ -voice ]}{[ -voice ]}\\\phonl{\textipa{s}}{$\varnothing$}{[ +coronal +sonorant ]}\\\phonr{[ +coronal -sonorant ]}{$\varnothing$}{\textipa{s}}\\\phonr{\textipa{e}}{\textipa{i}}{C V}\\\phonb{$\varnothing$}{\textipa{e}}{C}{[ +coronal +sonorant ] \#}\\\phonb{\textipa{s}}{\textipa{r}}{V}{V}\\\phonr{[ +vowel -back ]}{[ -high ]}{\textipa{r}}\\\phonr{$1$}{$\varnothing$}{C \#}
 \end{tabular}

### 2018-06-18 manuscript.tex
progress on science - still needs a discussion

@@
 
+\usepackage{verbatim}
 \usepackage{amsmath}
+\usepackage{dsfont}
 \usepackage{amssymb}
@@
 \footskip 1.0cm
+
+\newcommand{\indicator}{\mathds{1}} %{{\rm I\kern-.3em E}}
 
@@
 
-An age-old aspiration within artificial intelligence research is to build a machine that automates the scientific process~\cite{paul1990autonomous,langley1987scientific}, either by designing experiments~\cite{sparkes2010towards} or
-synthesizing theories and models~\cite{schmidt2009distilling}.
-However, gaining traction on the problem of theory induction requires
-a spectrum of tractable theory synthesis problems:
-while early work attempted to recapitulate
-classic discoveries like Kepler's laws~\cite{BACON},
-these results proved brittle and difficult to generalize.
-Despite these early works,
-theory induction has never attained the same level of success as
-machine vision or natural language processing.
-How can the artificial intelligence community get theory induction off the ground?
-This is an especially difficult question,
-because the current mainstream in machine intelligence focuses on
-learning to solve a qualitatively different class of problems (e.g., prediction tasks like classification and regression),
-whereas theory induction requires
-synthesizing human-understandable causal models of real-world phenomena~\cite{pearl2009causality},
-so that human scientists can understand and learn from the AI's outputs.
+An age-old aspiration within artificial intelligence research is to
+build a machine that helps automate the scientific process by
+synthesizing theories or
+models~\cite{paul1990autonomous,langley1987scientific,schmidt2009distilling}.
+This aspiration remains largely unrealized: despite a few
+small-scale successes on toy problems~\cite{Langley1981BACON5TD},
+practicing scientists do not use machines to generate theories.
+In contrast, AI has made great strides
+on problems like machine vision and natural language processing.  How
+can the artificial intelligence community get theory induction off the
+ground?  This is an especially difficult question, because the current
+mainstream in machine intelligence focuses on qualitatively different
+classes of problems (e.g., prediction tasks like classification and
+regression), whereas theory induction requires synthesizing
+human-understandable causal models of real-world
+phenomena~\cite{pearl2009causality}, so that human scientists can
+understand and learn from the AI's outputs.
 
 We propose theory induction research start with theories of human
-language, and scope out a collection of problems and techniques for a
+language, and introduce a model of theory induction for a
 key module of natural language: \emph{morphophonology}, the
-relationship between pronunciations of words and their meaning.
+relationship between word pronunciation and meaning.
 Acquiring the morphophonology of a language involve solving a basic
-problem that confronts both linguists and children: given a collection
+problem confronting both linguists and children: given a collection
 of utterances, together with aspects of their meaning, what is the
-causal relationship between form and meaning?  Morphophonology, and
+causal relationship between form and meaning?
+\begin{comment}
+Morphophonology, and
 natural language more broadly, presents a tractable challenge to
@@
 acquire language, and field linguists can be trained to build theories of grammar.
+\end{comment}
 
-Our contribution to theory induction is
-an algorithm for synthesizing
-theories of natural language morphophonology.
-Like the linguist,
-the algorithm starts with
-a collection of utterances 
-annotated with their meanings,
-and then constructs a
-causal, interpretable model explaining how
-those meanings gave rise to the utterances.
-An important aspect of a scientific theory is
-that it makes predictions; here,
-the theory predicts pronunciations of unobserved words conditioned on their meaning.
-We evaluate our algorithm on \# data sets spanning \# languages,
-automatically finding theories that can model a wide swath of a core component of human natural language.
+Our contribution is a model for synthesizing theories of natural
+language morphophonology.  Like linguists, the model starts with a
+collection of utterances paired with meanings, and then constructs a
+causal, interpretable model explaining how those meanings gave rise to
+the utterances, roughly mirroring the process by which theorists
+construct theories starting from experimental data.  In addition to
+building theories, scientists distill those theories into higher-level
+kinds of knowledge spanning multiple theories (e.g., energy
+conservation and Lorentz invariance in physics; XXX in chemistry;
+universal grammar in linguistics).  We argue that this higher-level
+abstract knowledge is a crucial part of theory induction, imparting
+prior knowledge and constraints on what would otherwise be an
+ill-posed inductive reasoning problem.  Our model also acquires this
+higher-level knowledge by jointly inferring theories for multiple
+languages, and then shifting its inductive bias over theories to more
+closely match the attested distribution of languages.  We evaluate our
+algorithm on \# data sets spanning \# languages, automatically finding
+theories that can model a wide swath of a core component of human
+language.
 
-\begin{table}\centering
-\begin{tabular}{lcl}
+\begin{figure}
+  \includegraphics[width = \textwidth]{naturalLanguageGenerativeModel-crop.pdf}
+  \caption{Agent induces theories (teal) for a range of languages, given observed form/meaning pairs (orange).
+Grammars are expressed as programs drawn from a universal grammar, or programming language (magenta).
+}\label{generativeModel}
+  \end{figure}
+
+\section*{Discovering Theories by Synthesizing Programs}
+
+We frame our approach as Bayesian Program Learning (BPL:
+see~\cite{lake2015human}), where the model explains a set of
+utterance/meaning pairs $\left\{(u_n,m_n) \right\}_{n = 1}^N$ by
+inferring a theory $T$, which we model as a program.  Formalizing
+grammars (theories) as generative programs has a long history in
+linguistics~\cite{chomsky1968sound},
+for an intuitive reason: being procedural,
+programs can capture the causal nature of
+both grammars and theories.
+Written as a probabilistic
+inference problem, the model seeks the theory $T$ maximizing
+$P(\{u_n,m_n\}_{n = 1}^N | T)P(T|UG)$, where $UG$ is a ``universal
+grammar'' encapsulating  higher-level abstract knowledge across
+different languages.  In this BPL setting we model $UG$ as a prior
+distribution over theories (programs), and represent programs as
+Context-Sensitive Rewrites, a Turing-complete program representation,
+but restrict the rewrites in such a way as to make them equivalent to
+finite state transducers --- this approach comes from the computational
+linguistics literature~\cite{beesley2003finite}.
+
+A language's morphophonology can generate infinitely many
+utterances,
+depending on what words are in the language,
+just as theories in other sciences
+can generate an infinite set of
+possible observations --- in Newtonian mechanics,
+the theory prescribes
+how bodies will interact,
+but does not prescribe
+the number of bodies or their masses.
+Thus, for a theory to explain a set of observations,
+it must introduce additional latent (unobserved)
+variables on a dataset-by-dataset basis.
+For the theories of language considered here,
+this latent variable is the \emph{lexicon},
+which is a mapping between
+the meaning of a stem and its pronunciation.
+Taking into account the latent lexicon,
+we refine the theory-induction objective into
+finding the theory $T$ and lexicon $L$ maximizing
+$\left[\prod_{n = 1}^N \indicator[T\text{ and }L\text{ predict }u_n\text{ for }m_n] \right]P(L)P(T|UG)$.
+Fig.~\ref{generativeModel} illustrates this set up for three different languages.
+
+Although this framing captures the problem a BPL theory inductor needs to solve,
+it offers no guidance on how to solve that problem:
+the space of all programs (theories) is infinitely large and
+sharply discontinuous, lacking the local smoothness that
+enables local optimization algorithms (e.g., gradient descent; MCMC)
+to succeed.
+We adopt a strategy currently unconventional to the
+AI community: constraint--based program synthesis,
+where the optimization problem is translated
+into a combinatorial constraint satisfaction 
+problem and solved using
+a SMT solver~\cite{solar2006combinatorial}.
+To scale these solvers to
+large and complex theories 
+we wrap the solver in
+an outer loop that incrementally
+introduces new (utterance, meaning) pairs,
+incrementally modifying the theory to explain
+new data points (Supplementary materials).
+
+We apply our model to textbook morphophonology problems taken from~\cite{9780511808869}.
+Each textbook problem requires synthesizing a
+causal theory of a subset of a language.
+Fig.~\ref{languageComparison}
+compares model outputs against 
+ground-truth textbook solutions.
+These problem span a range of difficulties
+and cover a diverse set of natural language phenomena:
+systems for assigning tone (e.g., in Kerewe,
+`to count' is kubala, but `to count it' is \textipa{kuk\'ib\'ala}, where accents mark high tones),
+for ``harmonizing'' vowels (found across many languages, e.g. Kikuria
+has s\underline{ii}ka meaning `close' but s\underline{ee}kera meaning `close for';
+Latin has [ad\underline{e}ps] meaning `fat' (nominative)' but [ad\underline{i}pis] for the genitive `fat'),
+and many other linguistic phenomena like
+assimilation, epenthesis, and degemination (Fig.~\ref{exampleGrammars} and Supplementary materials).
+
+Our theory induction model covers a wider range of languages and
+phenomena than prior grammar induction algorithms from the
+computational linguistics literature: prior approaches either recover
+interpretable causal models
+(e.g.,~\cite{Albright03rulesvs,gildea1996learning,rasin2015learning})
+but do not scale to a wide range of challenging and realistic data
+sets, or abandon theory induction and instead learn opaque
+probabilistic models~\cite{cotterell-peng-eisner-2015} that may
+nonetheless predict the data well but which do not help human
+scientists generate theories.  Our improvement here stems from two
+modeling choices: First, we use a generic computational substrate ---
+context-sensitive rewrites --- giving the expressive power needed to
+explain diverse linguist phenomena.  Second, we leverage solver-based
+program synthesis techniques, borrowing decades of
+research in the programming languages community that have honed these
+tools to the point that we can scale them to
+realistic data sets using rich program representations.
+
+%2
+
+\begin{figure}
+  \begin{tabular}{lcc}    
     \toprule
-  Example data&Morphophonology\\\midrule
+    &  Tibetan Count System&Catalan Nouns\\\midrule
+    \rotatebox[origin=c]{90}{Subset of Data}&
+
 \begin{tabular}{ll}
@@
 &
+\begin{tabular}{lllll}
+  \textipa{@kelj}$\sim$\textipa{@kelj@} &&\textipa{mal}$\sim$\textipa{mal@} \\
+\textipa{siBil}$\sim$\textipa{siBil@} &&\textipa{@skerp}$\sim$\textipa{@skerp@} \\
+\textipa{Sop}$\sim$\textipa{Sop@} &&\textipa{sEk}$\sim$\textipa{sEk@} \\
+\textipa{@spEs}$\sim$\textipa{@spEs@} &&\textipa{gros}$\sim$\textipa{gros@} \\
+\textipa{baS}$\sim$\textipa{baS@} &&\textipa{koS}$\sim$\textipa{koS@} \\
+\textipa{tot}$\sim$\textipa{tot@} &&\textipa{brut}$\sim$\textipa{brut@} \\
+\textipa{pOk}$\sim$\textipa{pOk@} &&\textipa{pr@sis}$\sim$\textipa{pr@siz@} \\
+\textipa{fr@nses}$\sim$\textipa{fr@nsez@} &&\textipa{gris}$\sim$\textipa{griz@} \\
+\textipa{k@zat}$\sim$\textipa{k@zaD@} &&\textipa{bwit}$\sim$\textipa{bwiD@} \\
+\textipa{rOt\super S}$\sim$\textipa{rOZ@} &&\textipa{bot\super S}$\sim$\textipa{boZ@} \\
+\textipa{orp}$\sim$\textipa{orB@} &&\textipa{ljark}$\sim$\textipa{ljarG@} \\
+\textipa{sek}$\sim$\textipa{seG@} &&\textipa{f@Suk}$\sim$\textipa{f@SuG@} \\
+\textipa{grok}$\sim$\textipa{groG@} &&\textipa{puruk}$\sim$\textipa{puruG@} \\
+\textipa{kandit}$\sim$\textipa{kandiD@} &&\textipa{frEt}$\sim$\textipa{frED@} \\
+\textipa{s@Gu}$\sim$\textipa{s@Gur@} &&\textipa{du}$\sim$\textipa{dur@} \\
+\textipa{s@G@Do}$\sim$\textipa{s@G@Dor@} &&\textipa{kla}$\sim$\textipa{klar@} \\
+\textipa{nu}$\sim$\textipa{nu@} &&\textipa{kru}$\sim$\textipa{kru@} \\
+\textipa{flO\~nd\super Zu}$\sim$\textipa{flO\~nd\super Z@} &&\textipa{dropu}$\sim$\textipa{drop@} \\
+\textipa{@gzakt@}$\sim$\textipa{@gzakt@} &&\textipa{@lBi}$\sim$\textipa{@lBin@} \\
+\textipa{sa}$\sim$\textipa{san@} &&\textipa{pla}$\sim$\textipa{plan@} \\
+\textipa{bo}$\sim$\textipa{bon@} &&\textipa{s@rE}$\sim$\textipa{s@rEn@} \\
+\textipa{suBlim}$\sim$\textipa{suBlim@} &&\textipa{al}$\sim$\textipa{alt@} \\
+\textipa{fOr}$\sim$\textipa{fOrt@} &&\textipa{kur}$\sim$\textipa{kurt@} \\
+\textipa{sor}$\sim$\textipa{sorD@} &&\textipa{bEr}$\sim$\textipa{bErD@} \\
+\textipa{san}$\sim$\textipa{sant@} &&\textipa{k@lEn}$\sim$\textipa{k@lEnt@} \\
+\textipa{prufun}$\sim$\textipa{prufund@} &&\textipa{f@kun}$\sim$\textipa{f@kund@} \\
+\textipa{d@sen}$\sim$\textipa{d@sent@} &&\textipa{dulen}$\sim$\textipa{dulent@} \\
+\textipa{@stuDian}$\sim$\textipa{@stuDiant@} &&\textipa{blaN}$\sim$\textipa{blaNk@} & 
+  \end{tabular}
+\\\midrule 
+\rotatebox[origin=c]{90}{Theory}&
 \begin{tabular}{l}
-  \emph{Consonant cluster reduction:}\\
   \phonb{C}{$\varnothing$}{\#}{C}\\
@@
   \end{tabular}
-
+&
+\begin{tabular}{l}
+Morphology:   stem$\sim$stem$+$\textipa{@}\\
+  \phonr{[ +coronal +sonorant -lateral ]}{$\varnothing$}{\#}
+\\\phonr{[ -sonorant ]}{[ -voice ]}{\#}
+\\\phonl{C}{\textipa{k}}{\textipa{N}}
+\\\phonb{[ +voice -nasal ]}{[ +continuant ]}{[ -nasal ]}{V}
+\\\phonb{V}{$\varnothing$}{[ -sonorant ]}{V}
+\\\phonb{\textipa{t}}{$\varnothing$}{C}{\#}
+\end{tabular}
   \\
   \bottomrule  \end{tabular}
-\caption{Tibetan count system.}
-\label{Tibetan}
-\end{table}
+\caption{Example morphophonologies}
+\label{exampleGrammars}
+    \end{figure}
 
 \begin{figure}
-  \includegraphics[width = \textwidth]{naturalLanguageGenerativeModel.pdf}
-  \caption{Theory learner induces grammars (blue) for a range of languages, given observed form/meaning pairs (pink).
-Grammars are expressed as programs drawn from a universal grammar, or programming language.
-}\label{generativeModel}
+  \includegraphics[width = \textwidth]{languageComparison.png}
+  \caption{Basically a placeholder -- we still need to decide exactly what to put in here and how to measure success.}\label{languageComparison}
 \end{figure}
 
-\section*{Discovering Theories by Synthesizing Programs}
+\section*{The Role of Higher-Level Knowledge}
 
-We frame our approach as Bayesian Program Learning (BPL: see~\cite{lake2015human}),
-where the agent explains a
-set of utterance/meaning pairs $\left\{(u_n,m_n) \right\}_{n = 1}^N$
-by inferring a theory $T$, which we model as a program.
-Formalizing grammars (theories) as generative programs has a long history in linguistics~\cite{chomsky1968sound}.
-Written as a probabilistic inference problem, the agent seeks the $T$ maximizing $\left[\prod_{n = 1}^N P(u_n,m_n | T) \right] P(T|UG)$,
-where $UG$ is a ``universal grammar'' (\textbf{TIM}: what should I cite here?)
-constraining the space of allowed theories and imparting an inductive bias over the space of grammars.
-In this BPL setting we model $UG$ as a prior distribution over theories (programs).
-To model the space of programs
-use Context-Sensitive Rewrites,
-a Turing-complete program representation,
-but restrict the rewrites
-in such a way as to make them equivalent to finite state transducers --- this insight comes from the computational linguistics literature~\cite{beesley2003finite}.
-For morphophonology,
-the relationship between meaning $m_n$ and utterance $u_n$
-involves a latent \emph{lexicon},
-which is a mapping between lexemes and the pronunciation of their stem.
-For children, linguists, and our model, the lexicon
-must also be inferred (Fig.~\ref{generativeModel}).
+No theory is built from scratch: instead, researchers borrow concepts
+and constraints from other successful theories.  Linguistics has long
+acknowledged the importance of constraints and inductive biases in
+language learning, and these principles collectively go by the name
+Universal Grammar: innate for child language learners and discovered
+empirically by linguists.
+In practice,
+both children and linguists
+may have very little data to go off,
+and thus success often hinges upon
+having the right kind of high-level knowledge.
 
-Although this framing captures the problem a BPL theory inductor needs to solve,
-it offers no guidance on how to solve that problem:
-the space of all programs (theories) is infinitely large and
-sharply discontinuous, lacking the local smoothness that
-enables local optimization algorithms (e.g., gradient descent; MCMC)
-to succeed.
-We adopt a strategy currently unconventional to the
-AI community: constraint--based program synthesis,
-where the optimization problem is translated
-into a combinatorial constraint satisfaction 
-problem and solved using
-a SMT solver~\cite{solar2006combinatorial}.
+Our model represents and acquires this cross-theory knowledge
+by jointly inferring $UG$ along with the grammars for each data set.
+Assuming we have $D$ datasets (e.g., from different languages),
+notated $\left\{\left\{(u_n^d,m_n^d) \right\}_{n = 1}^{N_d} \right\}_{d = 1}^D$,
+we propose that a
+theory inductor construct $D$ theories, $\left\{T_d \right\}_{d = 1}^D$,
+along with a universal grammar $UG$, maximizing
+$$
+P(UG)\prod_{d = 1}^D P(T_d|UG)P(\left\{(u_n^d,m_n^d) \right\}_{n = 1}^{N_d}|T_d)
+$$
+where $P(UG)$ is a prior distribution over universal grammars.
+As a first approximation to this goal,
+we modeled the space of universal grammars
+using a formalism known as Fragment Grammars~\cite{tim},
+which work by saving and reusing pieces (`fragments')
+of the symbolic structure of tree-shaped representations.
+Here the trees are programs,
+so by inferring a Fragment Grammar across the theories for the different data sets,
+the model learns
+pieces of the higher-level structure
+found across languages.
+This automatically learned higher-level knowledge
+serves two functions:
+First, it is human interpretable:
+manually inspecting the contents of the fragment grammar reveals
+cross-language motifs previously discovered by linguists (e.g., word-final devoicing of obstruents,
+which occurs in XX\% of the world's languages).
+Second,
+it aids further theory induction:
+learning $UG$ allows the model to
+find better solutions to the textbook problems than if it solves
+each problem in isolation using a fixed,
+uninformative $UG$ (Fig.~\ref{languageComparison}).
 

### 2018-06-20 manuscript.tex
discussion

@@
 linguistics~\cite{chomsky1968sound},
-for an intuitive reason: being procedural,
+for two intuitive reasons: being procedural,
 programs can capture the causal nature of
-both grammars and theories.
+both grammars and theories;
+and being highly structured,
+programs can be interpreted by human scientists.
 Written as a probabilistic
-inference problem, the model seeks the theory $T$ maximizing
+inference problem, our model seeks the theory $T$ maximizing
 $P(\{u_n,m_n\}_{n = 1}^N | T)P(T|UG)$, where $UG$ is a ``universal
@@
 to succeed.
-We adopt a strategy currently unconventional to the
-AI community: constraint--based program synthesis,
+We adopt a strategy based on constraint--based program synthesis,
 where the optimization problem is translated
@@
 language learning, and these principles collectively go by the name
-Universal Grammar: innate for child language learners and discovered
-empirically by linguists.
+Universal Grammar: innate for child language learners and 
+empirically probed by linguists.
 In practice,
@@
 
+Theory induction is a grand challenge for AI,
+and our work here
+captures only small slices of
+the theory building process.
+Like our model, human theorists
+do craft models
+by examining experimental data,
+but also propose new theories by
+unifying existing theoretical frameworks,
+performing `thought experiments',
+and inventing new formalisms.
+Humans also deploy their theories more richly than our model:
+proposing new experiments to test theoretical predictions,
+engineering new tools based on the conclusions of a theory,
+and distilling higher-level knowledge that goes far beyond what our
+Fragment-Grammar approximation can represent.
+Continuing to push theory induction along these many dimensions 
+remains a prime target for future research.
+
 \bibliography{main}

### 2018-06-25 manuscript.tex
more introduction

@@
 models~\cite{paul1990autonomous,langley1987scientific,schmidt2009distilling}.
-This aspiration remains largely unrealized: despite a few
-small-scale successes on toy problems~\cite{Langley1981BACON5TD},
-practicing scientists do not use machines to generate theories.
-In contrast, AI has made great strides
-on problems like machine vision and natural language processing.  How
-can the artificial intelligence community get theory induction off the
-ground?  This is an especially difficult question, because the current
-mainstream in machine intelligence focuses on qualitatively different
-classes of problems (e.g., prediction tasks like classification and
-regression), whereas theory induction requires synthesizing
-human-understandable causal models of real-world
-phenomena~\cite{pearl2009causality}, so that human scientists can
-understand and learn from the AI's outputs.
+This aspiration remains largely unrealized: despite small-scale
+demonstrations of machine-assisted theory
+induction~\cite{Langley1981BACON5TD,schmidt2009distilling}, practicing
+scientists do not use machines to generate theories.  In contrast, AI
+has made great strides on problems like machine vision and natural
+language processing.  How can the artificial intelligence community
+get theory induction off the ground?  This is an especially difficult
+question, because the current mainstream in machine intelligence
+focuses on qualitatively different classes of problems (e.g.,
+prediction tasks like classification and regression), whereas theory
+induction requires synthesizing human-understandable causal models of
+real-world phenomena~\cite{pearl2009causality}, so that human
+scientists can understand and learn from the AI's outputs.
 
-We propose theory induction research start with theories of human
-language, and introduce a model of theory induction for a
+Theory induction is a challenge both for the natural sciences and AI,
+with connections to questions in cognitive development.
+Children, like scientists,
+build systems of laws and concepts
+to explain the world around them,
+building intuitive theories of kinship, biology, physics, number,
+grammar, and other domains, motivating the `child as scientist' metaphor~\cite{childAsScientist}.
+To bridge these different kinds of theory induction---in scientists, children, and machines---
+we propose theory induction research start with theories of human language,
+for several reasons.
+First, scientists, specifically linguists,
+%% build theories of language starting from
+%% experimental data (utterances paired with their meanings),
+%% and 
+have compiled large corpora 
+from a variety of languages,
+giving a rich and varied dataset for benchmarking theory induction algorithms.
+Second, children easily acquire language from
+modest amounts of data,
+suggesting that
+inducing theories of language is tractable,
+even from sparse data;
+and also suggesting that accounts of
+linguistic theory induction
+could shed light on language acquisition.
+Going back to at least Chomsky,
+linguists have sometimes thought of the acquisition problem as
+approximating the problem facing the linguist~\cite{chomsky1956three,chomsky1988current}.
+Third,
+theories of language
+are traditionally formalized in computational terms,
+exposing a suite of formalisms ready to be deployed by AI researchers.
+These three features of human language --- the availability of many highly-varied datasets,
+the interfaces with cognitive development,
+and the computational formalisms within linguistics ---
+conspire to single out language as an especially suitable target for
+research in automated theory induction.
+
+We have proposed theory induction research start with theories of human
+language, and here introduce a model of theory induction for a
 key module of natural language: \emph{morphophonology}, the
@@
 \end{comment}
-
 Our contribution is a model for synthesizing theories of natural

### 2018-10-17 manuscript.tex
updated data

@@
 
-\author
-{John Smith,$^{1\ast}$ Jane Doe,$^{1}$ Joe Scientist$^{2}$\\
-\\
-\normalsize{$^{1}$Department of Chemistry, University of Wherever,}\\
-\normalsize{An Unknown Address, Wherever, ST 00000, USA}\\
-\normalsize{$^{2}$Another Unknown Address, Palookaville, ST 99999, USA}\\
-\\
-\normalsize{$^\ast$To whom correspondence should be addressed; E-mail:  jsmith@wherever.edu.}
-}
+%% \author
+%% {John Smith,$^{1\ast}$ Jane Doe,$^{1}$ Joe Scientist$^{2}$\\
+%% \\
+%% \normalsize{$^{1}$Department of Chemistry, University of Wherever,}\\
+%% \normalsize{An Unknown Address, Wherever, ST 00000, USA}\\
+%% \normalsize{$^{2}$Another Unknown Address, Palookaville, ST 99999, USA}\\
+%% \\
+%% \normalsize{$^\ast$To whom correspondence should be addressed; E-mail:  jsmith@wherever.edu.}
+%% }
 
@@
   \includegraphics[width = \textwidth]{languageComparison.png}
-  \caption{Basically a placeholder -- we still need to decide exactly what to put in here and how to measure success.}\label{languageComparison}
+  \caption{Program-synthesis-based phonological rule learner applied to 37 problems from a standard phonology textbook~\cite{9780511808869}.
+\% data covered by rules: fraction of the phonology problem that is explained by the learned program (so 100\% does not mean a \emph{perfect} solution, but a \emph{consistent} solution).
+Problems marked with an asterisk are very easy problems; the other problems are more challenging.
+Not shown: last dozen problems in the textbook, which are very hard.}\label{languageComparison}
 \end{figure}

### 2018-10-17 topics.tex
renamed problem solutions
updated data

@@
 
+\usepackage{verbatim}
 \usepackage{amsmath}
@@
 \newcommand{\expect}{\mathds{E}} %{{\rm I\kern-.3em E}}
-\newcommand{\probability}{\mathds{P}} %{{\rm I\kern-.3em P}}
 
@@
 
-\section{Automatically generating grammars with Bayesian program learning}
+\section{Hypothesis generation}
+\label{althyp}
 
-\subsection{A Probabilistic Framing}
+\newcommand{\probability}[1]{P\left(#1 \right)} %{{\rm I\kern-.3em P}}
 
@@
 a collection of utterances.
-One way of
-modeling this inference problem is to (1) place a prior distribution over grammars, for example, a distribution that puts more probability mass on shorter or simpler grammars; (2) equip each grammar with a
+In reference to the previous section (Sec.~\ref{modelselection}),
+we can think of this as a kind of \emph{exploratory data analysis}.
+One way of formalizing
+ this inference problem is to (1) place a prior distribution over grammars, for example, a distribution that puts more probability mass on shorter or simpler grammars; (2) equip each grammar with a
 \emph{likelihood model} that specifies exactly how likely a grammar is to produce a given utterance;
@@
 \begin{equation}
-  \probability\left[\text{grammar}|\text{utterances} \right]\propto\underbrace{\probability\left[\text{utterances}|\text{grammar} \right]}_{\text{Likelihood}}\times\underbrace{\probability\left[\text{grammar} \right]}_{\text{Prior}}\label{posterior}
+  \probability{\text{grammar}|\text{utterances}}\propto\underbrace{\probability{\text{utterances}|\text{grammar} }}_{\text{Likelihood}}\times\underbrace{\probability{\text{grammar} }}_{\text{Prior}}\label{posterior}
 \end{equation}
@@
 prior and likelihood.
-In this work we will use this probabilistic framing
-to explore spaces of grammars
-that an AGL learner could use to explain
-the pronunciations of words.
-Here the grammars will be SPE-style rewrite rules~\cite{chomsky1968sound},
+Our goal will be both to automatically recover the single best grammar maximizing~\ref{posterior},
+as well as systematically exploring and generating the space of possible grammars (i.e. hypotheses).
+
+Here we consider an AGL learner tasked with learning a grammar generating different word forms,
+for example those in~\citep{marcus1999rule,gerken2010infants}.
+We model these grammars as context-sensitive rewrite rules~\citep{chomsky1968sound},
 commonly used in generative phonology,
 but at a high level this framing can apply to
-learning other aspects of grammar like semantics~\cite{piantadosi2011learning}, syntax~\cite{perfors2011learnability}, and morphology~\cite{tim}.
-
-\noindent\textbf{SPE-style rewrite notation:} Each SPE-style rewrite rule is a function that both inputs and outputs a sequence of phonemes.
+learning other aspects of grammar like semantics~\citep{piantadosi2011learning}, syntax~\citep{perfors2011learnability}, and morphology~\citep{tim}.
+%
+%\noindent\textbf{Context-sensitive rewrite notation:} 
+Each  rewrite rule is a function that both inputs and outputs a sequence of phonemes.
 These rewrites are written as ``\phonb{input}{output}{left}{right}'' %$\text{input}\to\text{output }/\text{ left } \_ \text{ right}$'',
 which means that ``input'' gets rewritten to ``output'' whenever ``left'' is to the left and ``right'' is to the right.
-Some special symbols used in these rules are \# to mean the start or end of the input;
-C/V to mean a consonant/vowel; $\sigma $ to mean a syllable;
-and $\varnothing$ to mean the empty string.
-For example, the rule
-\phonb{$\sigma $}{$\varnothing$}{\#}{C}
-deletes a syllable at the start of a word if it is followed by a consonant.
-Rules can refer to sets of phonemes by writing down vectors of phonological features -- for example,
-the rule \phon{[-son]}{[+voice]} rewrites segments that are -sonorant (e.g., \textipa{k,d,v,S,...}) to their voiced counterparts (e.g., \textipa{g,d,v,Z,...}).
-Rules can also bind variables to phonemes or syllables,
-which are written using subscripts;
-for example,
-\phonr{C$_i$}{$\varnothing$}{C$_i$}
+%JZ: examples and notation moves to figure caption
+%Some special symbols used in these rules are \# to mean the start or end of the input; C/V to mean a consonant/vowel; $\sigma $ to mean a syllable; and $\varnothing$ to mean the empty string.
+For example, the rule \phonb{$\sigma $}{$\varnothing$}{\#}{C} deletes a syllable at the start of a word if it is followed by a consonant.
+Rules can refer to sets of phonemes by writing down vectors of phonological features;
+%-- for example, the rule \phon{[-son]}{[+voice]} rewrites segments that are -sonorant (e.g., \textipa{k,d,v,S,...}) to their voiced counterparts (e.g., \textipa{g,d,v,Z,...}).
+rules can also bind variables to phonemes or syllables.
+%which are written using subscripts;
+%for example, \phonr{C$_i$}{$\varnothing$}{C$_i$}
 %$\text{\texttt{C}}_i\to\varnothing / \_ \text{\texttt{C}}_i$
-deletes a consonant if it is followed by the same consonant,
-and
-\phonb{$ \varnothing $}{$\sigma _i$}{#}{$\sigma _i$}
+%deletes a consonant if it is followed by the same consonant,and
+%\phonb{$ \varnothing $}{$\sigma _i$}{\#}{$\sigma _i$}
 %$\varnothing\to\sigma_i / \text{\texttt{\#}}\_\sigma_i$
-inserts a copy of a word initial syllable.
-The special subscript ``0" means zero or more occurrences of the subscripted segment,
-so for example \phonl{V}{[+hi]}{[+hi][ ]$_0$} means that vowels become +hi whenever there exists a +hi phoneme anywhere to the left.
+%inserts a copy of a word initial syllable.
+%The special subscript ``0" means zero or more occurrences of the subscripted segment,
+%so for example \phonl{V}{[+hi]}{[+hi][ ]$_0$} means that vowels become +hi whenever there exists a +hi phoneme anywhere to the left.
 
-\noindent \textbf{Prior \& Likelihood:} A simple and intuitive prior over grammars is one which penalizes longer grammars and favors parsimonious grammars, for example defining $\probability\left[\text{grammar} \right]\propto\exp\left(- \text{\# symbols in grammar} \right)$.
+%\noindent \textbf{Prior \& Likelihood:} 
+Given a collection of rewrite rules that define possible sequences of phonemes, we can define \emph{prior} probabilities over grammars, and the \emph{likelihood} of the data given a grammar. 
+A simple and intuitive prior over grammars is one which penalizes longer grammars and favors parsimonious grammars, for example defining
+$$\probability{\text{grammar} }\propto\exp\left(- \text{\# symbols in grammar} \right)$$
 An example of a likelihood model,
 and the one we use here,
-is just to count the number of extra bits or symbols needed to encode an utterance given the grammar\footnote{This is equivalent to a minimum description length (MDL) interpretation}.
+is just to count the number of extra bits or symbols needed to encode an utterance given the grammar\footnote{This is equivalent to a minimum description length (MDL) interpretation.}.
 
-Now in theory this probabilistic framing
-has now bought us a procedure for automatically discovering
+In theory, this probabilistic framing
+has now bought us a procedure for automatically generating
 grammars that do a good job explaining the data.
@@
 and so cashing out this framing requires a
-efficient procedure for searching the space of grammars and honing in on those that maximize Equation~\ref{posterior}.
+efficient procedure for searching the space of grammars and honing in on those that maximize equation~\ref{posterior}.
 A subfield of computer science, called \textbf{program synthesis},
@@
 like those that arise in grammar learning.
-We use a program synthesis tool called Sketch~\cite{solar2008program}
-to search for grammars maximizing~\ref{posterior},
+We use a program synthesis tool called Sketch~\citep{solar2008program}
+to search for grammars maximizing equation~\ref{posterior},
 which was previously applied to language learning problems in~\cite{ellis2015unsupervised}.
@@
 This cashing out of the framing makes
-our approach an instance of  \textbf{Bayesian Program Learning (BPL)}; see~\cite{lake2015human}.
+our approach an instance of  \textbf{Bayesian Program Learning}; see~\cite{lake2015human}.
 
-\subsection{Modeling solutions to AGL problems}
+%\subsection{Modeling solutions to AGL problems}
 
+\begin{comment}
 The fundamental hypothesis underlying AGL research is that
@@
 instead studying AGL in controlled experiments.
-Our demonstration here is that
-a BPL model can explain how a learner can quickly (e.g., from very little data) infer a grammar represented as SPE--style rules.
-We collected grammars from AGL datasets~\cite{gerken2010infants,marcus1999rule,frank2011three}
-and presented our BPL learner with data drawn from these grammars (Table~\ref{artificialGrammarTable}),
-finding that our model automatically discovers the grammars that
+Our contribution here is the demonstration that
+a single Bayesian program learning model can explain how a learner can quickly infer a grammar represented as context--sensitive rewrites.
+We collected grammars from AGL datasets~\citep{gerken2010infants,marcus1999rule,frank2011three}
+and presented our model with data drawn from these grammars (Table~\ref{artificialGrammarTable}),
+finding that the learner automatically discovers the grammars that
  underly these data sets.
-\begin{table}[h!]
+\end{comment}
+%
+\begin{table}[htbp]
 \centering\begin{tabular}{lcll}
@@
   Grammar&\begin{tabular}{c}
-    Example input \\to learner
+    Example input \\to model
     \end{tabular}&Inferred grammar&Natural language analogues\\\midrule
@@
     ABA \\
-  Marcus et al. 1999.     
+%  Marcus et al. 1999.     
     \end{tabular}&
@@
       ABB\\
-  Marcus et al. 1999. 
+%  Marcus et al. 1999. 
     \end{tabular}&
@@
   \begin{tabular}{l}
-    AAx \\Gerken 2006. %~\cite{gerken2010infants}
+    AAx \\%Gerken 2006. %~\cite{gerken2010infants}
     \end{tabular}&
@@
   \begin{tabular}{l}
-    AxA \\Gerken 2006. %~\cite{gerken2010infants}
+    AxA \\%Gerken 2006. %~\cite{gerken2010infants}
     \end{tabular}&
@@
 \\  \bottomrule  \end{tabular}
-\caption{Using BPL to model artificial grammar learning. Learner given 5 examples.}\label{artificialGrammarTable}
+\caption{Using Bayesian program learning to model artificial grammar learning. Each model given 5 examples. Grammars ABA and ABB are from Marcus et al. 1999; grammars AAx and AXA from Gerken, 2006.
+Some special symbols used in the grammar rules are \# to mean the start or end of the input; C/V to mean a consonant/vowel; $\sigma $ to mean a syllable; and $\varnothing$ to mean the empty string.
+For example, the rule \phonb{$\sigma $}{$\varnothing$}{\#}{C} deletes a syllable at the start of a word if it is followed by a consonant.
+The rule \phon{[-son]}{[+voice]} rewrites segments that are -sonorant (e.g., \textipa{k,d,v,S,...}) to their voiced counterparts (e.g., \textipa{g,d,v,Z,...}).
+The rule \phonr{C$_i$}{$\varnothing$}{C$_i$}
+%$\text{\texttt{C}}_i\to\varnothing / \_ \text{\texttt{C}}_i$
+deletes a consonant if it is followed by the same consonant,
+and
+\phonb{$ \varnothing $}{$\sigma _i$}{\#}{$\sigma _i$}
+%$\varnothing\to\sigma_i / \text{\texttt{\#}}\_\sigma_i$
+inserts a copy of a word initial syllable.
+The special subscript ``0" means zero or more occurrences of the subscripted segment,
+so for example \phonl{V}{[+hi]}{[+hi][ ]$_0$} means that vowels become +hi whenever there exists a +hi phoneme anywhere to the left.
+}\label{artificialGrammarTable}
 \end{table}
 
-\subsection{Automatically exploring the space of alternative hypotheses}
+%\subsection{Automatically exploring the space of alternative hypotheses}
 
-Just like in first language acquisition,
-these AGL learning setups introduce a trade-off between
-grammars that are likely under the prior (and are therefore small; ``parsimony'')  and grammars that
-assign high likelihood to the data (and therefore closely ``fit the data'', having high likelihood).
-For example, a learner could infer a grammar that just memorized to the data (perfect fit but poor parsimony)
+The fundamental hypothesis underlying AGL research is that
+artificial grammar learning must engage some cognitive resource shared with first language acquisition. To gain insights into first language acquisition, we 
+instead study AGL in controlled experiments.
+In particular, we are interested in how the human mind deals with the trade-off between
+grammars that are \emph{a priori} probable (the prior prefers small grammars; ``parsimony''), and grammars that
+assign high likelihood to the data (and therefore closely ``fit the data'').
+For example, a learner could infer a grammar that just memorized the data (perfect fit but poor parsimony)
 or it could infer a grammar that can generate every
@@
 
-To analyze this trade-off, we calculated its shape in the form of a
-\textbf{pareto front}~\cite{mattson2005pareto}.
-In our setting, a pareto front is the set of all grammars that
+To analyze this trade-off, we present the model with artificial data drawn from grammars used in a series of AGL experiments~\citep{gerken2010infants,Marcus1999,frank2011three}.
+The grammars used (and the notation we use for them) are described in Table~\ref{artificialGrammarTable}.
+
+Rather than producing a single grammar -- a single hypothesis on the grammar a child might acquire in such an experiment -- we use the model to search a massive space of possible grammars and compute all those grammars that are optimal solutions to the trade-off given some weighting of parsimony and fit to data.
+That is, we search for the set of grammars that
 are not worse than another grammar along
-the two competing axes of parsimony and fit to data.
+the two competing axes; this set is called the \textbf{pareto front}~\citep{mattson2005pareto}.
 Intuitively, grammars on the pareto front
-are ones which an ideal Bayesian or MDL learner prefers,
+are ones which an ideal Bayesian learner prefers,
 \emph{independent} of how the learner decides to
 relatively weight the prior and likelihood.
+
 Figure~\ref{quadrupleFrontier}
@@
 as measured by the prior (parsimony) and the likelihood (fit to the data).
-We believe that the Pareto front is a useful way of
-viewing the space of possible grammars,
-and understanding the different trade-offs
-between the grammars.
+%We believe that the Pareto front is a useful way of viewing the space of possible grammars, and understanding the different trade-offs between the grammars.
 
@@
 \begin{itemize}
-\item \textbf{A grammar that does nothing:} In the lower right-hand corner of~\ref{quadrupleFrontier}
+\item \textbf{A grammar that generates every possible word:} In the lower right-hand corner of Figure~\ref{quadrupleFrontier}
   is the grammar labeled ``surface=underlying'',
@@
   giving a fit to the data of -6.
-\item \textbf{A grammar that duplicates syllables:} Highlighted in dark red in~\ref{quadrupleFrontier}
+\item \textbf{A grammar that duplicates syllables:} Highlighted in dark red in Figure~\ref{quadrupleFrontier}
   is a grammar with the single rule
@@
   %$\varnothing\to\sigma_i/\_\sigma_i$\texttt{\#},
-  which inserts a copy of the last syllable syllable.
+  which inserts a copy of the last syllable.
   This rule generates the word [\textipa{vEfefe}] by starting with the stem (i.e., underlying form) /\textipa{vEfe}/ (which has 4 symbols, fit to data of -4) and then applying the rule in the grammar,
@@
 \item \textbf{A grammar that duplicates syllables and appends
-  morphemes:} Highlighted in pink in~\ref{quadrupleFrontier} are
+  morphemes:} Highlighted in pink in Figure~\ref{quadrupleFrontier} are
   grammars that duplicate a syllable but also append or prepend extra
@@
 
-\begin{figure}[H]
+\begin{figure}[htbp]
 
@@
 
-  \caption{Pareto fronts for ABB  (Marcus~1999; top two) \& AAX  (Gerken~2006; bottom two) learning problems fo either one example word (left column) or three example words  (right column).
+  \caption{Pareto fronts for ABB  (Marcus~1999; top two) \& AAX  (Gerken~2006; bottom two) learning problems for either one example word (left column) or three example words  (right column).
 Rightward on x-axis corresponds to more parsimonious grammars and upward on y-axis corresponds to grammars that best fit the data, so the best grammars live in the upper right corners of the graphs.
@@
 As the number of examples increases, the Pareto fronts develop a sharp kink around the ground truth grammar, which indicates a stronger preference for the correct grammar.
-With one example the kinks can still exists but are less pronounced.}\label{quadrupleFrontier}
+With one example the kinks can still exist but are less pronounced.}\label{quadrupleFrontier}
 \end{figure}
 
-\subsection{Beyond artificial grammars: Applying BPL to natural language}
+%\margincomment{Tim: Is this an okay explanation of why we might want to explore the set of different possible hypotheses as opposed to just taking one of them?}
+
+Grammar induction is an under constrained problem, and in general there are infinitely many grammars that could explain a collection of utterances.
+Despite this ambiguity, both children and linguists can make plausible inferences about which grammars best explain a data set.
+The computational tools shown here offer a generic way of generating and visualizing the range of alternative hypotheses, and explaining why a child, linguist, or computer program might prefer one over the other.
+
+\begin{comment}
+
+\subsection{Beyond artificial grammars: Applying Bayesian program learning to natural language}
 
@@
 
-\begin{figure}[H]\centering
+\begin{figure}\centering
   \begin{minipage}{7cm}\centering
@@
 10 & \textipa{bd\super Zu}
-\bottomrule
+\\\bottomrule
 \end{tabular}    
@@
   \end{figure}
+\end{comment}
 

### 2018-10-24 manuscript.tex
language comparison figure

@@
 \begin{figure}
-  \includegraphics[width = \textwidth]{languageComparison.png}
-  \caption{Program-synthesis-based phonological rule learner applied to 37 problems from a standard phonology textbook~\cite{9780511808869}.
+\centering\vspace{-3cm}  \includegraphics[width = 0.7\textwidth]{figures/languageComparison.eps}
+  \caption{Program-synthesis-based phonological rule learner applied to 38 problems from a standard phonology textbook~\cite{9780511808869}.
 \% data covered by rules: fraction of the phonology problem that is explained by the learned program (so 100\% does not mean a \emph{perfect} solution, but a \emph{consistent} solution).
 Problems marked with an asterisk are very easy problems; the other problems are more challenging.
-Not shown: last dozen problems in the textbook, which are very hard.}\label{languageComparison}
+Not shown: last eleven problems in the textbook, which are very hard: those results are pending.}\label{languageComparison}
 \end{figure}
